---
title: "Practice Session 1 Answers"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---

# Part 1:  Quarto
  
  
  Quarto documents allow one to create reproducible analyses that include both
written explanations as well as R code. We will use Quarto documents for all the
class code and for the homework. For more information on Quarto see <https://quarto.org/>
  
  
### 1.1 Rendering Quarto documents
  
  
  We can render a Quarto document to a pdf document by pressing the **Render button** 
which is located at the top of the Quarto file in RStudio (the button has an arrow symbol). This will generate a pdf that includes both content as well as the output of any embedded R code chunks within the document.


**Note**: When working on the homework, please be sure to render the Quarto 
document often so that you can tell right away when you make a mistake.
Otherwise, it will be very difficulty to debug your code at the end when you try
to submit the homework.

$\\$

### 1.2 LaTeX symbols

We can include LaTeX symbols in our Quarto documents. For example, we can 
write the Greek letter alpha using the code $\alpha$. 

We can also write subscripts using underscores. For example, $x_{original}$.


Sometimes we will want to put marks above characters such as a hat above a letter.
For example, we can write $\hat{y}$ to get a hat above the letter y. We can also
write $\bar{y}$ to get a bar above it.


**Exercises**: Try writing the following:
  
* x-bar subscript 20
* Pi subscript 1
* $\beta$-hat subscript 0


**Answers:**
  
  * $\bar{x}_{20}$
  * $\pi_1$
  * $\hat{\beta}_0$
  
$\\$

# Part 2: Introduction to R
  
  
  Let's now discuss some of the basics of R!


### 2.1 Running R code

We can run R code on the console or inside of R chunks. If we press the green
"play" button at the top of the R chunk, this is the same as running the code
in the console. 


```{r simple_math}
1000 + 9999 # Addition
10 * 4 # Multiplying 2 numbers
7 * 7 * 20 # Multiplying 3 numbers
3^4 # Exponents (i.e., 3 to the fourth power)
132 / 11 # Division
```


**Exercises**: Compute the area of the following shapes:

* Square with side length 4
* Circle with radius 3 (you can use 3.14 for $\pi$)
* Triangle with base 5 and height 6


**Answers:**

* `4 * 4 (16)`
* `3.14 * 3^2 (28.26)`
* `0.5 * 5 * 6 (15)`

$\\$

### 2.2 Assigning values to objects


We can assign values to objects using the assignment operator `<-`. 

Note, there cannot be any spaces in between the `<` and `-` symbols.

Any object that is placed on a line by itself will print the value stored in
that object. This is how you will "show your work" on the homework; i.e., show
that you have the correct answer.


```{r assignment_example}

a <- 45
b <- 55

z <- a + b

z

fruit <- "apple"
vegetable <- "carrot"

```

**Exercises**: Assign the areas that you computed above to corresponding variables.
You can name the variables based on the shape (e.g., square <- $\ldots$).

Area of : 

1. Square  : $A = s^2$

2. Circle  : $A = \pi r^2$

3. Triangle: $A = \frac{1}{2} b h$


**Answers:**

* square = 4 * 4
* circle = 3.14 * 3^2
* triangle = 0.5 * 5 * 6


$\\$


### 2.3 Functions

Functions are used to perform specific tasks. Usually they take input values
and return an output value.

Function use syntax that ends with parenthesis: functionName(x)

Any ideas what the functions below do?

1. round(49.209, 2)	

2. toupper("panama!")


To get information about a function, use the ? before the function name. 

For example: ?round

**Exercises**: Find the help page for the `toupper` function.

**Answers:**

* ?toupper


$\\$

### 2.4 Vectors

Vectors are used to store multiple values in a single object. We can create
vectors using the c() function.

We can access values in a vector using square brackets [].


```{r vectors_example}

v <- c(1, 3, 5, 7, 9) # defining a vector called v

v

states <- c("Alabama", "Alaska", "Arizona", "Arkansas")

states[4] # indexing or selecting the fourth element in the vector

w = 1:10 # Sequence of numbers from 1 through 10

w
```


**Exercises**: Answer the following using the vector `v` that we defined above.

* Find the sum of `v` above using the `sum` function. Save the result to a vector called `v_sum`
* Take the square root of `v` using the `sqrt` function. Save the result to a vector called `v_sqrt`
* Subtract `v_sqrt` from `v_sum`, and round it to 2 decimal places


**Answers:**

* v_sum = sum(v) 
* v_sqrt = sqrt(v)
* round(v_sum - v_sqrt, 2)

$\\$

# Part 2: Quick review of categorical data analysis in R

Let's quickly review how to analyze categorical data in R using a fictional
dataset of voters' opinions of president Trump. 



### 2.1 Getting a simulated sample of voters


We can use the SDS1000 function `get_approval_sample()` to get a vector of random
sprinkle colors. 

```{r voter_sample}

# Load the SDS1000 package
library(SDS1000)

# Setting the seed so that we all get the same "random sample"
set.seed(1000)  



# Get a random sample of 1000 fictional voter's opinion of president Trump
approval_sample <- get_approval_sample(1000)


# Show the first 10 voters in the sample
approval_sample [1:10]

```

$\\$
  
### 2.2 Get the proportion:

#### Example: Get the proportion of voters that appropove of trump
  
We can use the `table()` and `prop.table()` functions to create a frequency and
relative frequency table of approval ratings.

We can also use the SDS1000 `get_proprortion()` function to get the proportion
of voters that approve of president Trump.


```{r voter_frequency_table}

# Create a table of the approval ratings
approval_table <- table(approval_sample)
approval_table

# Create a relative frequency table of the approval ratings
approval_proportions <- prop.table(approval_table)
approval_proportions

# Get the proportion that approves
get_proportion(approval_sample, "approve")

```

$\\$
  
### 2.3 Bar plots and pie charts
  
  We can use the `barplot()` and `pie()` functions to visualize the data.

Note: Both functions take a frequency table as input, and **not** the original
vector of data. 

```{r voter_barplot_and_pie_chart}


# Create a bar plot of the sprinkle colors
barplot(approval_table, ylab = "Count")


# Create a pie chart of the sprinkle colors
pie(approval_table)

```


Can you figure out how to change the color of the segments in the pie chart?
  
$\\$

# Part 3:  Analyzing Quantitative Data 

### Mean, Median, and histogram: 
  
  Generate histograms for each of the following data sets. Use the `$` command to access the individual data sets. 
  
  1. For each histogram, add the **mean**  and the **median** to the plot using `abline()`. 
  
  2. Describe the shape of the histogram, the position of the mean, and the position of median.
  
  3. Do you see any potential outliers? 
  
```{r}
set.seed(999)

mydata <- data.frame(
  dat1 = -rchisq(1000, df = 1),
  dat2 = rchisq(1000, df = 1),
  dat3 = runif(1000),
  dat4 = rnorm(1000),
  dat5 = sample(c(rnorm(1000, mean = 2), rnorm(1000, mean = 10)), size = 1000)
)


```

**Answers**: 

```{r}

mean1 <- mean (mydata$ dat1 )
mean1

median1 <-  median (mydata$ dat1 )
median1

hist1<- hist( mydata$ dat1,
  
     col = "pink1",                           # bins color   
     main = "Histogram for mydata$dat1",      # title
     xlab = "mydata$dat values",              # x-axis label
     ylab = "Frequencies of mydata$dat",      # y-axis label
     border = "darkgreen",                    # bins border's color 
     lwd = 2,                                 # border thickness
     cex.main = 1.5,                          # title size
     cex.lab = 1,                             # axis labels size
     cex.axis = 1.2,                          # tick labels size
     font.main = 2,                           # bold title
     font.lab = 3)                            # italic axis labels
     

abline(v= mean1, col="blue" )
abline(v= median1, col="red2" )


```


