---
title: "Homework 2"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---




$\\$




In the following homework, you will practice analyzing quantitative data using
histograms, box plots, and z-scores. Please render this Quarto document and
submit the compiled pdf document with your answers to Gradescope
*by 11pm on Sunday February 1st*. Please also make sure to mark the correct
pages to each question's answers on Gradescope to allow the TAs to grade each
problem.


Some useful functions for completing this worksheet are: `dim()`, `length()`, 
`hist()`, `mean()`, `median()`, `max()`, `sd()`, `sum()`,  `fivenum()`, and 
`boxplot()`. Also see the "R function" page on Canvas for a list of functions we
have used in class. You might also find the following symbols useful:  $\mu$, 
$\bar{x}$, $\pi$,  $\hat{p}$. 


For all questions that involve R, please use the R chunks (gray sections) to
create relevant plots and calculations, and then answer the questions in the
"answer" section below each chunk. Be sure to "show your work" by printing out
information that answers the question that have been asked (although please do
not print out pages of irrelevant data). Also, make sure to add appropriate
labels to all plots.


If you need help with any of the homework assignments, please attend the TA
office hours which are listed on Canvas and/or ask questions on Ed Discussions
forum. Also, if you have completed the homework, please help others out by
answering questions on Ed Discussions. Finally, reviewing the class slides and
videos could be helpful if you run into difficulties.





$\\$





# Part 1:  Lock 5 questions



Please answer the following questions that are based on questions from the Lock5
textbook. Some symbols that might be useful for answering these questions
include $\hat{p}$, $\bar{x}$, $\pi$, and $\mu$. You can also write subscripts in
LaTeX by using an underscore, e.g., $\beta_{treatment}$.



$\\$







**Exercise 1.1 (6 points)**:  A set of lucky numbers are: 41, 53, 38, 32, 115,
47, and 50. For these lucky numbers, find:


a. the mean $\bar{x}$

b. the median m

c. indicate whether there appears to be any outliers, and if so, what are they?


For **this question** you **do not** need to "show your work" by including R code, 
although you are welcome to use R to get the answers. (In general, there will
be an R chunk given in the homework Quarto document for all problems where we 
expect you to show how you used R to solve a problem). 



**Answers:**  


a. 



b. 



c. 






$\\$




**Exercise 1.2 (6 points)**: Using only the whole numbers 1 through 9 as
possible data values, create a data set with n = 6, and $\bar{x}$ = 5 with:

a. The standard deviation as small as possible.

b. The standard deviation as large as possible.


Note: There can be repeated numbers in your data set (e.g., it's ok to use the
value 3 twice, etc.).


**Answers:**  

a. 

b. 




$\\$




**Exercise 1.3 (9 points)**: The number of days it took 8 different rowers to
row solo across the Atlantic Ocean is: 40, 87, 78, 106, 67, 70, 153, and 81. Use
the R chunk below to calculate the z-score for the shortest and longest row
times and interpret what these values tell us in terms of the mean and standard
deviation of the sample.

**Hint**: it will be useful to create objects that hold intermediate values before
calculating your final answers. For example, start by creating a vector called
`row_times` that has the times for the different rowers. Then create objects
called `mean_time` and `sd_time` which have the mean and standard deviations of
rowers' times. Finally, calculate the z-scores for the shortest and longest row
times.



```{r row_times}










```



**Answer**: 








$\\$




# Part 2: Analyzing quantitative data



Let's now practice analyzing a single quantitative variable by examining data on
the heights of Major League Baseball (MLB) players.

We will look at all MLB players who died after 1900. The data goes up until the 
2024 baseball season. 




$\\$





**Exercise 2.1: (4 points)** The code below loads information about every MLB 
player who has played through the 2023 season in a data frame called `People`. 
Please examine the data frame and answer the questions: 

a. What does a case correspond to in this data frame?

b. How many cases are there in the data frame? 


```{r q2_1}

# load a data frame called 'People' with baseball player information 
library(Lahman)

# A data frame with information about baseball players
People <- subset(People, deathYear > 1900 | is.na(deathYear))





```


**Answers:** 

a. 

b. 





$\\$









**Exercise 2.2 (5 points):** The code below extracts a vector of heights of all 
the baseball players in inches. Please create a histogram of their heights, and be 
sure to use an appropriate number of bins and to put appropriate labels on your
axes. Describe the shape of the histogram. Are there any noticeably small or
large outliers?


```{r q2_2}


# extract a vector with baseball player heights
player_heights <- People$height








```


**Answers:** 








$\\$








**Exercise 2.3 (8 points):** Now create a box plot of the heights, and report 
the 5 number summary. Do you notice any extreme outliers? Describe what you
should do when you see an extreme outlier and then do it! Hint: Google could be
helpful here.


```{r q2_3}








```



**Answers**: 












$\\$








**Exercise 2.4 (6 points):** Next, calculate the mean and standard deviation of 
the heights of baseball players. Note, you will need to use the `na.rm = TRUE`
argument as well since there is missing data which we will ignore here.  

If the heights are normally distributed, what range of heights would you expect
for the middle 95% of the data to lie? Use the mean, and standard deviation to 
calculate this range (i.e., calculate the end  points of the interval that you 
would expect the middle 95% of the data to lie). Print out the values in the R 
chunk below and report the range of values in the answer section.


```{r q2_4}







```


**Answers**: 






$\\$






**Exercise 2.5 (5 points):**: Now calculate the actual actual 2.5th and
97.5th percentile values for the heights (also known as end points) where the
middle 95% of the data actually lies. Are the endpoints you calculated using the 
actual data close to those calculated under the assumption that the heights are 
normally distributed (from the previous exercise)?


```{r q2_5}




```


**Answers**: 







$\\$



**Exercise 2.6 (7 points):** Finally, calculate the z-score for the minimum value 
in the data set (again use the `na.rm = TRUE` argument when calculating the
mean, the standard deviation, and the minimum height in the data). Report what
this value is, and what this value means.


```{r q2_6}










```


**Answers**: 







$\\$






# Part 3: Calculating statistics "by hand"

In the following exercises you will calculate a few statistics "by hand" (i.e.,
without using the built in R functions).  The purpose of this exercise is to gain 
a better understanding of the mathematical notation used in class (and to become
thankful that we have computers!). As you answer these questions, we recommend you
look at the rendered pdf to see the formulas more clearly.


The data we will look at is the number from [Nathan’s Famous Hot Dog Eating
Contest](https://en.wikipedia.org/wiki/Nathan%27s_Hot_Dog_Eating_Contest) where
contestants  need to eat as many hot dogs as possible in 10 minutes. The data we
will look at is the number of hot dogs eaten by the winner of the competition in
the years 2004 to 2011, which is shown in the table below.

| Year | Hot Dogs |
|------|----------|
| 2011 |    62    |
| 2010 |    54    |
| 2009 |    68    |
| 2008 |    59    |
| 2007 |    66    |
| 2006 |    54    |
| 2005 |    49    |
| 2004 |    54    |


Note: Doctor's recommend you limit the number of hot dogs you eat in a day as
discussed [here](https://www.tiktok.com/@thewhitestkidsuknow/video/7542005809690070285).



$\\$




**Exercise 3.1 (12 points):** Calculating the standard deviation

In this first exercise you will calculate the standard deviation "by hand" by 
filling in the marked sections below. As you will recall from class, the formula
for calculating the standard deviation is:

$$s = \sqrt{\frac{1}{n-1} \sum_{i=1}^n ( x_i - \bar{x})^2}$$


Please use this formula, and the following steps, to calculate the standard
deviation for the number of hot dogs eaten by the winners. Show your work by
filling in the intermediate values in the all places where there are XXX symbols
(i.e., replace all XXX with values you calculate). All the values you fill in
should have an precision of two decimal places. You are also welcome to use R to
help with the calculations, but you must fill in the values showing all
intermediate steps everywhere there is a XXX.


## Steps for calculating the standard deviation

1. Calculate the mean number of hot dogs eaten and write your answer here: $\bar{x} = XXX$

2. Fill in column `b` in the table below showing the deviations of each data point 
from the mean, i.e., $x_i - \bar{x}$.

3. Fill in column `c` in the table below showing the squared deviations; i.e., $(x_i - \bar{x})^2$


| Num hot dogs | b. Deviations: $x_i - \bar{x}$ | c. Deviations squared: $(x_i - \bar{x})^2$ |
|--------------|---------------------------------|-----------------------------------------|
| 62           |             XXX                 |                  XXX                    |
| 54           |             XXX                 |                  XXX                    |
| 68           |             XXX                 |                  XXX                    |
| 59           |             XXX                 |                  XXX                    |
| 66           |             XXX                 |                  XXX                    |
| 54           |             XXX                 |                  XXX                    |
| 49           |             XXX                 |                  XXX                    |
| 54           |             XXX                 |                  XXX                    |



4. Write the sum of squared deviations value here:  $\sum_{i=1}^n (x_i - \bar{x})^2 = XXX$

5. Write the value of the squared deviations divided by $n-1$ here: $\frac{\sum_{i=1}^n (x_i - \bar{x})^2}{n-1} = XXX$

6. Write the value for square root of the sum of squared deviations here: $s = XXX$





$\\$



**Exercise 3.2 (6 points):** Calculating the five-number summary 

Now please calculate the five-number summary "by hand". To do this, sort the
data from the smallest value to the largest value, and from **looking at this sorted data**, 
use the definitions for the five-number summary discussed in class, to write down 
the five-number summary in the answer section below. 

For this question, you can use R to check your work, but you should write down
the five-number summary without using any R functions, since the purpose of this 
exercise is to make sure you understand how five-number summary is calculated.

Note: As we discussed in class, there are different ways for calculating
quartiles. For this exercise, you should calculate the quartiles based on values
that are **equal to or below** a given value in your data; i.e., each
value in your five number summary (apart from  the median) should be equal to
one of the values in your original data sample.


$\\$



**Answer** (Please write down your five-number summary below)






$\\$




## Part 4: Scatter plots and correlation 




$\\$





**Exercise 4.1 (4 points)**:  Let's explore two quantitative variables related 
to cars. Describe whether you expect a positive or negative association between
the *distance* a car was driven since the last fill-up of the gas tank, and the
*amount of gas* left in the tank. In 1-3 sentences, please explain the rationale
behind your answer.




**Answer**:  






$\\$





  
**Exercise 4.2 (12 points)**:  Suppose we record the husband's age and the 
wife's age for 100 randomly selected heterosexual married couples. Please answer 
the following questions:

a. What would it mean about ages of couples if these two variables had a
negative relationship?

b. What would it mean about ages of couples if these two variables had a
positive relationship?

c. Which do you think is more likely, a negative or a positive relationship? 

d. Do you expect a strong or weak relationship in the data? Why? 

e. Would a strong correlation imply there is an association between husband age
and wife age?

f. What symbol (of the symbols discussed in class) should we use to denote the
correlation discussed in this problem?



**Answers**:


a. 



b. 



c. 



d. 



e. 



f. 






$\\$







**Exercise 4.3 (6 points)**: The table below shows the relationship between two 
quantitative variables X and Y (you might want to render the document to see the
table better). Please use R to create a scatter plot and calculate the
correlation between these variables.  Hint: the function to create vectors,
`c()`, might be useful here.  *Challenge (0 points):* try to calculate the
correlation without using the `cor()` function.



   |    |    |    |    |    |    |    |    |    
---|----|----|----|----|----|----|----|----|     
X  | 15 | 20 | 25 | 30 | 35 | 40 | 45 | 50 |
Y  |532 |466 |478 |320 |303 |349 |275 |221 |





```{r q1_3}











```







$\\$








**Exercise 4.4 (6 points)**:  When honeybee scouts find a food source or a nice 
site for a new home, they communicate the location to the rest of the swarm by
doing a ["waggle dance"](https://www.youtube.com/watch?v=-7ijI-g4jHg). They
point in the direction of the site and dance longer for sites farther away. The
rest of the bees use the duration of the dance to predict distance to the site.
The code below loads a data frame called `HoneybeeWaggle` that contains the
distance (in meters), and the duration of the dance (in seconds) for seven
honeybee scouts. Using this data:

a. Create a scatter plot of the data, and be sure to label the plot
appropriately. Also describe whether there appears to be a linear trend in the
data. If so, is it positive or negative?

b. Use the `cor()` function to find the correlation between the two variables. 


Note: if you learn more about correlations through (human) dance please check out
[this video](https://www.youtube.com/watch?v=VFjaBh12C6s)



```{r q1_4}


# download the data and load it into R
library(Lock5Data)
data(HoneybeeWaggle)


# Plot the data here







# Calculate the correlation coefficient here






```


**Answers:**

a. 

b. 







$\\$






## Reflection (3 points)


Please reflect on how the homework went by going to Canvas, going to the Quizzes
link, and clicking on Reflection on homework 2.



