---
title: "Homework 10"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---




$\\$





Welcome to homework 10, the last homework problem set of the semester! The
purpose of this homework is to practice running hypothesis tests for comparing
multiple proportions (chi-squared tests) and for comparing multiple means
(analysis of variance). Please submit a compiled pdf with your answers to the
exercises to Gradescope *by 11pm on Monday December 1st*.

Functions you might find useful are `dchisq()`, `pchisq()`, `barplot()`,
`sum()`, `df()`, and `pf()`, along with the list of functions found on Canvas.
For full credit on problems involving R, be sure to label all your figures and
to "show your work" by making sure all values that are answers to questions are
printed/visible in your Quarto pdf.

As always, if you need help with any of the homework assignments, please attend
the TA office hours which are listed on Canvas and/or ask questions on Ed
Discussions forum. Also, if you have completed the homework, please help others
out by answering questions on Ed Discussions. Finally, attending the practice
sessions and reviewing the class slides will be helpful to solidfy your
understanding.



<!-- This R chunk sets some parameters that will be used in the rest of the document. Please ignore it and do not modify it. -->
```{r message=FALSE, warning=FALSE, tidy=TRUE, echo=FALSE}
    
  library(SDS1000)
  library(knitr)
  options(scipen=999)
  set.seed(1000)
  
```







$\\$






## Part 1: Testing goodness-of-fit for a single categorical variable



In the first set of exercises you will run a hypothesis test to compare counts
in multiple categories. I encourage you to review the examples in the Lock5
textbook for additional practice. For all your answers be sure to "show your
work" by printing the output of your calculations in the R chunks.





$\\$





Rock-paper-scissors is a hand game played between two people, where people make
hand gestures in the shape of either a rock, a piece of paper, or a pair of
scissors. In the problem below we will analyze the first move people typically
play in games of rock-paper-scissors. If you are not familiar with the game,
please check out either [this youtube video](https://www.youtube.com/watch?v=cDwK-29meMI) 
or the [wikipedia article](https://en.wikipedia.org/wiki/Rock_paper_scissors) 
that explains the rules (and if you're still unsure, I am happy to discuss the 
rules during office hours).

The table below shows the choices made by 119 players on the first turn of a
rock-paper-scissors game. A player gains an advantage in playing this game if
there is evidence that the choices made on the first turn are not equally
distributed among the three options. In the following set of exercises we will
run a $\chi^2$  goodness-of-fit test to see if there is evidence that any of the
proportions are different from 1/3.



| Option Selected | Observed Counts (O) |
------------------|---------------------|
|  Rock           |         66          |
|  Paper          |         39          |
|  Scissor        |         14          |
|  Total          |        119          |




**Exercise 1.1 (5 points)**: As described above, in this set of exercises we will 
run a $\chi^2$ goodness-of-fit hypothesis test to see if there is evidence that
any of the proportions for the first move in rock-paper-scissors games differ
from 1/3. Let's start this hypothesis test by stating the null and alternative
hypotheses using the appropriate symbols. 



**Answers**








$\\$   
 




**Exercise 1.2 (points 8)**: Now let's complete step 2 of hypothesis testing by 
calculating the $\chi^2$ statistic. To start doing this, first fill in the table
below showing the counts you would expect to get for Rock, Paper and Scissors if
the null hypothesis was true.

After you have filled in this table, then use the code section below to calculate 
the $\chi^2$ statistic based on the observed count table from exercise 1.1 and
expected count table you filled in below. Recall that the $\chi^2$ statistic is
defined as: $\chi^2 = \sum_i^k \frac{(O_i - E_i)^2}{E_i}$, where $O_i$ is the
$i^{th}$ observed count (from the table above exercise 1.1) and $E_i$ is the
$i^{th}$ expected count (from the table you filled out below). Be sure to print
out the value you get for the $\chi^2$ in the code section below as well.



**Answers**


| Option Selected | Expected Counts (E) |
------------------|---------------------|
|  Rock           |       fill in       |
|  Paper          |       fill in       |
|  Scissors       |       fill in       |
|  Total          |         119         |




```{r q1_2, message=FALSE, warning=FALSE, tidy=TRUE}






```




$\\$




  

**Exercise 1.3 (7 points)**:  Next do step 3 of hypothesis testing by plotting 
the null distribution. Start by writing down the appropriate degrees of freedom
for the null distribution in this problem in the answer section. Then use the
`dchisq(x_vals, df)` function to plot the null chi-squared distribution for the
null distribution. In particular, use the following steps to plot the null
distribution.

1. Use the `x_vals` vector below for the appropriate x-values.

2. Create an object called `y_vals` that has the density function values by 
using `dchisq(x_vals, df)` with the appropriate degrees of freedom. 

3. Use `plot(x_vals, y_vals)` to plot this null distribution. 

4. Add a vertical red line at the observed $\chi^2$ statistic value you calculated 
in exercise 1.2.

Finally, in the answer section, write down what you expect the p-value to be 
based on looking at the plot you created of the null distribution. 


**Answers**








```{r q1_3, message=FALSE, warning=FALSE, tidy=TRUE}

x_vals <- seq(0, 40)







```







$\\$  







**Exercise 1.4 (5 points)**: Now do step 4 of hypothesis testing by calculating 
the p-value using the `pchisq(obs_stat, df)` function, and print out the p-value. 

Hint: To determine which tail to use, think about what a small $\chi^2$ means 
and what a $\chi^2$ means with respect to the null hypothesis.


```{r q1_4, message=FALSE, warning=FALSE, tidy=TRUE}




```





$\\$  





**Exercise 1.5 (4 points)**: Based on the p-value, what do you conclude? Do these 
results affect how you will play rock-paper-scissors in the future? 


**Answer**









$\\$  





 

 
## Part 2: One-way Analysis of Variance (ANOVA) for testing more than two means


On homework 7 part 3, you examined whether movies that have particular Motion
Picture Association of America (MPAA) ratings (i.e., G, PG, PG-13 or R) are
enjoyed more by movie critics using data from over 500 movies randomly selected
from the [Rotten Tomatoes website](https://www.rottentomatoes.com/). To do your
analysis, you ran a randomization test that created a null that was based on
using the MAD statistic. In the exercises below, you will reanalyze this
question, however instead of running a randomization test, you will run a
one-way analysis of variance (ANOVA) to determine whether several means are the
same.

As you will recall from homework 7, the code below loads the movie data and
creates two vectors. The first vector is called `mpaa_rating` and has the MPAA
rating for each movie (i.e., the vector contains categorical data with the
levels of: G, PG, PG-13, and R). The second vector is called `critics_score` and
contains ratings critics gave to movies on a scale from 1 to 100. You will use
these two vectors (`mpaa_rating` and `critics_score`) to examine whether the
average (mean) movie scores differ depending on the rating MPAA rating level. If
you are interested in learning more about this full dataset a codebook
describing the variables can be found on 
[this website](http://www2.stat.duke.edu/~mc301/data/movies.html).






$\\$





**Exercise 2.0 (2 points)**:

To start your analysis, report the sample size *n* for these two vectors, 
`mpaa_rating` and `critics_score`, using the `length()` function. 


```{r get_rotten_tomatoes_data}


# load the data
load('movies.Rda')


# only keep movies rated "G", "PG", "PG-13", "R" (you can ignore this)
movies2 <- movies[movies$mpaa_rating %in% c("G", "PG", "PG-13", "R"), ]
movies2$mpaa_rating <- droplevels(movies2$mpaa_rating)


# get the MPAA ratings and the rotten tomatoes critic scores 
# you will only use these two vectors for the rest of these exercises
mpaa_rating <- movies2$mpaa_rating
critics_score <- movies2$critics_score


# display the sample size of the two vectors you will use in your analyses





```






$\\$






**Exercise 2.1 (3 points)**:  Now start your hypothesis test with step 1 by 
stating the null and alternative hypotheses in symbols and words. Also state the
alpha level that is most commonly used. 



**In words**





**In symbols**






**The significance level**







$\\$







**Part 2.2 (4 points)**: In order for the F-distribution to be a valid null 
distribution for our F-statistic, two conditions (i.e., assumptions) must hold.
The two conditions that need to be met are: 1) the data from each group must be
relatively normal, and 2) the variances (or standard deviations) in each group
must be approximately the same.

We can check the first condition as to whether the data is relatively normal by
creating box plots of the data. As long as the boxplots look reasonably
symmetric and there are not too many extreme outliers this condition should be
met. Please use the R chunk below to create a boxplot of our data by group (in
this case, rating). Remember, we can create a box plot comparing several groups
using the syntax `boxplot(data ~ grouping)`. Then in the answer section report
whether the data in each group looks relatively normal.

We can check the second condition as to whether the variances within each group
are approximately the same by comparing the standard deviations between the
groups. As long as the largest standard deviation is not twice as big as the
smallest standard deviation this condition is met. The code below uses the
`by()` function to calculate the standard deviation separately for each group
(the `by(data, grouping, function)` works by taking three arguments and applies
the `function` arguments separately to `data` separately for each `grouping`).
In the answer section below, describe whether the output of the `by()` function
indicates whether the equal variance condition is met. 



**Answer**






```{r check_anova_conditions}


# create a boxplot comparing the groups



# calculating the standard deviation separately for each group
by(critics_score, mpaa_rating, sd)


```







**Part 2.3 (3 points)**:  Next let's compute the F-statistic that we discussed 
in class to compare the mean critic scores between the different MPAA rating
levels. To do this we can use the `get_F_stat(quantitative_data, grouping_data)`
function from the SDS1000 library. Save the F statistic value to an object
called `obs_stat` and report the value of this statistic below.

```{r movies_obs_stat}


# Calculate the F statistic




```

**Answer:** 





$\\$






**Exercise 2.4 (6 points)**: Now let's do step 3 of the hypothesis test by 
plotting the null distribution. Start by calculating the values of the two
parameters, degrees of freedom 1 and 2, and write down their values in the
answer section below. Then create your y-values using `df(x_vals, df1, df2)`,
using the `x_vals` object that is given in the R chunk. Plot your null
distribution with `plot(x_vals, y_vals, type = "l")`. Finally, add the observed
statistic as a red vertical line to the plot. Based on looking at this null
distribution, what do you think the p-value is?




**Answer**

The degrees of freedom are: 






```{r F_null_distribution}

# use these x values to plot the density function
x_vals <- seq(0, 10, by = .01)


# created the denisty function values and plot them 



# add a vertical line at the value of the observed statistic




```




**Answers** 







$\\$






**Exercise 2.5 (4 points)**: Let's do step 4 of the hypothesis test by 
calculating the p-value using the `pf(obs_stat, df1, df2)` function. Report what 
the p-value is (and make sure you look at the correct tail). Is this close to 
what you estimated by looking at the null distribution above? 


```{r p_value}



```


**Answers** 






$\\$




**Exercise 2.6 (3 points)**:  Now complete step 5 of hypothesis testing by 
making a judgement. Are you able to reject the null hypothesis? What do you conclude? 


**Answer** 





$\\$






**Exercise 2.7 (5 points)**: R also has a built in ANOVA function that can do 
all the steps of an ANOVA for you. To create an ANOVA table using this function 
requires doing two steps: 

1. You need to fit the model using the syntax:  `fit <- aov(data ~ grouping)`

2. You can then print the ANOVA table using `summary(fit)`

Please run the code described above to create a ANOVA table in the R chunk below
and print out this ANOVA table. Do the results match what you got above? (Note
that the p-value is labelled 'Pr(>F)' in the summary output.)


**Answer**  Do the results match? 





```{r r_anova}





```




$\\$





**Exercise 2.8 (6 points)**:: As mentioned above, on homework 7 you ran a 
hypothesis test examining this same question of whether critics scores differed
based on MPAA rating but you did it using a permutation test using the MAD
statistic. Please repeat this permutation test analysis by running hypothesis
test step 3-5 in the R chunk below, but use the F-statistic instead of the MAD
statistic (i.e., use the `get_F_stat()` function, etc.). Be sure to display your
null distribution as a histogram, and put a red vertical line on the null
distribution at the value of the observed statistic. Do you come to the same
conclusions?


**Answer**





```{r F_rand_null_distribution}


# create the null distribution using permutation (randomization) methods






# plot the null distribution with a red vertical line for the statistic value





# calculate the p-value




```







$\\$
 




 
## Part 4: Final project data and question (5 points)


Please state the question you will address in your final project and where you 
will get the data from. Also, load the data into R in the chunk below to verify 
that you can get the data into R. 

 
```{r message=FALSE, warning=FALSE, tidy=TRUE}



```




$\\$ 
 
 
 

## Reflection (3 points)


Please reflect on how the homework went by going to Canvas, going to the Quizzes 
link, and clicking on Reflection on homework 10.




 
 
 
 
 
 
 
 
 
 
  
