## Practice: Hypothesis test for the difference of two means


A study is interested to check if the mean exercise hours for female are less than the mean exercise hours for male students. Use data `ExerciseHours` and the two variables `Exercise` and `Sex`.

1.) **Step 1**:  Write the `null hypothesis`  and `alternative hypothesis`  in words and in symbols.

*a.)* Create a boxplot to describe hours of exercise for `female` versus `male`.

```{r}
# your code here
```

*b.)* Find some favorites statistics of `Exercise hours` for female and male students. You might find the function: `mosaic::favstats` useful.
*Note*: you can search online for this function arguments.

```{r}
#your code here
```
 
*c.)* Subset the data `ExerciseHours` to two groups: `F` and `M`.

$\\$

```{r}
#your code here
```

2.) **Step 2**: Compute the observed statistic (mean difference of exercise hours for Female and Male).

```{r}
#your code here
```

$\\$

3.) **Step 3**: Create null hypothesis distribution

*a.)* Shuffle the two groups of `female` and `Male` into two samples, and find the mean difference of the two shuffled samples.

*b.)* Create the Null hypothesis Distribution using `do_it()` function.  

*c.)* Plot a `histogram' of the null distribution and show the `line` of the `observed mean difference` using the `abline()` function.

```{r}
# your code here
```

$\\$

4.) **Step 4:** Calculate p-value 

```{r}
# your code here
```

$\\$

**Step 5:**  Make decision/Judgment

```{r}
#your code here
```

$\\$

**Answers:**

```{r}
library(Lock5Data)
library(SDS1000)
data(ExerciseHours)
```

1.) **Step 1**:  Write the `null hypothesis`  and `alternative hypothesis`  in words and in symbols.

$\ H_0:\mu_{f}= \mu_{m}$   vs    $\ H_a: \mu_{f} < \mu_{m}$


*a.)* Create a boxplot to describe hours of exercise for `female` versus `male`.


```{r}

boxplot(Exercise ~ Sex , data = ExerciseHours)

```

*b.)* Find some favorites statistics of `Exercise hours` for female and male students. You might find the function: `mosaic::favstats` useful.

*Note*: you can search online for this function arguments.

```{r}

mosaic::favstats( Exercise ~ Sex, data = ExerciseHours)
```

*c.)* Subset the data `ExerciseHours` to two groups: `F` and `M` using `subset()` function.

```{r}
# we will use the function `subset`
excercise_fem<- subset(  ExerciseHours$Exercise, ExerciseHours$Sex == "F")
excercise_fem
excercise_mal<- subset(  ExerciseHours$Exercise, ExerciseHours$Sex == "M")
excercise_mal

length(excercise_fem)
length(excercise_mal)
## 30
## 20
```

2.) **Step 2**: Compute the observed statistic (mean difference of exercise hours for Female and Male).

```{r}

obs_stat <- mean(excercise_fem) - mean(excercise_mal)
obs_stat
## -3
```

3.) **Step 3:** Create null hypothesis distribution

*a.)* Shuffle the two groups of `female` and `Male` into two samples, and find the mean difference of the two shuffled samples.

```{r}
combined_sample <- c(excercise_fem, excercise_mal)
shuffled_sample <- sample(combined_sample )

shuff_fem <- shuffled_sample[1:30]
shuff_mal <- shuffled_sample[31:50]

shuff_stat <- mean(shuff_fem) - mean(shuff_mal)
shuff_stat
# answers may vary
```

*b.)* Create the Null hypothesis Distribution  

```{r}

null_dist <- do_it(10000) * {
shuffled_sample <- sample(combined_sample )

shuff_fem <- shuffled_sample[1:30]
shuff_mal <- shuffled_sample[31:50]

shuff_stat <- mean(shuff_fem) - mean(shuff_mal)
}
```

*c.)* Plot histogram of the null distribution and show the line of the observed mean difference

```{r}
hist(null_dist , xlab = "Difference in means of Exercise hous bewteen Female and Male", main = "Null distribution")

abline(v = obs_stat, col = "blue")
```

4.) **Step 4:** Calculate p-value 

```{r}

p_value <- pnull(obs_stat, null_dist, lower.tail = T) 
p_value
#0.1038 (# answers may vary)
```

5.) **Step 5:**  Make decision/Judgment

```{r}
#There are no enough evidence to conclude that there is  a mean difference in Exercise Hours between Female and Male.
```
