---
title: "Class 18 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---







## Part 1: Generating random data from a normal distribution


IQ scores are normally distributed with a mean of 100 and a standard deviation of 15.

Let's generate/simulate 1,000 random IQ scores and visualize the distribution.



```{r normal_data}

set.seed(123)  # for reproducibility



# Generate random IQ scores



# Visualize the distribution




```





$\\$





## Part 2: Visualzing the distribution of IQ scores with a density plot


We can visualize the distribution of IQ scores using a density plot.


```{r density_plot}

# Get a range of x values


                
# Calculate the density values for the normal distribution




# Plot the density





```



We can also overlay the density plot on the histogram of the simulated IQ scores by normalizing the histogram so that the area sums to 1.




```{r overlay_density}

# Plot the histogram with density




# Overlay the density plot - lines() adds to the previous plot



```






$\\$





## Part 3: Calculating probabilities using the normal distribution

We can calculate probabilities using the normal distribution. For example, what is the probability that a randomly selected individual has an IQ score less than 120?


```{r calculate_probability}

# Calculate the probability









```


We can also get probability ranges by subtracting cumulative probabilities.

What is the probability that a randomly selected individual has an IQ score between 85 and 115?


```{r calculate_probability_range}

# Calculate the probability










```





$\\$





## Part 4: Finding IQ score cutoffs using quantiles


Before we get quantiles from a theoretical normal distribution, let's get
quantiles from our simulated IQ scores.

Recall that quantiles divide the data into equal parts. For example, the 0.25
quantile (or 25th percentile) is the value below which 25% of the data falls.

Also, recall that we can get quantiles in R using the `quantile()` function.
Let's try this now...


```{r quantiles_simulated}

# Let's get the 25th percentile for our simulated IQ scores





# get the quantiles for the middle 95% of our simulated IQ scores



```


Let's now get quantiles from the theoretical normal distribution of IQ scores.



```{r find_cutoff}

# Calculate the quantiles




# we can visualize this using the mosaic package xqnorm function


```


Can you get the range of IQ scores that the middle 95% of the population falls within?

Also, try it using the `mosaic::xcnorm()` function.


```{r find_95_cutoff}




```
