---
title: "Class 6 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---







# Part 1: Practice with scatter plots and correlation 

Let's continue to use the data collected by [Fraumeni (1968)](https://academic.oup.com/jnci/article/41/5/1205/903436) 
to examine the relationship between the number of cigarettes sold per capita
and the rates of different types of cancer. 

The data is loaded below into a data frame named `smoking`. Cancer rates are 
reported as the number of deaths per 100,000 citizens. 


```{r smoking_data}

load("states_smoking.rda")

```



$\\$



### Part 1.1: Practice with scatter plot and correlation coefficient


Last class we examined the relationship between cigarettes sold per capita and
lung cancer rates and saw that there seemed to be a pretty strong association.

Let's get a little more practice with scatter plots and correlation by examining
the relationship between cigarettes sold per capita (`CIG`) and the rate of 
bladder cancer (`BLAD`) and/or leukemia (`LEUK`) by doing the following:


1. Create scatter plots of the relationships between cigarettes sold per capita
and these types of cancer. Based on the scatter plots, guess what the
correlation values are.

2. Calculate the correlation values between cigarettes sold per capita and these
types of cancer. Do the correlation values match your expectations based on the
scatter plots?

3. Based on these results, do you think there is a real relationship between the
number of cigarettes sold per capita and these types of cancer?



```{r smoking_additional_explorations}

# Scatter plot of CIG vs BLAD
plot(smoking$CIG, smoking$BLAD,
     xlab = "Cigarettes sold per capita (1960)",
     ylab = "Bladder cancer deaths per 100,000",
     main = "Scatter plot of cigarettes sold vs bladder cancer deaths")


# Scatter plot of CIG vs LEUK
plot(smoking$CIG, smoking$LEUK,
     xlab = "Cigarettes sold per capita (1960)",
     ylab = "Leukemia deaths per 100,000",
     main = "Scatter plot of cigarettes sold vs leukemia deaths")


# Correlation between CIG and BLAD
cor(smoking$CIG, smoking$BLAD)


# Correlation between CIG and LEUK
cor(smoking$CIG, smoking$LEUK)


```





$\\$






# Part 2: Linear regression


Let's continue analyzing the relationship between cigarettes sold per
capita (`CIG`) and lung cancer rates (`LUNG`) but now using linear regression.




$\\$




### Part 2.1: Fitting a linear model

To start, let's fit a linear model predicting lung cancer deaths from the number of
cigarettes sold per capita. We can then recreate our scatter plot and add the regression
line to the plot.


```{r linear_regression}

# Fit a linear model predicting lung cancer deaths from cigarettes sold
lung_model <- lm(smoking$LUNG ~ smoking$CIG)

# Plot the data and the regression line
plot(smoking$CIG, smoking$LUNG,
     xlab = "Cigarettes sold per capita (1960)",
     ylab = "Lung cancer deaths per 100,000",
     main = "Scatter plot of cigarettes sold vs lung cancer deaths")


abline(lung_model, col = "red")

```




$\\$





### Part 2.2: Interpreting the model output

We can use the `coef()` function to extract the coefficients from the model.

Please do the following:

1. Extract the coefficients from the model using the `coef()` function.

2. Write down the equation of the regression line 
for predicting lung cancer deaths from the number of cigarettes sold per capita. 

3. Then interpret the coefficients in the context of this problem.



```{r linear_regression_coefficients}

coef(lung_model)

```

**Answer:**

1. The coefficients from the model are approximately:
   - Intercept: 6.47
   - Slope: 0.0053

2. The equation of the regression line is: $\hat{y} =  0.0053x + 6.47$, where $\hat{y}$ is the predicted number of lung cancer deaths per 100,000 and $x$ is the number of cigarettes sold per capita.

3. The intercept (6.47) represents the estimated number of lung cancer deaths per 100,000 when no cigarettes are sold per capita. This value may not have a practical interpretation in this context, as it is unlikely that there would be zero cigarette sales. 

The slope (0.0053) indicates that for each additional cigarette sold per capita, the number of lung cancer deaths per 100,000 is expected to increase by 0.0053. This suggests a positive association between cigarette sales and lung cancer deaths.




$\\$





### Part 2.3: Making predictions

Next, let's use the regression model to make a prediction.

To predict the number of lung cancer deaths per 100,000 when 2,000 cigarettes
are sold per capita, substitute $x = 2000$ into the regression equation and/or
use R!


```{r linear_regression_prediction}

the_coefs <- coef(lung_model)

the_coefs[1] + the_coefs[2] * 2000


```




**Answer:**


$\hat{y} = 0.0053(2000) + 6.47 = 10.6 + 6.47 = 17.07$

So, the predicted number of lung cancer deaths per 100,000 when 2,000 cigarettes are sold per capita is approximately 17.1.




$\\$





### Part 2.4: Exploring the residuals

Finally, let's explore the residuals from the model by doing the following:

1. Let's calculate the residual for the first state (Alabama) in the data set.

2. Let's use the `residuals()` function to extract all the residuals from the model.

3. Let's calculate the sum of the squared residuals. R selects the optimal 
intercept (`a`) and slope (`b`) to minimize this quantity. 

To explore finding the optimal `a` and `b` by hand, try the
[simple regression app](https://gallery.shinyapps.io/simple_regression/).



```{r linear_regression_residuals}

# Calculate the residual for Alabama
AL_residual <- smoking$LUNG[1] - (the_coefs[1] + the_coefs[2] * smoking$CIG[1])

AL_residual


# Extract all residuals
all_residuals <- residuals(lung_model)

# Calculate the sum of squared residuals
sum(all_residuals^2)


```


**Answer:**

1. The residual for Alabama is about 0.95, which means Alabama's actual lung
cancer death rate was about 0.95 deaths per 100,000 higher than the model
predicts from its cigarette sales.

2. The `residuals()` function returns all 44 residuals, one for each state.

3. The sum of the squared residuals is about 394.8. This is the quantity that
least squares regression minimizes: no other choice of intercept and slope would
give a smaller value for this data set.



