---
title: "Class 6 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---






# Part 1: Scatter plots and correlation


Let's use the data collected by [Fraumeni (1968)](https://academic.oup.com/jnci/article/41/5/1205/903436) 
to examine the relationship between the number of cigarette's sold per capita
and the rates of different types of cancer. 

The data is loaded below into a data frame named `smoking`. Cancer rates are 
reported as the number of deaths per 100,000 citizens. 


```{r smoking_data}

load("states_smoking.rda")

```




$\\$





### 1.1 Scatter plots

Start by creating a scatter plot between the number of cigarette's sold per 
capita (`CIG`) and the number of lung cancer deaths (`LUNG`). 


```{r smoking_scatter_plot}





```



$\\$




### 1.2 Scatter plots

Now let's examine the correlation between the number of cigarette's sold per 
capita (`CIG`) and the number of lug cancer deaths (`LUNG`). 

```{r smoking_correlation}




```




$\\$



### 1.3 Additional explorations

Do you think there is a correlation between the number of cigarette's sold per
capita (`CIG`) and the rate of bladder cancer (`BLAD`) and/or luekemia (`LEUK`)?

Try the following to explore these questions:

1. Create scatter plots of these relationships cigarette's sold per and these 
types of cancer. Based on the scatter plot guess what the correlation values are.

2. Calculate the correlation values between cigarette's sold per capita and these
types of cancer. Do the correlation values match your expectations based on the
scatter plots?

3. Based on these results, do you think there is a real relationship between the
number of cigarette's sold per capita and these types of cancer?



```{r smoking_additional_explorations}

# Scatter plot of CIG vs BLAD




# Scatter plot of CIG vs LEUK





# Correlation between CIG and BLAD



# Correlation between CIG and LEUK



```






$\\$







# Part 2: Linear regression


Let's continue to use the smoking data to explore linear regression.


$\\$



### Part 2.1: Fitting a linear model

To start, let's fit a linear model predicting lung cancer deaths from the number of
cigarettes sold per capita. We can then recreate our scatter plot and add the regression
line to the plot.


```{r linear_regression}

# Fit a linear model predicting lung cancer deaths from cigarettes sold


# Plot the data and the regression line






```




$\\$





### Part 2.2: Interpreting the model output

We can use the `coef()` function to extract the coefficients from the model.

Please do the following:

1. Extract the coefficients from the model using the `coef()` function.

2. Write down the equation of the regression line 
for predicting lung cancer deaths from the number of cigarettes sold per capita. 

3. Then interpret the coefficients in the context of this problem.



```{r linear_regression_coefficients}



```

**Answer:**

1.



2. 



3. 






$\\$





### Part 2.3: Making predictions


Next, let's use the regression model to make a prediction.

Please predict the number of lung cancer deaths per 100,000 when 2,000
cigarettes are sold per capita, we can substitute $x = 2000$ into the regression
equation and/or use R! 


```{r linear_regression_prediction}






```




**Answer**







$\\$






### Part 2.4: Exploring the residuals


Finally, let's explore the residuals from the model by doing the following:

1. Let's calculate the residual for the first state (Alabama) in the data set.

2. Let's use the `residuals()` function to extract all the residuals from the model.

3. Let's caluculate the sum of the squared residuals. R selects the optimal 
intercept (`a`) and slope (`b`) to minimize this quantity. 

To explore finding the optimal `a` and `b` by hand, try the online app at: https://gallery.shinyapps.io/simple_regression/




```{r linear_regression_residuals}

# Calculate the residual for Alabama





# Extract all residuals



# Calculate the sum of squared residuals



```


     



