---
title: "Class 3 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---




# Part 1: Quick review of basics of R



### 1.1 Number journey 1!

1. Create an object called `year` and assign the value 2026
2. Create an object called `days` and assign 365.25 to it
3. Create an object called `days_since_bc` and assign `year` times `days` to it
4. Print out `days_since_bc` to "show your work"


```{r number_journey}








```




$\\$




### 1.2 Number journey 2!


1. Create a vector object called `my_vec` and assign the values 3, 4, and 5 to it
2. Take the square root of the values in `my_vec` and assign it to an object called `my_vec2` 
3. Multiple `my_vec2` by the number of elements in `my_vec2` and assign it to an object called `my_vec3` 
4. Sum the values in `my_vec3` and print out the result to show your work

```{r number_journey2}







```






$\\$






# Part 2: More practice analyzing categorical data


Let's get more practice analyzing categorical data in R! To do this let's look 
at the [Gapminder data set](https://www.youtube.com/watch?v=jbkSRLYSojo) which 
contains information about different countries. We will focus on data from the
year 2007 only, which is loaded below.


Key variables in the Gapminder data set include:

- `country`: The country of the observation
- `continent`: The continent of the observation
- `year`: The year of the observation
- `lifeExp`: Life expectancy at birth (in years)
- `pop`: Population of the country
- `gdpPercap`: Gross domestic product per capita (in dollars)



$\\$




### 2.1 Load the data

The code below loads the Gapminder data set and shows the number of rows and
columns in the data frame using the `dim()` function.


```{r load_gapminder_data}

load("gapminder_2007.rda")

# Show the number of rows and columns
dim(gapminder_2007)


```




$\\$




### 2.2 Extracting a vector of data from a data frame

We can extract a vector of data from a data frame using the syntax `df$my_var`,
where `df` is the name of the data frame and `my_var` is the name of the column
of data we want to extract. 

Let's start by creating an object called `continents` which contains the continent 
that each country is in.



```{r extract_cat_variable}

# Create an object called continents by extract data from the gapminder_2007 data frame




```




$\\$




### 2.3 More practice analyzing categorical data

What proportion of countries in the data set are in Africa? 

Hint: The `get_proportions(vec, category)` function could be useful


```{r}

library(SDS1000)






```




$\\$




### 2.4 Frequency tables and bar plots

Now create a frequency table for how many countries are in each continent in the
data and create a bar plot of the data. 


```{r practice_categorical_data}

# 1. Create a frequency table showing how many countries are in each continent





# 2. Create a barplot of the frequency of countries in each continent






```






$\\$




# Part 3: Analyzing a single quantitative variable

Let's now use the Gapminder data to practice analyzing quantitative data. 



$\\$



### 3.1 Extracting data from a data frame


Let's create the following vector object by extracting data from
the `gapminder_2007` dataframe:

- `life_expectancy`: should contains the life expectancy data (for the year 2007)
- `population`: should contains the population of each country in 2007 



```{r}

# Create a new variable for life expectancy




# Create a new variable for population



```




$\\$




### 3.2 Histograms

Let's now create a histogram of the life expectancy data for the year 2007. We
can do this using the `hist()` function along with the vector of life expectancy
data we created in part 2.1 which is stored in the object `life_expectancy`.


Questions:

1. What shape does this histogram have? Is it symmetric, right-skewed, or left-skewed?
2. Are there any extreme outliers in the data? 


```{r histogram_life_expectancy}

# Create a histogram of life expectancy





```


**Answer:** 

1. 

2. 





$\\$







# Part 4:  Descriptive statistics for the central tendency: the mean and median


We can use the `mean()` and `median()` functions to calculate the mean and median
of a sample of data. 

Let's calculate the mean and median of the life expectancy data for the year 2007.

Are these values what you would expect based on the skew of the data?

```{r mean_median_life_expectancy}

# Calculate the mean of life expectancy




# Calculate the median of life expectancy



```




$\\$




### 4.2 Showing the mean and median on a histogram


We can add lines to the histogram to show where the mean and median are located using
the `abline()` function.

Let's add lines for the mean and median to the histogram we created in part 2.1.


```{r add_mean_median_lines}

# Create a histogram of life expectancy



# Add a line for the mean



# Add a line for the median



# Add a legend


```





$\\$





### 4.3 Internet pings (try this at home)

When an individual visits a webpage a 'ping' is generated

Below is a random sample of ping counts from 7 people who pinged a website at least once:

   12, 45, 6, 4, 158, 10, 59 

Question: What is the mean and median ping count in this sample? 

Also, see if you can calculate these values without using the `mean()` and 
`median()` functions. Hint: the `sum()`, `length()`, and `sort()` functions may 
be useful.

```{r}

# Create the ping data as a vector 




# Calculate the mean ping count




# Calculate the median ping count





```




