---
title: "Homework 1"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---




$\\$





In the following homework, you will explore descriptive statistics and
visualizations of categorical data. Please render this Quarto document and
submit the compiled pdf document with your answers to Gradescope
*by 11pm on Sunday September 7th*. Please also make sure to mark the correct
pages to each question's answers on Gradescope to allow the TAs to grade each
problem.


Some functions that could be useful for completing this worksheet are: `dim()`,
`length()`, `table()`, `prop.table()`, `get_proportion()`, `barplot()`,
`hist()`, `mean()`, `median()`, `max()`. You might also find the following
symbols useful: $\mu$, $\bar{x}$, $\pi$, $\hat{p}$


If you need help with any of the homework assignments, please attend the TA
office hours which are listed on Canvas and/or ask questions on [Ed Discussions
forum](https://edstem.org/us/courses/78564/discussion). Also, please help
members of the class by answering questions on Ed Discussions. Finally, if you
have conceptual questions, reviewing the class slides and videos will be
helpful.






$\\$





# Part 1:  Lock 5 questions



Please answer the following questions that are based on questions from the Lock5
textbook. Some symbols that might be useful for answering these questions
include $\hat{p}$, $\bar{x}$, $\pi$, and $\mu$. You can also write subscripts in
LaTeX by using an underscore, e.g., $\beta_{treatment}$.






$\\$






**Exercise 1.1 (6 points)**:  A study by Hamscher et al. (Environmental Health
Perspectives, 2003) shows that antibiotics added to animal feed to accelerate
growth can become airborne. Some of these drugs can be toxic if inhaled and may
increase the evolution of antibiotic-resistant bacteria. The researchers
analyzed 20 samples of dust particles from animal farms. Tylosin, an antibiotic
used in animal feed that is chemically related to erythromycin, showed up in 16
samples. Based on this study please report:

a. What are the individual cases in this study?

b. What is the variable of interest in this study? 

c. What is the statistic of interest in this study, what symbol should be used
for this statistic, and what is the value of this statistic? Your choice of
statistic and symbol should be one of the statistics/symbol we used in class.




**Answers:**  

a. 



b. 



c. 





$\\$


**Exercise 1.2 (6 points)**: A study by Valasek and Walter (Pediatric and 
Adolescent Concussions, Springer, 2012) examined all sports-related concussions
reported to an emergency room for children ages 5 to 18 in the United States
over the course of one year. The table below displays the number of concussions
in each of the major activity categories.


| Activity              | Number of concussions |
|-----------------------|-----------------------|
| Bicycles              |      23,405           |
| Football              |      20,293           |
| Basketball            |      11,507           |
| Playground            |      10,414           |
| Soccer                |       7,667           |
| Baseball              |       7,433           |
| All-Terrain Vehicles  |       5,220           |
| Hockey                |       4,111           |
| Skateboarding         |       4,408           |
| Swimming              |       3,846           |
| Horseback Riding      |       2,648           |
|-----------------------|-----------------------|
| Total                 |      100,952          |



Based on this study please report:

a. Are these results from a population or a sample. In 1-3 sentences please
explain your answer.

b. What proportion of concussions came from playing football? Please write down
your answer and denote it with the appropriate symbol (e.g., symbol = value). 

c. What proportion of concussions came from riding bicycles? Please write down
your answer and denote it with the appropriate symbol.

d. Can we conclude that, at least in terms of concussions, riding bicycles is
more dangerous to children in the US than playing football? Why or why not?




**Answers:**

a.  




b.  




c.  





d. 







$\\$




**Exercise 1.3 (6 points)**: A study by Reinhold et al. (Science, 2009) examined 
animal behavior and found that rats can quickly learn to play hide-and-seek with
humans. The rats seem to love the game, squeaking and jumping for joy when they
find the human or are found by the human. The game is played in a large room
with multiple boxes. Some of the boxes are transparent and easy to see through,
while others are opaque and the inside cannot be seen. Rats go in and out of
these boxes and the researchers studied whether they treated the boxes
differently depending on whether they were hiding (and didn't want to be seen) or
seeking (and didn't care if they were seen). For rats who had been learning the
game for about a week, the data on the number of times the rats entered the two
types of boxes in each situation is shown in the table below:


| Situation | Transparent boxes | Opaque boxes | Total |
|-----------|-------------------|--------------|-------|
| Hiding    |        15         |      38      |  53   |
| Seeking   |        17         |      14      |  31   |
| Total     |        32         |      52      |  84   |


Based on this study please report:

a. When the rats were hiding, what proportion of the time did they go into
opaque boxes?

b. When the rats were seeking, what proportion of the time did they go into
opaque boxes?

c. Find the difference in proportions and give the notation for the difference
in proportions using the notation we used in class. Hint: using subscripts on
the symbols could be useful (see the top of part 1 of the homework for how to
create subscripts in LaTeX). 



**Answers:**

a. 



b. 



c. 





$\\$






# Part 2: Oxford commas


The Oxford comma is a comma used before the conjunction in a list of three or more
items. For example, in the list "apples, oranges, and bananas," the Oxford comma
is the comma before the "and." As discussed in the book Eats Shoots & Leaves,
Oxford commas can clarify meaning and avoid ambiguity in sentences. However, to
paraphrase a song by [Vampire Weekend](https://youtu.be/P_i1xk07o4g?si=7aF-wGA1_B4W_60D), 
does anyone really care about the Oxford comma?

To address the question of whether most Americans prefer Oxford commas, the
website FiveThirtyEight conducted a survey where they asked people "How much, if
at all, do you care about the use (or lack thereof) of the serial (or Oxford)
comma in grammar?" They also collected other information about their
demographics and grammatical preferences from each survey participant. The
results were published in an article titled ["Elitist, Superfluous, Or Popular?
We Polled Americans on the Oxford Comma"](https://fivethirtyeight.com/features/elitist-superfluous-or-popular-we-polled-americans-on-the-oxford-comma/).

The R chunk below load the data that was collected in their survey into a data
frame called `comma_survey`. Each row (case) corresponds to one survey
respondent, and the variable we are most interested in is `care_oxford_comma`,
which indicates how much the respondent cares about the Oxford comma. The
possible values for this variable are "Not at all", "Not much", "Some",
and "A lot.

Let's analyze this data now to address the very important question of whether 
most Americans care about Oxford commas!


```{r load_oxford_comma_data, message = FALSE}

library(SDS1000)
library(fivethirtyeight)
library(dplyr)

data(comma_survey)

head(select(comma_survey, respondent_id, care_oxford_comma, everything()))

```




$\\$





**Exercise 2.1 (6 points):** To start, please extract the column `care_oxford_comma` 
as a vector from the `comma_survey` data frame. Assign this vector the to an
object called `oxford_care`.

Then, using the `oxford_care` vector, please print out how many people were surveyed 
in their study and report the number of people surveyed in the answer section as
well. Also, in answer section, please report what you believe the population of
interest is that the survey is trying to estimate.


```{r}





```


**Answer**: 








$\\$






**Exercise 2.2 (6 points):** Now that you have the `oxford_care` vector, please 
use the `table()` function on this object to create a count of how many people
care about the Oxford comma at each factor level of the variable. Assign the
results to an object called `oxford_comma_counts`. Finally, report which factor
level has the most responses.


```{r oxford_comma_counts}





```

**Answer**: 







$\\$





**Exercise 2.3 (6 points):** Next, please calculate the **relative frequencies**
(i.e., proportions) of how much each level of care was reported in the survey, and 
print out the results to "show your work". In the answer section, report what is 
the proportion of people who care "A lot" about the Oxford comma and what proportion 
of people care "Not at all"?

```{r oxford_comma_freqs}








```


**Answer**: 








$\\$






**Exercise 2.4 (6 points):** Now create a bar chart of the counts of how many people
care about the Oxford comma at each level of the variable. Also, suppose had conducted
another survey 2 years later which asked the same questions. In the answer
section, report whether you think the proportion of people who care about the
Oxford comma would be different in this new survey? Why or why not?



```{r barplot}







```



**Answers:**  










$\\$





**Exercise 2.5 (6 points):** Next create a pie chart of the proportions of how 
many people care about the Oxford comma at each level of the variable. In the
answer section, report whether you think the bar chart or pie chart is a better
way to visualize this data and why?


```{r piechart}






```


**Answer**: 









$\\$





**Exercise 2.6 (6 points):** Finally, the survey conducted by FiveThirtyEight also
collected information about several other variables. To see a list of these
variables, you can type `? comma_survey` in the R console, and you can view the
full `comma_survey` using the `View()` function. You can also learn more about
the survey by reading the original FiveThirtyEight article (link is at the top
of this part of the homework).

Please look through the `comma_survey` data frame and explore some of the other
variables in the data frame by visualizing them and/or calculating summary
statistics. In the R chunk below, please show your work exploring some of these
other variables, and in the answer section, please report what your analysis
shows.

Note: Our grading will be fairly relaxed on this question so do not stress too
much about coming up with the perfect analysis here, but do see if you can
find something interesting and explain why you think it is interesting. 


```{r explore_other_variables}











```

**Answer**: 







$\\$







## Reflection (6 points)


Please reflect on how the homework went by going to Canvas, going to the Quizzes link, and clicking on 
[Reflection on homework 1](https://yale.instructure.com/courses/109046/quizzes/89888)

Also, if you have not done so already, please be sure to fill out the class 
[background survey](https://yale.instructure.com/courses/109046/quizzes/89879).
Three points will be deducted from this homework if the background survey has not
been filled out by the time homework 1 is due. 





