---
title: "Class 11 notes and code"
format: pdf
editor: source
editor_options:
  chunk_output_type: console
---




#  Part 1: Simple example of sampling with replacement



```{r}


library(SDS1000)


# simple vector
my_sample <- c(3, 1, 4, 1, 5, 9)


# resampling values in the vector with replacement
boot_sample  <-  sample(my_sample,  6,  replace = TRUE)


# print the bootstrap sample (some values can appear more than once)
boot_sample


```




$\\$




#  Part 2: Simple fake data bootstrap example


```{r}

# simple bootstrap example

my_sample <- c(21, 29, 25, 19, 24, 22, 25, 26, 25, 29)


# calculate the x-bar statistic
obs_mean <- mean(my_sample)
obs_mean


# create the bootstrap distribution
bootstrap_dist <-  do_it(10000) * {

  curr_boot <- sample(my_sample, length(my_sample), replace = TRUE)
  mean(curr_boot)

}


# visualize the bootstrap distribution
hist(bootstrap_dist, 100,
     xlab = "Mean of the bootstrap sample",
     main = "Bootstrap distribution")


# calculate the standard error
SE_boot <- sd(bootstrap_dist)
SE_boot


# 95% confidence interval (2 is approximately cnorm(.95) = 1.96)
CI_lower <-  obs_mean  - 2 * SE_boot
CI_upper <-  obs_mean  + 2 * SE_boot


c(CI_lower, CI_upper)



```




$\\$



#  Part 3: Confidence interval for Gettysburg mean word length


```{r}


# set the random seed
set.seed(123)


# load the gettysburg data
load("gettysburg.Rda")


# store the word lengths in a vector object called word_lengths
word_lengths <- gettysburg$num_letters


# get a random sample from the word lengths

n <- 10

my_sample <- sample(word_lengths, size = n)



# calculate the x-bar statistic
obs_mean <- mean(my_sample)
obs_mean


# create the bootstrap distribution
bootstrap_dist <-  do_it(10000) * {

  curr_boot <- sample(my_sample, n, replace = TRUE)
  mean(curr_boot)

}


# visualize the bootstrap distribution
hist(bootstrap_dist, 100,
     xlab = "Mean word length of the bootstrap sample",
     main = "Bootstrap distribution of the mean word length")


# calculate the standard error
(SE_boot <- sd(bootstrap_dist))


# 95% confidence interval
CI_lower <-  obs_mean  - 2 * SE_boot
CI_upper <-  obs_mean  + 2 * SE_boot


c(CI_lower, CI_upper)



# Is the population parameter (mu) in our confidence interval?
mean(word_lengths)


```


Yes, the 95% confidence interval contains the population mean word length $\mu$
of about 4.29.






$\\$





#  Part 4: Confidence interval examples



$\\$




##  Confidence interval example 1: One true love?


A survey asked 2625 people whether they agreed with the statement:
"There is only one true love for each person"

1812 of the respondents disagreed.

Compute a 90% confidence interval for the proportion who agreed (i.e., the
proportion who believe there is only one true love for each person).




```{r}

library(SDS1000)


# Create the data set

believes_in_true_love <- c(rep(FALSE, 1812), rep(TRUE, 2625 - 1812))


# Calculate the proportion of people who believe in one true love
# Note: if you use the get_proportion function the word "TRUE" needs to be in quotes

(obs_prop <- get_proportion(believes_in_true_love, "TRUE"))




# Get an example of a bootstrap statistic

# We can use the sample(data_vec, n) to get a sample of length n:

boot_sample <- sample(believes_in_true_love, 2625, replace = TRUE)
(bootstrap_prop <- get_proportion(boot_sample, "TRUE"))



# Create a bootstrap distribution in R

boot_dist <- do_it(10000)  *  {

  boot_sample <- sample(believes_in_true_love, 2625, replace = TRUE)

  get_proportion(boot_sample, "TRUE")

}



# Visualize the bootstrap distribution

hist(boot_dist, 20,
     xlab = "Proportion who believe in one true love",
     main = "Bootstrap distribution")




# Get the bootstrap standard error SE*

(SE_boot <- sd(boot_dist))





# Create a 90% confidence interval

CI_lower <-  obs_prop  - cnorm(.90) * SE_boot
CI_upper <-  obs_prop  + cnorm(.90) * SE_boot

c(CI_lower, CI_upper)



```







$\\$






## Confidence interval example 2: Average body temperature


Let's create a 95% confidence interval for the mean body temperatures of humans
using a sample of 50 people from the data collected by Mackowiak et al (1992).

The data is loaded below from the Lock5 package. Start by visualizing the data
then please create a 95% confidence interval for the mean body temperature using
the bootstrap!



```{r bodytemp_viz}

# Load the data
library(Lock5Data)

body_temps <- BodyTemp50$BodyTemp




# Start by visualizing the data

hist(body_temps, main = "Body Temperatures",
     breaks = 20,
     xlab = "Temperature (°F)", col = "lightblue", border = "black")


```




Now create a 95% confidence interval for the mean body temperature using the
bootstrap!

See if you can also create an 80% and 99% confidence interval as well!




```{r bodytemp_CI}


# Calculate the observed mean

(obs_mean <- mean(body_temps))





# Create a bootstrap distribution in R

boot_dist <- do_it(10000)  *  {

  boot_sample <- sample(body_temps, length(body_temps), replace = TRUE)

  mean(boot_sample)

}



# Visualize the bootstrap distribution

hist(boot_dist, 100,
     xlab = "Mean body temperature of the bootstrap sample (°F)",
     main = "Bootstrap distribution")




# Get the bootstrap standard error SE*

(SE_boot <- sd(boot_dist))




# Create a 95% confidence interval (2 is approximately cnorm(.95) = 1.96)

CI_lower <-  obs_mean  - 2 * SE_boot
CI_upper <-  obs_mean  + 2 * SE_boot

c(CI_lower, CI_upper)



# Create a 80% confidence interval

(CI_80 <-  c(obs_mean - cnorm(.80) * SE_boot, obs_mean + cnorm(.80) * SE_boot))



# Create a 99% confidence interval

(CI_99 <-  c(obs_mean - cnorm(.99) * SE_boot, obs_mean + cnorm(.99) * SE_boot))




```



Based on your confidence intervals, is 98.6$^\circ$F (the value that is commonly
quoted as the average human body temperature) a plausible value for the mean body
temperature?


**Answer:** No. 98.6$^\circ$F is above all of these confidence intervals (even the
99% interval), so it is not a plausible value for the mean body temperature; the
data suggest that the mean body temperature is a little lower than 98.6$^\circ$F.



