---
title: "Class 11 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---




# Part 1: Normal quantiles

We use quantiles from a standard normal distribution to create confidence
intervals for different confidence levels.

For a given confidence level *C*, we want to find the quantile value q* such
that C percent of the area of a standard normal distribution is between -q* and
q*.



```{r}

library(SDS1000)











```





$\\$




#  Part 2: Simple example of sampling with replacement



```{r}


library(SDS1000)


# simple vector
my_sample <- c(3, 1, 4, 1, 5, 9)



# resampling values in the vector with replacement







```




$\\$




#  Part 3: Simple fake data bootstrap example


```{r}

# simple bootstrap example

my_sample <- c(21, 29, 25, 19, 24, 22, 25, 26, 25, 29)



# calculate the x-bar statistic



# create the bootstrap distribution






# visualize the bootstrap distribution




# calculate the standard error





# 95% confidence interval






```




$\\$






#  Part 4: Confidence interval examples




$\\$






## Confidence interval example 1: One true love? 


A survey asked 2625 people whether they agreed with the statement: 
"There is only one true love for each person"

1812 of the respondents disagreed.

Compute a 90% confidence interval for the proportion who disagreed.





```{r}

library(SDS1000)


# Create the data set

belives_in_true_love <- c(rep(FALSE, 1812), rep(TRUE, 2625 - 1812))



# Calculate the proportion of people who believe in one true love
# Note: if you use the get_proportion function the word "TRUE" needs to be in quotes





# Get an example of a bootstrap statistics 

# We can use the sample(data_vec, n) to get a sample of length n:








# Create a bootstrap distribution in R









# Visualize the bootstrap distribution






# Get the bootstrap standard error SE*






# Create a 90% confidence interval








```







$\\$






## Confidence interval example 2: Average body temperature


Let's create a 95% confidence interval for the mean body temperatures of humans
using a sample of 50 people from the data collected by Machowaik et al (1992). 

The data is loaded below from the Lock5 package. Start by visualizing the data
then please create a 95% confidence interval for the mean body temperature using
the bootstrap!



```{r bodytemp_viz}

# Load the data 
library(Lock5Data)

body_temps <- BodyTemp50$BodyTemp


# Start by visualizing the data






```




Now create a 95% confidence interval for the mean body temperature using the
bootstrap!

See if you can also create a 80% and 99% confidence interval as well!




```{r bodytemp_CI}


# Calculate the observed mean




# Create a bootstrap distribution in R






# Visualize the bootstrap distribution




# Get the bootstrap standard error SE*



# Create a 95% confidence interval





# Create a 80% confidence interval




# Create a 99% confidence interval





```








