---
title: "Class 7 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---




# Part 1: Linear regression


Let's continue to use the smoking data to explore linear regression.

We will continue to use the data collected by [Fraumeni (1968)](https://academic.oup.com/jnci/article/41/5/1205/903436) 
to examine the relationship between the number of cigarette's sold per capita
and the rates of different types of cancer. 

The data is loaded below into a data frame named `smoking`. Cancer rates are 
reported as the number of deaths per 100,000 citizens. 


```{r smoking_data}

load("states_smoking.rda")

```




$\\$




### Part 1.1: Fitting a linear model

To start, let's fit a linear model predicting lung cancer deaths from the number of
cigarettes sold per capita. We can then recreate our scatter plot and add the regression
line to the plot.


```{r linear_regression}

# Fit a linear model predicting lung cancer deaths from cigarettes sold




# Plot the data and the regression line





# Add the regression line to the plot




```


$\\$




### Part 1.2: Interpreting the model output

We can use the `coef()` function to extract the coefficients from the model.

Please do the following:

1. Extract the coefficients from the model using the `coef()` function.

2. Write down the equation of the regression line 
for predicting lung cancer deaths from the number of cigarettes sold per capita. 

3. Then interpret the coefficients in the context of this problem.



```{r linear_regression_coefficients}



```

**Answer:**

1. 



2. 




3. 






$\\$




### Part 1.3: Making predictions

Next, let's use the regression model to make a prediction.

Please predict the number of lung cancer deaths per 100,000 when 2,000
cigarettes are sold per capita, we can substitute $x = 2000$ into the regression
equation and/or use R! 


```{r linear_regression_prediction}






```




**Answer**








$\\$






### Part 1.4: Exploring the residuas

Finally, let's explore the residuals from the model by doing the following:

1. Let's calculate the residual for the first state (Alabama) in the data set.

2. Let's use the `residuals()` function to extract all the residuals from the model.

3. Let's caluculate the sum of the squared residuals. R selects the optimal 
intercept (`a`) and slope (`b`) to minimize this quantity. 

To explore finding the optimal `a` and `b` by hand, try the online app at: https://gallery.shinyapps.io/simple_regression/



```{r linear_regression_residuals}

# Calculate the residual for Alabama




# Extract all residuals




# Calculate the sum of squared residuals





```













