---
title: "Class 21 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---





# Part 1: Warm-up - Parametric hypothesis test for a proportion

Background:

Most people are right handed, right eye dominant, etc.  Biologists have
suggested that human embryos tend to turn their heads to the right as
well.German bio-psychologist Onur Güntürkün conjectured that this tendency
manifests itself in other ways, so he studies which ways people turn their heads
when they kiss.

In a study of 124 couples kissing, 80 of the couples turned their heads to the
right.  Is there evidence that more than half of all couples turn their heads to
the right when they kiss? Let's run a parametric hypothesis test for a
proportion to find out!




$\\$


Step 1: State the null and alternative hypotheses

In words: 

- Null hypothesis: 

- Alternative hypothesis: 



In symbols:







```{r hypothesis_test_proportion}



# Step 2: Calculate the z-statistic

# first assess that np and n(1-p) are both at least 10



# calculate the z-statistic






# Step 3: Visualize the null distribution (optional)








# Step 4: Calculate the p-value







```



Step 5: Make a decision








$\\$




Note: we could also get a p-value without converting to a z-statistic (and if
you take additional statistics/probability courses, you will learn you could
also use a binomial distribution as the null distribution too).

```{r hypothesis_test_proportion_alternative}






```



  




$\\$






# Part 2: Calculating quantities from a t-distribution


Suppose we have a sample size is n = 16, please calculate the following values: 

1. The 2.5th and 97.5th percentiles from a t-distribution

2. The probability that a t-statistic is more than 1.5

3. Also, calculate these same values for the standard normal distribution


```{r t_distribution_calculations, message = FALSE}


# Calculate the degrees of freedom




# 1. Calculate the 2.5th and 97.5th percentiles from a t-distribution





# We can also use the mosaic::ct() function to do this calculation








# 2. Calculate the probability that a t-statistic is more than 1.5










# 3. Calculate the same values for the standard normal distribution











```





$\\$





# Part 3: Parametric confidence interval for a single mean



A study by Loyd et al (2013) in Biological Conservation, used KittyCams to
record all activity of n = 55 domestic cats that hunt outdoors.

The video footage showed that the mean number of kills per week for these cats
was 2.4 with a standard deviation of 1.51.

Find and interpret a 99% confidence interval for the mean number of kills per
week by US household cats that hunt outdoors.


$$CI ~ = ~ \bar{x} ~ \pm ~ t^* \cdot \frac{s}{\sqrt{n}}$$



```{r confidence_interval_mean}

# Given data
n <- 55
x_bar <- 2.4
s <- 1.51


# Degrees of freedom




# Calculate the standard error (SE)





# Find the critical t-value for the desired confidence level







# Calculate the margin of error (ME)






# Construct the confidence interval






```







$\\$







# Part 4: Parametric hypothesis test for a single mean


In the mid-1990s a Nabisco marking campaign claimed that there were at least
1000 chips in every bag of Chips Ahoy! cookies.

A group of Air Force cadets tested this claim by dissolving the cookies from 42
bags in water and counting the number of chips.

They found the average number of chips per bag was 1261.6, with a standard
deviation of 117.6 chips

Test whether the average (mean) number of chips per bag is greater than 1000.  

Do the results confirm Nabisco’s claim?  



$\\$

Step 1: State the null and alternative hypotheses


In words: 

- Null hypothesis: 
- Alternative hypothesis: The average number of chips per bag is greater than 1000.


In symbols:  






```{r hypothesis_test_mean}



# Step 2: Calculate the t-statistic







# Step 3: Visualize the null distribution








# Step 4: Calculate the p-value






```



Step 5: Make a decision






Does this confirm Nabisco's claim? 











