---
title: "Practice Session 2 Answers"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---

```{r}
library(SDS1000)
```



# Part 1: Measures of central tendency for quantitative data	


## 1.1 Calculating the Sample Standard Deviation by Hand


### Here is the formula for the sample mean:

$$\bar{x} = \frac{1}{n} \sum_{i = 1}^n x_i$$



### And here is the formula for the sample standard deviation:

$$s = \sqrt{\frac{1}{n - 1} \sum_{i = 1}^n (x_i - \bar{x})^2}$$



#### Using the following data, complete the table for calculating the sample standard deviation.



|Cost(Dollars) |b. Deviations: $x_i - \bar{x}$  | c. Deviations squared: $(x_i - \bar{x})^2$|
|--------------|--------------------------------|-----------------------------------------  |
| 850          | -151.5                         | 22952.25                                  |
| 900          | -101.5                         | 10302.25                                  |
| 1400         | 398.5                          | 158802.25                                 |
| 1200         | 198.5                          | 39402.25                                  |
| 1050         | 48.5                           | 2352.25                                   |
| 750          | -251.5                         | 63252.25                                  |
| 1250         | 248.5                          | 61752.25                                  |
| 1050         | 48.5                           | 2352.25                                   |
| 565          | -436.5                         | 190532.25                                 |
| 1000         | -1.5                           | 2.25                                      |
                                                                   

a. The mean of the costs: $$\bar{x} =  1001.5$$


d. Write the sum of squared deviations value here: $$\sum_{i=1}^n (x_i - \bar{x})^2 =  551702.5     $$


e. Write the value of the squared deviations divided by $n-1$ here: 
$$\frac{\sum_{i=1}^n (x_i - \bar{x})^2}{n-1} =  61300.28$$


f. Write the value for the square root of the sum of squared deviations here: 
$$s = \sqrt{ 61300.28}=   247.5889   $$


$\\$



## 1.2 Five-Number Summary

Using the numbers from the previous exercise, do the following:

* Find the 5-number summary (minimum, Q1, median, Q3, maximum)

* Check your work using R functions 


**Answers**

```{r}

v <- c( 565, 750, 850, 900, 1000, 1050, 1050, 1200, 1250, 1400 )
v

## you can use the function fivenum()

fivenum(v)


```


$\\$



## 1.3 Boxplots and Histograms

Consider the `mtcars` data set. This dataset is built into R, so you can access it directly; no downloads required!

You can learn more about this dataset using `?` in the console or the `View()` function. You can also use the functions `head()` and `tail()`. 

1. First, create a histogram of the variable `mpg`. Then,  create a boxplot of `mpg`. How do these two plots compare? 

2. What proportion of cars have fuel consumption less than(or equal to) 15 mpg? 

3. What proportion of cars have fuel consumption more than(or equal to) 35 mpg? 

4. Create a boxplot of `mpg` by number of cylinders `cyl`.




**Answer**

```{r}

#1# Create a histogram of 'mpg'
dim(mtcars)

hist(mtcars$mpg, breaks= 10, xlim = c(5, 35), 
     main = "Histogram of Miles Per Gallon (mpg)",
     xlab = "Miles Per Gallon (mpg)",
     ylab = "Frequency",
     col  = "lightblue",
     border= "darkblue")


#1 Create a boxplot of 'mpg':

boxplot(mtcars$mpg, 
        main = "Boxplot of Miles Per Gallon (mpg)",
        xlab = "Miles Per Gallon (mpg)",
         col = "lightgreen")





#2 proportion of cars have a fuel consumption of less than(or equal to) 15 mpg 

prop_less_15<- ptail(15, mtcars$mpg )
prop_less_15 


#3 proportion of cars have a fuel consumption more than(or equal to)35 mpg 

prop_more_35 <- ptail(35, mtcars$mpg, lower.tail = FALSE )
prop_more_35



#4# create boxplot of mpg per cylinder: 


mtcars$cyl <- as.factor(mtcars$cyl)

boxplot(mpg ~ cyl, data = mtcars,
        main = "Miles Per Gallon by Number of Cylinders",
        xlab = "Number of Cylinders",
        ylab = "Miles Per Gallon",
        col  = c( "lightblue", "pink", "lightgreen"))
      


## method 2 to create a boxplot mpg per cylinder: 

cyl_4 <- subset(mtcars$mpg, mtcars$cyl=="4") # create a vector of mpg for cars with cyl =4
cyl_4


cyl_6 <- subset(mtcars$mpg, mtcars$cyl=="6") # create a vector of mpg for cars with cyl =6
cyl_6


cyl_8 <- subset(mtcars$mpg, mtcars$cyl=="8") # create a vector of mpg for cars with cyl =8
cyl_8

boxplot(cyl_4,cyl_6, cyl_8, 
        names = c( "4 cylinders", "6 cylinders", "8 cylinders"), 
         col  = c( "brown1", "brown3", "brown4"),
         xlab = "Number of Cylinders", 
         ylab = "Miles per Galon " )



      

```

The histogram shows the overall shape and spread of `mpg`, revealing that the distribution is slightly right-skewed with most cars between about 15 and 25 mpg. The boxplot summarizes the same data with the five-number summary (median, quartiles, and whiskers), making it easier to quickly identify the center, spread, and any potential outliers, but it does not show the shape (e.g., skewness or multiple peaks) as clearly as the histogram does.

$\\$





## 1.4 Quantitative data : histograms and outliers

Generate histograms for each of the following data sets. Use the `$` command to access the individual data sets. 

1. For each histogram, add the mean to the plot using `abline()`.

2. Add the median using `abline()`.

3. Do you see any potential outliers? Also calculate the five-number summary for each using R.

```{r}

set.seed(999)
data = data.frame(
  dat1 = -rchisq(1000, df = 1),
  dat2 = rchisq(1000, df = 1),
  dat3 = runif(1000),
  dat4 = rnorm(1000),
  dat5 = sample(c(rnorm(1000, mean = 2), rnorm(1000, mean = 10)), size = 1000)
)

```



**Answers**:

```{r}

mean1 <- mean(data$dat1 )
mean1

median1 <-  median(data$dat1 )
median1

hist1 <- hist(data$dat1,
  
     col = "pink",                           # bins color   
     main = "Histogram for data$dat1",      # title
     xlab = "mydata$dat values",              # x-axis label
     ylab = "Frequencies of data$dat1",      # y-axis label
     border = "darkgreen",                    # bins border's color 
     lwd = 2,                                 # border thickness
     cex.main = 1.5,                          # title size
     cex.lab = 1,                             # axis labels size
     cex.axis = 0.5,                          # tick labels size
     font.main = 2,                           # bold title
     font.lab = 3)                            # italic axis labels
     

abline(v= mean1, col="blue" )
abline(v= median1, col="red2" )

text(x = median1, y = 600,  cex= 0.5, labels = "median", 
     adj = 0.5, col = "red") # Add label slightly above the line

text(x = mean1, y = 650,  cex= 0.5, labels = "mean", 
     adj = 0.5, col = "blue") # Add label slightly above the line

```


$\\$

## 1.5 Percentiles

Compute the 25th, 50th, and 75th percentiles for the 5 datasets in the `data` data.frame. Which has the smallest median? Which has the largest?
You can use the function `quantile()`.

```{r}
# data 1 
percentiles <- quantile(data$dat1, c( 0.25, 0.5, 0.75))
percentiles

# data 2
percentiles <- quantile(data$dat2, c( 0.25, 0.5, 0.75))
percentiles

# data 3
percentiles <- quantile(data$dat3, c( 0.25, 0.5, 0.75))
percentiles

# data 4
percentiles <- quantile(data$dat4, c( 0.25, 0.5, 0.75))
percentiles

# data 5
percentiles <- quantile(data$dat5, c( 0.25, 0.5, 0.75))
percentiles



```


$\\$



# Part 2: Normal Distribution and +/- 2 Standard Deviations


## 2.1 Example: 

The normal distribution (also known as the "bell-curve") occurs very frequently in mathematics, statistics, and the natural and social sciences.

Which of the 5 data sets in the `data` data.frame appears to be normally distributed?


**Answers:**


```{r}
#par(mfrow=c(3,2))

  hist1 <- hist(data$dat1)
  hist2 <- hist(data$dat2)
  hist3 <- hist(data$dat3)
  hist4 <- hist(data$dat4)
  hist5 <- hist(data$dat5) 
  
```


It appears that `dat4` has a normal distribution:


$\\$




## 2.2 Calculate the endpoints of one standard deviation above and below the mean.

Using this data set, find the mean and standard deviation, then calculate the endpoints of **1 standard deviation above**, and **1 standard deviation below the mean**. What percentiles do these values correspond to?

**Note**: Approximately 68% of the data lie between 1 standard deviation above and 1 standard deviation below the mean.


**Answers**

* We will use `dat4`. 

```{r}

## Method 1:

mean_4 <- mean(data$dat4)
mean_4

sd_4 <- sd(data$dat4)
sd_4

endpoints_1sd_below_above_mean <- c(mean_4 - sd_4, mean_4 + sd_4)
endpoints_1sd_below_above_mean


## Method 2:

percentiles_middle_68percent <- quantile(data$dat4, c( 0.16, 0.84) )
percentiles_middle_68percent
```


$\\$


### 2.3 *z*-Scores 

Expressing a distance from the mean in standard deviations standardizes the
performance. To **standardize** a value, we subtract the mean and then divide
this difference by the standard deviation:

$$
z = \frac{y-\bar{y}}{s}
$$

The values are called **standardized values**, and are commonly denoted with
the letter $z$. Usually we just call them **z-scores**.


*z*-scores measure the distance of a value from the mean in standard
deviations. A *z*-score of 2 says that a data value is two standard deviations
above the mean. It doesn't matter whether the original variable was measured
in fathoms, dollars, or carats; those units don't apply to *z*-scores.

Data values below the mean have negative *z*-scores, so a *z*-score of
$-1.6$ means that the data value is 1.6 standard deviations below the mean.

Of course, regardless of the units, the farther a *z*-score is from 0, the
more unusual it is. Thus, a *z*-score of $-1.3$ is more extraordinary than a
*z*-score of $1.2$.


$\\$


### 2.4 Example of comparing *z*-Scores

A town's January high temperatures average $2^\circ$C with a standard
deviation of $6^\circ$, while in July the mean high temperature is
$24^\circ$C and the standard deviation is $5^\circ$.

**In which month is it more unusual to have a day with a high temperature of
$13^\circ$C? Explain.**

**Answers:**

```{r}
# January *z*-score
jan_z_score<- (13-2)/6
jan_z_score

# July *z*-score
jul_z_score<- (13-24)/5
jul_z_score

```

A $13^\circ$C day has a *z*-score of about $1.83$ in January but about $-2.2$ in July. Since the July *z*-score is farther from 0 in absolute value, a $13^\circ$C day is more unusual in July than in January.

$\\$



