---
title: "Class 20 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---





# Part 1: Inference using the normal distribution

We can use the normal distribution to perform inference about population parameters
based on sample statistics.



$\\$





## Example: Do goalies guess the direction of a penalty shot less than 50% of the time? 


From 1982 to 1994 there were 128 penalty shots in the World Cup.  
Goal keepers correctly guessed the direction 41% of the time. Please run a 
hypothesis test to assess whether goalies guessed the correct direction of the
kick less than 50% of the time. 




### Step 1: State the null and alternative hypotheses







$\\$




### Step 2: Calculate the observed statistic (and check conditions are met)

We can calculate the z-statistic using the formula: 

$$z = \frac{\hat{p} - \pi_0}{SE} $$

where:

$$SE = \sqrt{\frac{\pi_0 \cdot (1 - \pi_0)}{n}}$$



```{r}

library(SDS1000)



# step 2: calculate the z-statistic








# check conditions are met






```




$\\$




### Step 3: Visualize the null distribution

Our null distribution is the standard normal distribution.We can visualize this
null distribution using `dnorm()` function.


```{r}










```



$\\$



### Step 4: Calculate the p-value using the standard normal distribution 

Since our conditions were met, we cna use the standard normal distribution as 
our null distribution. Thus we can get p-values using the `pnorm()` function. 


```{r}








```



$\\$



### Step 5:  Make a decision 










$\\$




# Part 2: Parametric confidence intervals for proportions


A survey asked 2625 people whether they agreed with the statement 
"There is only one true love for each person". 1812 of the respondents disagreed.

Please compute a 90% confidence interval for the proportion who disagreed.

Recall that the the formula for an estimate of the standard error for a proportion
is:

$$\hat{SE} = \sqrt{ \frac{\hat{p} \cdot (1 - \hat{p}) }{n}}$$
And a confidence interval is given by: 

$$\hat{p} \pm z^* \cdot \hat{SE}$$

```{r}

# The data






# Checking conditions 





# calculate the SE estimate





# create the confidence interval





```




$\\$





# Part 3: Practice problem - sinister lawyers


10% of American population is left-handed. 

A study found that out of a random sample of 105 lawyers, 16 were left-handed

Test whether the proportion of left-handed lawyers is greater than the proportion 
found in the American population. 



$\\$



### Step 1: State the null and alternative hypotheses







$\\$





### Step 2: Calculate the observed statistic (and check conditions are met)

We can calculate the z-statistic using the formula: 

$$z = \frac{\hat{p} - \pi_0}{SE} $$

where:

$$SE = \sqrt{\frac{\pi_0 \cdot (1 - \pi_0)}{n}}$$



```{r}

# step 2: calculate the z-statistic









# check conditions are met







```




$\\$




### Step 3: Visualize the null distribution

Our null distribution is the standard normal distribution.We can visualize this
null distribution using `dnorm()` function.


```{r}









```



$\\$



### Step 4: Calculate the p-value using the standard normal distribution 

Since our conditions were met, we can use the standard normal distribution as 
our null distribution. Thus we can get p-values using the `pnorm()` function. 


```{r}








```



$\\$



### Step 5:  Make a decision 








$\\$









## Practice: confidence interval for sinister lawyers


Please compute a 95% confidence interval for the proportion of lawyers who are left-handed.

Recall that the the formula for an estimate of the standard error for a proportion
is:

$$\hat{SE} = \sqrt{ \frac{\hat{p} \cdot (1 - \hat{p}) }{n}}$$
And a confidence interval is given by: 

$$\hat{p} \pm z^* \cdot \hat{SE}$$

```{r}

# The data






# Checking conditions 






# calculate the SE estimate







# create the confidence interval






```


Q: What does the ocnfidence interval tell you? 















