---
title: "Class 20 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---





# Part 1: Inference using the normal distribution

We can use the normal distribution to perform inference about population parameters
based on sample statistics.



$\\$




## Example: Hypothesis test for a single proportion with a known SE


Do goalies guess the direction of a penalty shot less than 50% of the time? 


From 1982 to 1994 there were 128 penalty shots in the World Cup.  
Goal keepers correctly guessed the direction 41% of the time With SE* = 0.043                             



### Step 1: 





$\\$




### Step 2-5: 

2. Calculate the z-statistic using the formula: $$z = \frac{\hat{p} - \pi_0}{SE^*} $$

3. Visualize the null distribution using `dnorm()` function.

4. Calculate the p-value using the standard normal distribution (`pnorm()` function)

5. Make a decision based on the p-value and significance level (e.g., $\alpha$ = 0.05)



```{r}


# step 2: calculate the z-statistic




# step 3: visualize the null distribution







# step 4: use the pnorm function to get the p-value




```


5. Make a decision!








$\\$



Can you also create a 99% confidence interval for the proportion of correct
guesses by goalies during penalty shots using the known SE?

```{r ci_proportion}

# Given data






# Calculate the margin of error




# Calculate the confidence interval






```



$\\$




# Part 2: Sampling distribution for proportions


In the United States, 65.1% of all houses are owned. Suppose that we sampled n = 50
houses at random from the population of all houses in the United States.

Can you visualize what the sampling distribution of the sample proportion would look like
if we took many samples of size n = 50?

What if we sampled n = 200 houses instead?

```{r sampling_dist_proportion, fig.width=7, fig.height=4}





# Calculate the standard error



# Create a sequence of x values



# Calculate the density values for the sampling distribution



# Plot the sampling distribution







# Repeat for n = 200










```





$\\$




# Part 3: Parametric confidence intervals a proportion

A survey asked 2625 people whether they agreed with the statement “There is only
one true love for each person”. 1812 of the respondents disagreed. Compute a 90%
confidence interval for the proportion who disagreed.

Hint: this formula could help: CI = $\hat{p} \pm \cdot \sqrt{\frac{\hat{p}(1 - \hat{p})}{n}}$


```{r ci_proportion_parametric}

# Given data
n <- 2625
x <- 1812




# calculate the critical value




# Calculate the standard error



# Calculate the margin of error



# Calculate the confidence interval





```



Bonus: Create a confidence interval using the bootstrap for the same data and compare the results!





$\\$





# Part 4: Parametric hypothesis test for a proportion

A telephone survey of 1000 randomly selected US adults found that 31% of them
say they believe in ghosts. Does this provided evidence that more than 1 in 4 US
adults believe in ghosts?

Please run a hypothesis test at the $\alpha$ = 0.05 significance level to assess
whether more than 25% of US adults believe in ghosts.



$\\$


Step 1: state the null and alternative hypotheses

In words:

Null hypothesis: 

Alternative hypothesis: 



In symbols:






```{r hypothesis_test_proportion}


# Step 2: Calculate the z-statistic






# Step 3: Visualize the null distribution









# Step 4: Calculate the p-value





```



Step 5: Make a decision








$\\$













