---
title: "Class 20 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---





# Part 1: Inference using the normal distribution

We can use the normal distribution to perform inference about population parameters
based on sample statistics.



$\\$





## Example: Do goalies guess the direction of a penalty shot less than 50% of the time? 


From 1982 to 1994 there were 128 penalty shots in the World Cup.  
Goal keepers correctly guessed the direction 41% of the time. Please run a 
hypothesis test to assess whether goalies guessed the correct direction of the
kick less than 50% of the time. 




### Step 1: State the null and alternative hypotheses


$H_0: \pi$ = .5
$H_A: \pi$ < .5



$\\$




### Step 2: Calculate the observed statistic (and check conditions are met)

We can calculate the z-statistic using the formula: 

$$z = \frac{\hat{p} - \pi_0}{SE} $$

where:

$$SE = \sqrt{\frac{\pi_0 \cdot (1 - \pi_0)}{n}}$$



```{r}

library(SDS1000)



# step 2: calculate the z-statistic

SE <- sqrt( (.5 * (1-.5)) / 128)

z_stat <- (.41 - .5)/SE




# check conditions are met

128 * .5

128 * (1 - .5)


```




$\\$




### Step 3: Visualize the null distribution

Our null distribution is the standard normal distribution.We can visualize this
null distribution using `dnorm()` function.


```{r}


x_values_null <- seq(-4, 4, length.out = 1000)
density_values_null <- dnorm(x_values_null, mean = 0, sd = 1)


plot(x_values_null, density_values_null, type = "l",
     main = "Null Distribution of Z-statistic",
     xlab = "Z-score", 
     ylab = "Density", 
     col = "blue")


abline(v = z_stat, col = "red", lwd = 2)


```



$\\$



### Step 4: Calculate the p-value using the standard normal distribution 

Since our conditions were met, we cna use the standard normal distribution as 
our null distribution. Thus we can get p-values using the `pnorm()` function. 


```{r}


p_value <- pnorm(z_stat, 0, 1, lower.tail = TRUE)

p_value


```



$\\$



### Step 5:  Make a decision 


Since our p-value is less than 0.05 we would reject the null hypothesis and
conclude that goalies guess the direction of a penalty shot less than 50% of the
time.





$\\$




# Part 2: Parametric confidence intervals for proportions


A survey asked 2625 people whether they agreed with the statement 
"There is only one true love for each person". 1812 of the respondents disagreed.

Please compute a 90% confidence interval for the proportion who disagreed.

Recall that the the formula for an estimate of the standard error for a proportion
is:

$$\hat{SE} = \sqrt{ \frac{\hat{p} \cdot (1 - \hat{p}) }{n}}$$
And a confidence interval is given by: 

$$\hat{p} \pm z^* \cdot \hat{SE}$$

```{r}

# The data
n <- 2625
p_hat <- 1812/n

p_hat


# Checking conditions 
n * p_hat
n * (1 - p_hat)


# calculate the SE estimate

SE <- sqrt(  (p_hat * (1 - p_hat)) /n  )
SE 


# create the confidence interval

CI_lower <- p_hat - cnorm(.9) * SE
CI_upper <- p_hat + cnorm(.9) * SE

c(CI_lower, CI_upper )


```




$\\$





# Part 3: Practice problem - sinister lawyers


10% of American population is left-handed. 

A study found that out of a random sample of 105 lawyers, 16 were left-handed

Test whether the proportion of left-handed lawyers is greater than the proportion 
found in the American population. 



$\\$



### Step 1: State the null and alternative hypotheses

$H_0: \pi = 0.10$

$H_A: \pi > 0.10$


$\\$





### Step 2: Calculate the observed statistic (and check conditions are met)

We can calculate the z-statistic using the formula: 

$$z = \frac{\hat{p} - \pi_0}{SE} $$

where:

$$SE = \sqrt{\frac{\pi_0 \cdot (1 - \pi_0)}{n}}$$



```{r}

# step 2: calculate the z-statistic

pi_0 <- .1
n <- 105
p_hat <- 16/105
p_hat


SE <- sqrt( (pi_0  * (1 - pi_0 )) / n) 
SE


z_stat <- (p_hat - .1)/SE
z_stat



# check conditions are met

105 * .1
105 * (1 - .1)


```




$\\$




### Step 3: Visualize the null distribution

Our null distribution is the standard normal distribution.We can visualize this
null distribution using `dnorm()` function.


```{r}

x_values_null <- seq(-4, 4, length.out = 1000)
density_values_null <- dnorm(x_values_null, mean = 0, sd = 1)


plot(x_values_null, density_values_null, type = "l",
     main = "Null Distribution of Z-statistic",
     xlab = "Z-score", 
     ylab = "Density", 
     col = "blue")


abline(v = z_stat, col = "red", lwd = 2)


```



$\\$



### Step 4: Calculate the p-value using the standard normal distribution 

Since our conditions were met, we can use the standard normal distribution as 
our null distribution. Thus we can get p-values using the `pnorm()` function. 


```{r}


p_value <- pnorm(z_stat, 0, 1, lower.tail = FALSE)

p_value


```



$\\$



### Step 5:  Make a decision 


Since our p-value is slightly more than 0.05 we fail to reject the null hypothesis.
Thus we can't conclude that lawyers are more sinister. If we wanted to show this,
we would have to redo our whole experiment with a larger sample size and then
perhaps we would be able to reject the null hypothesis. 



$\\$



## Practice: confidence interval for sinister lawyers


Please compute a 95% confidence interval for the proportion of lawyers who are left-handed.

Recall that the the formula for an estimate of the standard error for a proportion
is:

$$\hat{SE} = \sqrt{ \frac{\hat{p} \cdot (1 - \hat{p}) }{n}}$$
And a confidence interval is given by: 

$$\hat{p} \pm z^* \cdot \hat{SE}$$

```{r}

# The data
n <- 105
p_hat <- 16/n

p_hat


# Checking conditions 
n * p_hat
n * (1 - p_hat)


# calculate the SE estimate

SE <- sqrt(  (p_hat * (1 - p_hat)) /n  )
SE 


# create the confidence interval

CI_lower <- p_hat - cnorm(.95) * SE
CI_upper <- p_hat + cnorm(.95) * SE

c(CI_lower, CI_upper )


```


Q: What does the ocnfidence interval tell you? 


A: Somewhere between 8.3% and 22.1% of lawyers are left-handed (so it is feasible
that 10% of lawyers are left-handed)











