---
title: "Practice Session 4 "
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---

# Part 1: Sampling Distributions

## 1.1 Introduction: 

Recall that a "sample" is a subset of a population. We can use samples to calculate "statistics", such as the sample mean, median, and standard deviation. If we were to **repeatedly** sample from the same population, and calculate the **mean** of each sample, we would obtain what is known as a **Sampling Distribution** of the **Sample Mean**. If we were to instead calculate a **Proportion** from each sample, we would obtain the Sampling Distribution of the **Sample Proportion**. 

In general, Sampling Distributions are the distributions of a **statistic**. This is in contrast to the Population, which is the distribution of all individuals or sampling units.


## 1.2 Sampling Distribution Terminology

Match each term with its definition.\ 

**Terms**

* (a) s
* (b) $\sigma$
* (c) Standard Error (SE)
* (d) x
* (e) $\bar{x}$

**Definitions:**

* Standard deviation of a population
* Individual unit of the population
* Standard deviation of a sample
* Sample mean 
* Standard deviation of a sampling distribution


$\\$

## 1.3  Generating Sampling Distributions

Let's generate a sampling distribution using polling data from the president's approval rating. Make sure to load the `library(SDS1000)` to have access to the functions for this exercise.\ 

1.) Use the `get_approval_sample()` function to obtain a sample of `n = 1000` individual approval ratings. Use the `get_proportions()` function to obtain the proportion of voters in this sample that approve of the president.


```{r}
# Load the SDS1000 package
library(SDS1000)

# Setting the seed so that we all get the same "random sample"
set.seed(1000)  

# your code here #
```



2.) Next, use the `do_it` function from class to generate 10000 samples, and repeat the above steps to obtain the proportion for each sample. Assign this vector of proportions to the variable `polling_distribution`, and create a histogram of it. Make sure to have an appropriate title.


**Challenge: What do values on the x-axis of this histogram correspond to? How does this differ from histograms that we've previously encountered?**

```{r}

# your code here #
```



3.) Finally, calculate the mean and standard deviation of your collection of sample proportions. What is another name for the standard deviation?

```{r}

# your code here #
```


$\\$


## 1.4  Properties of Sampling Distribution

Answer the following True/False questions relating to sampling distributions.

* (a)  True/False: As the sample size $n$ decreases, the standard error of a sampling distribution also decreases.

* (b)  True/False: The mean of a sampling distribution is always **greater than** the mean of population.

* (c) True/False: The mean of a sampling distribution is always is always **less than** the mean of the population.

* (d) True/False: Assuming that you have an unbiased sampling procedure with a sufficient sample size, the sampling distribution of the sample mean is approximately normally distributed.

* (e) True/False: It is generally better to have a larger standard error than a smaller one.


$\\$

# Part 2:  Introduction to Confidence Interval

Match each term with its definition: 

**Terms:**

* Point Estimate
* Interval estimate
* Confidence Interval
* Confidence Level


**Definitions:**

* (a) An interval that is generated with a specific method that guarantees that a certain percentage of these intervals will contain the population parameter.

* (b) The percentage of confidence intervals that will contain the population parameter of interest. For example, a 95% confidence level means that 95% of confidence intervals generated will contain the population parameter of interest.

* (c) A sample statistic that we use to estimate a population parameter. The sample mean is an example as we use it to estimate the population mean.

* (d) An interval that contains a range of plausible values for a population parameter. The point estimate sits at the center of this interval.


$\\$


# Part 3 :  More Practices on Sampling Distributions 

## Practice:  Baseball Salaries

For this section, we will use the `BaseballSalaries2015` data set from the `Lock5Data` library.


1.) Using the `sample()` function, obtain a sample of $n = 100$ baseball player salaries. See the `help` page for `sample()` for details.


```{r}
library(Lock5Data)
set.seed(1000)

# your code here #
```

2.) Now, using the `do_it()` function, repeat this process $n = 10,000$ times, and calculate the mean each time.

```{r}

# your code here #
```


3.) Using your original sample of 100 salaries from part 1.), calculate a point estimate of the mean salary. 

```{r}

# your code here #
```

4.) Calculate the standard deviation of your sampling distribution from part 2.) to estimate the standard error.

```{r}

# your code here #
```

