---
title: "Class 14 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---



## Part 1: Hypothesis tests for two means


A randomized controlled trial by Lyle et al (1987) investigated whether calcium
lowered blood pressure.

 - A treatment group of 10 men received a calcium supplement for 12 weeks.  
 - A control group of 11 men received a placebo during the same period.  

The blood pressure of these men was taken before and after the 12 weeks of the study.

Let's run a hypothesis test to see if there was a larger decrease in blood
pressure for participants that received calcium compared to those that received
a placebo.




$\\$



### Step 1: State the null and alternative hypotheses

Words:

* Null hypothesis: 
* Alternative hypothesis: 
 
 
Symbols:


 




$\\$






### Step 2: Calculate the observed statistic


```{r calcium_obs_stat}


# the data from the calcium study
treat <- c(7, -4, 18, 17, -3, -5,  1, 10, 11, -2)
control <- c(-1,  12,  -1,  -3,   3,  -5,   5,   2, -11,  -1,  -3)



# observed statistic




```






$\\$






### Step 3: create the null distribution


```{r calcium_null_distribution}

library(SDS1000)


#  Combine data from both groups


# Create the null distribution


       # shuffle data



       # create fake treatment and control groups



       # calculate the shuffled statistic of interest







# plot the null distribution




# add a red vertical line at the observed statistic




```







$\\$







### Step 4: calculate the p-value


```{r calcium_p_value}




```





$\\$






### 5. Step 5: Make decision








$\\$





## Part 2: Hypothesis tests for two means (Husbands and Wives data)


Let's look at another example to test if the mean average of husbands tends to be
greater than the mean average of wives.

To address this qeuestion, we will use a data from a sample of 100 marriage
licences from St. Lawrence County, NY that gives the ages of husbands and wives
for newly married couples.

But before we look at the data, let's start with step 1 by stating the null and
alternative hypotheses, and then go from there...



$\\$




### Step 1: State the null and alternative hypotheses


Words:

* Null hypothesis:  
* Alternative hypothesis: 


Symbols:  






$\\$



### Step 2a: Visualize the data

Before we calculate our observed statistic let's visualize the data. 

Please use an appropriate method we discussed in class to visualize the data.


```{r husbands_and_wives_viz}

# load the data
library(Lock5Data)

husbands <- MarriageAges$Husband
wives <- MarriageAges$Wife



# visualize the data using side-by-side box plots





```


**Question**: From looking at the visualization you created, does it appear that 
husbands tend to be older than wives?


**Answer**: 






$\\$





### Step 2b: Calculate the observed statistic

Now let's do actual step 2 and calculate the observed statistic.


```{r husbands_and_wives_obs_stat}

# calculate the observed statistic




```




$\\$




### Step 3: Create the null distribution


```{r husbands_and_wives_null_distribution}

library(SDS1000)

# combine the data into a single vector


# create the null distribution










# plot the null distribution






# add a red vertical line at the observed statistic





```




$\\$




### Step 4: Calculate the p-value


```{r husbands_and_wives_p_value}





```




$\\$



### Step 5: Make decision






$\\$





## Part 3: Hypothesis tests for more than two means


A group of Hope College students wanted to see if there was an association
between a student’s major and the time it takes to complete a small Sudoku-like
puzzle


They grouped majors into four categories

* Applied science (as)
* Natural science (ns)
* Social science  (ss)
* Arts/humanities  (ah)

Let's run a hypothesis test to see if there is an association between a student's
major and the time it takes to complete the puzzle.



$\\$



### Step 1: State the null and alternative hypotheses



Words:

* Null hypothesis: 

* Alternative hypothesis: 



Symbols:  






$\\$




### Step 2a: Visualize the data

```{r sudoku_viz}

# load the data
sudoku_data <- read.table("MajorPuzzle.txt", header = TRUE)

completion_times <- sudoku_data$time
majors <- sudoku_data$major



# visualize the data using side-by-side box plots






```



**Question**: From looking at the visualization you created, does it appear that
the mean time to complete the puzzle is different for at least one of the four
groups?


**Answer**: 






$\\$




### Step 2b: Calculate the observed statistic


```{r sudoku_obs_stat}

# calculate the observed statistic





```




$\\$



### Step 3: Create the null distribution


```{r sudoku_null_distribution}

library(SDS1000)


# Create the null distribution








# plot the null distribution




# add a red vertical line at the observed statistic




```




$\\$




### Step 4: Calculate the p-value


```{r sudoku_p_value}




```




$\\$




### Step 5: Make decision








