---
title: "Class 14 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---



## Part 1: Hypothesis tests for two means


A randomized controlled trial by Lyle et al (1987) investigated whether calcium
lowered blood pressure.

 - A treatment group of 10 men received a calcium supplement for 12 weeks.  
 - A control group of 11 men received a placebo during the same period.  

The blood pressure of these men was taken before and after the 12 weeks of the study.

Let's run a hypothesis test to see if there was a larger decrease in blood
pressure for participants that received calcium compared to those that received
a placebo.




$\\$



### Step 1: State the null and alternative hypotheses

Words:

* Null hypothesis: The mean decrease in blood pressure is the same for both groups  
* Alternative hypothesis: The mean decrease in blood pressure is greater for the treatment group  
 
Symbols:

* $H_0: \mu_{treat} = \mu_{control}$  
* $H_A: \mu_{treat} > \mu_{control}$  
 




$\\$






### Step 2: Calculate the observed statistic


```{r calcium_obs_stat}


# the data from the calcium study
treat <- c(7, -4, 18, 17, -3, -5,  1, 10, 11, -2)
control <- c(-1,  12,  -1,  -3,   3,  -5,   5,   2, -11,  -1,  -3)



# observed statistic
obs_stat <- mean(treat) - mean(control)

obs_stat


```






$\\$






### Step 3: create the null distribution


```{r calcium_null_distribution}

library(SDS1000)


#  Combine data from both groups
combined_data <- c(treat, control)


null_dist <-  do_it(10000) * {

       # shuffle data
        shuff_data <- shuffle(combined_data)

        # create fake treatment and control groups
        shuff_treat   <-  shuff_data[1:10]
        shuff_control  <-  shuff_data[11:21]

        # calculate the shuffled statistic of interest
        mean(shuff_treat) - mean(shuff_control)

}


# plot the null distribution
hist(null_dist, breaks = 200,
     main = "Null distribution",
     xlab = "Difference in mean decrease in blood pressure")


# add a red vertical line at the observed statistic
abline(v = obs_stat, col = "red")


```







$\\$







### Step 4: calculate the p-value


```{r calcium_p_value}

pnull(obs_stat, null_dist, lower.tail = FALSE)

```





$\\$






### 5. Step 5: Make decision


Since the p-value is (just barely) greater than the typical cutoff of 0.05. So,
if we were to use the strict rule of only rejecting the null hypothesis if the
significance level is less than 0.05 we would fail to reject the null hypothesis
here.



$\\$





## Part 2: Hypothesis tests for two means (Husbands and Wives data)


Let's look at another example to test if the mean average of husbands tends to be
greater than the mean average of wives.

To address this qeuestion, we will use a data from a sample of 100 marriage
licences from St. Lawrence County, NY that gives the ages of husbands and wives
for newly married couples.

But before we look at the data, let's start with step 1 by stating the null and
alternative hypotheses, and then go from there...



$\\$




### Step 1: State the null and alternative hypotheses


Words:

* Null hypothesis: The mean age of husbands is the same as the mean age of wives. 
* Alternative hypothesis: The mean age of husbands is greater than the mean age of wives.  


Symbols:  

* $H_0: \mu_{husbands} = \mu_{wives}$  
* $H_A: \mu_{husbands} > \mu_{wives}$  



$\\$



### Step 2a: Visualize the data

Before we calculate our observed statistic let's visualize the data. 

Please use an appropriate method we discussed in class to visualize the data.


```{r husbands_and_wives_viz}

# load the data
library(Lock5Data)

husbands <- MarriageAges$Husband
wives <- MarriageAges$Wife

# visualize the data using side-by-side box plots
boxplot(husbands, wives,
        names = c("Husbands", "Wives"),
        main = "Marriage Ages",
        ylab = "Age (years)",
        col = c("lightblue", "lightpink"))

```


**Question**: From looking at the visualization you created, does it appear that 
husbands tend to be older than wives?


**Answer**: From my visualization it is hard to tell. Of course the box plot is 
showing that the median age of husbands is greater than the median age of wives,
but there is also a lot of overlap in the ages of husbands and wives. So, it is
hard to tell just by looking at the box plot if the mean age of husbands is
actually greater than the mean age of wives.




$\\$





### Step 2b: Calculate the observed statistic

Now let's do actual step 2 and calculate the observed statistic.


```{r husbands_and_wives_obs_stat}

# calculate the observed statistic
obs_stat <- mean(husbands) - mean(wives)

obs_stat

```




$\\$




### Step 3: Create the null distribution


```{r husbands_and_wives_null_distribution}

library(SDS1000)

# combine the data into a single vector
combined <- c(husbands, wives)

null_dist <- do_it(10000) * {

  # shuffle the combined data
  shuffled <- sample(combined)

  # split the shuffled data into two groups
  sim_husbands <- shuffled[1:length(husbands)]
  sim_wives <- shuffled[(length(husbands) + 1):length(combined)]

  # calculate the simulated statistic
  mean(sim_husbands) - mean(sim_wives)

}


# plot the null distribution
hist(null_dist, breaks = 200,
     main = "Null distribution",
     xlab = "Difference in mean ages")

# add a red vertical line at the observed statistic
abline(v = obs_stat, col = "red")


```




$\\$




### Step 4: Calculate the p-value


```{r husbands_and_wives_p_value}

p_value <- pnull(obs_stat, null_dist, lower.tail = FALSE)

p_value

```




$\\$



### Step 5: Make decision

Since the p-value is slightly less than the typical cutoff of 0.05, we can
(tentatively reject the null hypothesis and conclude that the mean age of
husbands is greater than the mean age of wives.




$\\$





## Part 3: Hypothesis tests for more than two means


A group of Hope College students wanted to see if there was an association
between a student’s major and the time it takes to complete a small Sudoku-like
puzzle


They grouped majors into four categories

* Applied science (as)
* Natural science (ns)
* Social science  (ss)
* Arts/humanities  (ah)

Let's run a hypothesis test to see if there is an association between a student's
major and the time it takes to complete the puzzle.



$\\$



### Step 1: State the null and alternative hypotheses



Words:

* Null hypothesis: The mean time to complete the puzzle is the same for all four groups  

* Alternative hypothesis: The mean time to complete the puzzle is not the same 
for all four groups  


Symbols:  

* $H_0: \mu_{as} = \mu_{ns} = \mu_{ss} = \mu_{ah}$   
* $H_A$: At least one of the means is different  




$\\$




### Step 2a: Visualize the data

```{r sudoku_viz}

# load the data
sudoku_data <- read.table("MajorPuzzle.txt", header = TRUE)

completion_times <- sudoku_data$time
majors <- sudoku_data$major

# visualize the data using side-by-side box plots
boxplot(completion_times ~ majors, 
        main = "Time to complete puzzle by major",
        ylab = "Time (seconds)",
        col = c("lightblue", "lightgreen", "lightpink", "lightyellow"))

```



**Question**: From looking at the visualization you created, does it appear that
the mean time to complete the puzzle is different for at least one of the four
groups?


**Answer**: From my visualization it appears that the mean time to complete the
puzzle is different for at least one of the four groups. The box plot is showing
that the median time for the natural science group is lower than the median times
for the other three groups. However, there is also a lot of overlap in the times
for the four groups. So, it is hard to tell just by looking at the box plot if
the mean time to complete the puzzle is actually different for at least one of
the four groups.




$\\$




### Step 2b: Calculate the observed statistic


```{r sudoku_obs_stat}

# calculate the observed statistic

obs_stat <- get_MAD_stat(completion_times, majors)

obs_stat

```



$\\$



### Step 3: Create the null distribution


```{r sudoku_null_distribution}

library(SDS1000)


# Create the null distribution

null_dist <- do_it(10000) * {

  # shuffle the completion times
  shuffled_majors <- shuffle(majors)

  # calculate the simulated statistic
  get_MAD_stat(completion_times, shuffled_majors)

}

# plot the null distribution
hist(null_dist, breaks = 200,
     main = "Null distribution",
     xlab = "Mean absolute deviation of group means")


# add a red vertical line at the observed statistic
abline(v = obs_stat, col = "red")

```




$\\$




### Step 4: Calculate the p-value


```{r sudoku_p_value}

p_value <- pnull(obs_stat, null_dist, lower.tail = FALSE)

p_value

```




$\\$




### Step 5: Make decision

Since the p-value is greater than the typical cutoff of 0.05, we fail to reject
the null hypothesis. Thus we can't conclude that the mean time to complete the
puzzle is different for the different majors.




