---
title: "Class 14 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---




## Part 1: Lie detector hypothesis test

A study by Hollien, Harnsberger, Martin and Hollien (2010) tried to assess the 
accuracy of lie detection software. 

A sample of 48 participants were gather and attached to a lie detection device. 
They were asked to read deceptive (lying) material out loud. 

The lie detector correctly reported that 31 out of the 48 participants were lying. 

Let's run a hypothesis test to see if the results provide evidence that lie 
detectors are more than 60% accurate? 





$\\$





### Step 1: State the null and alternative hypotheses 


Words:  



 
Symbols:  


  



$\\$




### Step 2: Calculate the observed statistic



```{r lie_detector_stat}


# step 2: compute observed statistic




```




$\\$





### Step 3: create the null distribution



```{r lie_detector_null_distribution}


library(SDS1000)



# create the null distribution






# plot the null distribution





# add a red vertical line at the observed statistic





```






$\\$






### Step 4: calculate the p-value


```{r lie_detector_p_value}




```





$\\$





### 5. Step 5: Make decision








$\\$




**Question**: What would have happened if we had tested whether lie detector 
tests get it correct more than 50% of the time? 

Try it!





$\\$




## Part 2: Two-sided hypothesis tests

Let's say we want to test whether the lie detector is different from 60% accurate.


$\\$


### Step 1: State the null and alternative hypotheses

Words:  



 
Symbols:  



   
   

$\\$   
   




### Step 2: Calculate the observed statistic


```{r lie_detector_stat_2sided}




```





$\\$






### Step 3: create the null distribution


```{r lie_detector_null_distribution_2sided}

library(SDS1000)


# create the null distribution





# plot the null distribution




# add a red vertical line at the observed statistic




```




$\\$





### Step 4: calculate the p-value


```{r lie_detector_p_value_2sided}

# create the margin between the observed statistic and the null hypothesis parameter value




# calculate the left and right tail p-values





# calculate the two-sided p-value





```




$\\$




### 5. Step 5: Make decision







$\\$






## Part 3: Hypothesis tests for two means


A randomized controlled trial by Lyle et al (1987) investigated whether calcium
lowered blood pressure.

 - A treatment group of 10 men received a calcium supplement for 12 weeks.  
 - A control group of 11 men received a placebo during the same period.  

The blood pressure of these men was taken before and after the 12 weeks of the study.

Let's run a hypothesis test to see if there was a larger decrease in blood
pressure for participants that received calcium compared to those that received
a placebo.




$\\$





### Step 1: State the null and alternative hypotheses

Words:




Symbols:





$\\$






### Step 2: Calculate the observed statistic


```{r calcium_obs_stat}


# the data from the calcium study
treat <- c(7, -4, 18, 17, -3, -5,  1, 10, 11, -2)
control <- c(-1,  12,  -1,  -3,   3,  -5,   5,   2, -11,  -1,  -3)



# observed statistic






```






$\\$






### Step 3: create the null distribution


```{r calcium_null_distribution}


#  combine data from both groups


# create the null distribution












# plot the null distribution



# add a red vertical line at the observed statistic




```







$\\$







### Step 4: calculate the p-value


```{r calcium_p_value}



```





$\\$






### 5. Step 5: Make decision








     







