---
title: "Class 5 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---







# Part 1: Quantiles, five number summary and boxplots



$\\$




## Part 1.2: Quantiles 

The $P^{th}$ percentile is the value of a quantitative variable which is greater 
than P percent of the data. 

We can calculate the percentiles of a sample of data using the `quantile()` function.

The code loads information about the winner for the best actor and best atress Oscars
award winners from 1929 to 2019, into vectors called `best_actor_age` and `best_actress_age`.

Let's calculate the age at the 20th, 50th, and 80th percentiles for the best actor award winners.


```{r}


# Load the data 
load("oscars.rda")

# create the vectors that have the ages of the best actor and best actress award winners
best_actor_age <- subset(oscars, award == "Best actor")$age
best_actress_age <- subset(oscars, award == "Best actress")$age


# Calculate the 20th, 50th, and 80th percentiles for best actor ages
quantile(best_actor_age, c(.2, .5, .8))


# As we discussed last class ptail() returns the percentile rank, which is the inverse of the quantile function
library(SDS1000)

ptail(37, best_actor_age)  # 37 is the smallest value x in our dataset, such that 20% of the data is less x


```

We interpret these results are:

1. 20% of the best actor award winners were age 37 or younger

2. 50% of the best actor award winners were age 42 or younger

3. 80% of the best actor award winners were age 51 or younger




$\\$





# Part 1.2: Five number summary and boxplots

A five number summary consists of the minimum, first quartile (Q1), median (Q2), 
third quartile (Q3), and maximum of a quantitative variable.

We can calculate the five number summary using the `fivenum()` function.

```{r}

fivenum(best_actor_age)

```



We can also visualize the five number summary using a boxplot. We can create boxplots using
the `boxplot()` function. 

Let's create side-by-side boxplots comparing the ages of best actor and best actress award winners.


```{r}

boxplot(best_actor_age, best_actress_age,
        names = c("Best actor", "Best actress"),
        ylab = "Age",
        main = "Ages of Best Actor and Best Actress Award Winners")

```


If you look at the boxplot of the ages of best actor you will notice that there is an outlier in the data. Is this outlier due to an error in the data or is it a valid data point?




$\\$





# Part 2: Scatter plots and correlation

Let's use the data collected by [Fraumeni (1968)](https://academic.oup.com/jnci/article/41/5/1205/903436) 
to examine the relationship between the number of cigarette's sold per capita
and the rates of different types of cancer. 

The data is loaded below into a data frame named `smoking`. Cancer rates are 
reported as the number of deaths per 100,000 citizens. 


```{r smoking_data}

load("states_smoking.rda")

```



$\\$




### 2.1 Scatter plots

Start by creating a scatter plot between the number of cigarette's sold per 
capita (`CIG`) and the number of lung cancer deaths (`LUNG`). 


```{r smoking_scatter_plot}

plot(smoking$CIG, smoking$LUNG,
     xlab = "Cigarettes sold per capita (1960)",
     ylab = "Lung cancer deaths per 100,000",
     main = "Scatter plot of cigarettes sold vs lung cancer deaths")


```


$\\$


### 2.2 Scatter plots

Now let's examine the correlation between the number of cigarette's sold per 
capita (`CIG`) and the number of lug cancer deaths (`LUNG`). 

```{r smoking_correlation}


cor(smoking$CIG, smoking$LUNG)

```


$\\$



### 2.3 Additional explorations

Do you think there is a correlation between the number of cigarette's sold per
capita (`CIG`) and the rate of bladder cancer (`BLAD`) and/or luekemia (`LEUK`)?

Try the following to explore these questions:

1. Create scatter plots of these relationships cigarette's sold per and these 
types of cancer. Based on the scatter plot guess what the correlation values are.

2. Calculate the correlation values between cigarette's sold per capita and these
types of cancer. Do the correlation values match your expectations based on the
scatter plots?

3. Based on these results, do you think there is a real relationship between the
number of cigarette's sold per capita and these types of cancer?



```{r smoking_additional_explorations}

# Scatter plot of CIG vs BLAD
plot(smoking$CIG, smoking$BLAD,
     xlab = "Cigarettes sold per capita (1960)",
     ylab = "Bladder cancer deaths per 100,000",
     main = "Scatter plot of cigarettes sold vs bladder cancer deaths")


# Scatter plot of CIG vs LEUK
plot(smoking$CIG, smoking$LEUK,
     xlab = "Cigarettes sold per capita (1960)",
     ylab = "Leukemia deaths per 100,000",
     main = "Scatter plot of cigarettes sold vs leukemia deaths")


# Correlation between CIG and BLAD
cor(smoking$CIG, smoking$BLAD)


# Correlation between CIG and LEUK
cor(smoking$CIG, smoking$LEUK)


```




$\\$


