---
title: "Practice Session 7 Answers"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---

# Intro

The focus of this practice session will be to perform hypothesis tests for the difference of two or more means. We will look at various test statistics that can be used. We will also conduct hypothesis testing for correlations.


## Question 1: Exercise Hours Hypothesis Test for the Difference of Two Means


A study is interested to check if the mean exercise hours for female students is less than the mean exercise hours for male students. Use data `ExerciseHours` and the two variables `Exercise` and `Sex`.

```{r}
library(SDS1000)
library(Lock5Data)
data(ExerciseHours)

```


1.a  Write the `null hypothesis`  and `alternative hypothesis` 

$\ H_0:\mu_{f}= \mu_{m}$   vs    $\ H_a: \mu_{f} < \mu_{m}$

&nbsp;

1.b Create a boxplot to describe hours of exercise for `female` versus `male`.


```{r}

boxplot(Exercise ~ Sex , data = ExerciseHours)
```

1.c Subset the data `ExerciseHours` to two groups: `F` and `M` .

```{r}

excercise_fem <- ExerciseHours$Exercise[ExerciseHours$Sex == "F"]
excercise_mal <- ExerciseHours$Exercise[ExerciseHours$Sex == "M"]

length(excercise_fem)
length(excercise_mal)

## 30
## 20
```

2. Compute the observed statistic (mean difference of exercise hours for Female and Male).

```{r}

obs_stat <- mean(excercise_fem) - mean(excercise_mal)
obs_stat
## -3
```


3. Create null hypothesis distribution

3.a  Shuffle the two groups of `female` and `Male` into two samples, and find the mean difference of the two shuffled samples.

```{r}

combined_sample <- c(excercise_fem, excercise_mal)
shuffled_sample <- shuffle(combined_sample)

shuff_fem <- shuffled_sample[1:30]
shuff_mal <- shuffled_sample[31:50]

shuff_stat <- mean(shuff_fem) - mean(shuff_mal)
shuff_stat
# answers may vary
```

3.b Create the Null hypothesis Distribution using the function do_it()

```{r}
 
null_dist <- do_it(10000) * {
shuffled_sample <- shuffle(combined_sample )

shuff_fem <- shuffled_sample[1:30]
shuff_mal <- shuffled_sample[31:50]

shuff_stat <- mean(shuff_fem) - mean(shuff_mal)
}
```

3.c Plot histogram of the null distribution and show the line of the observed mean difference

```{r}
hist(null_dist ,
     xlab = "Difference in means of Exercise hous bewteen Female and Male",
     main = "Null distribution")

abline(v = obs_stat, col = "blue")
```

4. Calculate p-value 

```{r}

p_value <- pnull(obs_stat, null_dist, lower.tail = T) 
p_value
#0.1038 (# answers may vary)
```


5. Make decision/Judgment

We fail to reject the **null hypothesis**. There are not enough evidence to conclude that the average exercise hours for female students is less than the average exercise hours for male students.

$\\$


## Question 2: Baseball Hits Hypothesis Test for Correlations

Do more home runs mean more wins? In other words, is there a positive correlation between the number of home runs a team hits and the number of wins? Test this hypothesis using a permutation test. Data from the 2019 Major League Baseball (MLB) season is available in the `BaseballHits2019` dataset in the `Lock5Data` library. Make sure to extract the appropriate columns from the data.

```{r}
library(Lock5Data)
library(SDS1000)
data(BaseballHits2019)
```


**Answers:**

1. Write the `null hypothesis`  and `alternative hypothesis`  in words and in symbols.

* Null: There is no linear association between home runs and wins
* Alternative: There is a positive, linear association between home runs and wins

&nbsp;

* $H_0: \rho = 0$
* $H_A: \rho > 0$


1.b Create a scatterplot to visualize the association between `HomeRuns` and `Wins`.

```{r}
runs <- BaseballHits2019$HomeRuns
wins <- BaseballHits2019$Wins

plot(runs, wins,
     xlab = "Home Runs", ylab = "Wins", 
     main = " Home Runs vs Wins  in the 2019 MLB Season")
```

2. Compute the observed statistic (correlation between `HomeRuns` and `Wins`).

```{r}
runs_wins_corr <-  cor(runs, wins)
runs_wins_corr
```

3.a Create null hypothesis distribution
  

```{r}

null_dist <- do_it(10000) * {
 cor(runs, shuffle(wins))
}

```

3.b Plot a `histogram` of the null distribution and show the `line` of the `observed correlation` using the `abline()` function.

```{r}

hist(null_dist,
     xlab = "Correlations between Home Runs and Wins", 
     main = "Null distribution", 
     xlim= c( -1, 1))

abline(v = runs_wins_corr, col = "blue")
```

4. Calculate p-value 

```{r}
p_value <- pnull(runs_wins_corr, null_dist, lower.tail = F) 
p_value
```

5. Make a decision and state your conclusion:

Since our p-value is less than 0.05, we will reject the null hypothesis. We therefore have evidence that there is a positive correlation between wins and home runs.

$\\$



## Question 3: Comparing Multiple Samples/ The Mean Absolute Deviation (MAD) Statistic

The **mean absolute deviation** (MAD) is a `statistic` that we can calculate to compare differences among more than two groups. Suppose we had 4 groups. Then the MAD statistic is given by:


$\text{MAD} = \frac{|\bar{x}_1 - \bar{x}_2| + |\bar{x}_1 - \bar{x}_3| + |\bar{x}_1 - \bar{x}_4| + |\bar{x_2} - \bar{x}_3| + |\bar{x}_2 - \bar{x}_4| + |\bar{x}_3 - \bar{x}_4|}{6}$

 
The Lock5Data library contains the data set `TextbookCosts`. Conduct a  test to see if the mean textbook cost differs among subjects. *Hint:* Follow the steps that you did for the previous exercises, except now use the `get_MAD_stat()` function to compute your observed statistic and to generate your null distribution.

```{r}
library(SDS1000)
library(Lock5Data)
data(TextbookCosts)

```

1. Write the `null hypothesis`  and `alternative hypothesis`  in words and in symbols.

* Null: There is no difference in average textbook costs among subjects
* Alternative: There is a difference in average textbook costs among subjects

&nbsp;

**In symbols**

$H_0: \mu_{Arts} = \mu_{H} = \mu_{NS} = \mu_{SS}$

$H_A: \mu_i \ne \mu_j$ for some i, j in {Arts,H, NS,SS}


2. Compute the observed MAD statistic. Use the `get_MAD_stat()` function.


```{r}

mad_stat <-   get_MAD_stat(TextbookCosts$Cost, TextbookCosts$Field)
mad_stat
```



3.a Create null hypothesis distribution 


```{r}

null_dist <- do_it(10000) * {

  shuffled_Field <- shuffle(TextbookCosts$Field )

  get_MAD_stat(TextbookCosts$Cost, shuffled_Field)

}
```



3.b Plot a `histogram` of the null distribution and show the `line` of the `observed ` using the `abline()` function.

```{r}
hist(null_dist,
     xlab = "MAD Statistic for Textbook Costs by Field", 
     main = "Null distribution")

abline(v = mad_stat, col = "blue")
```




4. Calculate p-value 

```{r}

p_value <- pnull(mad_stat, null_dist, lower.tail = F) 
p_value
```


5. Make a decision and state your conclusion:

Since our p-value is less than 0.05, we will reject the null hypothesis. We therefore have evidence that there is a difference in textbook cost among subjects.


$\\$



# Extra practices: 

## Question 4: Caffeine Taps Hypothesis Test for the Difference of Two Means

A sample of male college students were asked to tap their fingers at a rapid rate. The sample was then divided at random into two groups of ten students each. Each student drank the equivalent of about two cups of coffee, which included about 200 mg of caffeine for the students in one group but was decaffeinated coffee for the second group. After a two hour period, each student was tested to measure finger tapping rate (taps per minute). The goal of the experiment was to determine whether caffeine produces an increase in the average tap rate.


```{r}
library(Lock5Data)
library(SDS1000)
data(CaffeineTaps)

```

1.a  Write the `null hypothesis` and `alternative hypothesis` in words and in symbols.

* Null: There is no difference in the average tapping rate between the caffeinated and decaffeinated groups 

* Alternative: The average tapping rate in the caffeinated group is different from the decaffeinated group

&nbsp;

* $H_0: \mu_{c} - \mu_{nc} = 0$
* $H_A: \mu_{c} - \mu_{nc} \neq 0$


1.b Create a boxplot to describe tap rates for `Caffeine` versus `No Caffeine`.

```{r}

boxplot(data = CaffeineTaps, Taps ~ Group)
```


1.c Subset the data `CaffeineTaps` to two groups: `Caffeine` and `NoCaffeine`.

```{r}

Caffeine <- CaffeineTaps$Taps[CaffeineTaps$Group == "Caffeine"]
NoCaffeine <- CaffeineTaps$Taps[CaffeineTaps$Group == "NoCaffeine"]

```


2. Compute the observed statistic (mean difference of tap number for the two groups).

```{r}
diff_caffeine = mean(Caffeine) - mean(NoCaffeine)
```


3.a Combine the two samples of `Caffeine` and `No Caffeine`. 

```{r}
combined_sample <- c(Caffeine, NoCaffeine)
```



3.b Create the Null hypothesis Distribution using `do_it()` function.  

```{r}
# Combine the two samples
combined_sample <- c(Caffeine, NoCaffeine)


null_dist <- do_it(10000) * {
  shuffled_sample <- shuffle(combined_sample)

  shuff_caff <- shuffled_sample[1:10]
  shuff_nocaff <- shuffled_sample[11:20]

  shuff_stat <- mean(shuff_caff) - mean(shuff_nocaff)
}
```

3.c Plot a `histogram` of the null distribution and show the `line` of the `observed mean difference` using the `abline()` function.

```{r}
hist(null_dist , 
     xlab = "Difference in Mean Tap Rate", main = "Null distribution")
abline(v = diff_caffeine, col = "blue")
```


4. Calculate p-value 

```{r}
p_value_upper <- pnull(diff_caffeine, null_dist, lower.tail = F) 
p_value_lower <- pnull(-1*diff_caffeine, null_dist) 

p_value <- p_value_upper + p_value_lower
p_value
```

5. Make a decision and state your conclusion:

Since our p-value is less than 0.05, we will reject the null hypothesis. We therefore have evidence that drinking caffeinated coffee produces an increase in the average tap rate compared to those that drink decaffeinated coffee.

$\\$

## Question 5:

We will use data on `Antibodies` (in μg/ml) production after receiving a `Vaccine` (Vaccine A, Vaccine B, Vaccine C). A hospital administered three different vaccines to six individuals and measured the antibody presence in their blood after a chosen time period. The data is saved in `patient_vaccine.csv`. Note that each patient received all three vaccines.

Is there a difference in antibody presence among the three vaccines? 

```{r}
library(SDS1000)
my_data<- read.csv("patient_vaccine.csv")
```

1. (a) Write the `null hypothesis`  and `alternative hypothesis` in words and in symbols.

* Null: The vaccines cause the same amount of antibodies to be produced 
* Alternative: At least one of the vaccines causes a different amount of antibodies to be produced

**In symbols**

$H_0: \mu_{A} = \mu_{B} = \mu_{C}$

$H_A: \mu_i \ne \mu_j$ for some i, j in {A, B, C}

1.b Create side-by-side boxplots to show the variation in antibody production among the three vaccines.

```{r}
library(SDS1000)

antibodi <-  my_data$Antibodies
vaccine  <- my_data$Vaccine

boxplot( antibodi ~ vaccine  )

```

2. Calculate the observed statistic
```{r}
obs_stat <- get_MAD_stat(antibodi ,  vaccine)
obs_stat
```

 3. Create the null distribution
 
```{r}
null_dist <- do_it(10000) * {

  # shuffle the values
  shuffled_Vaccine <- shuffle(vaccine)

  # calculate the simulated statistic
  get_MAD_stat(antibodi, shuffled_Vaccine)

}
```

4. Plot the null distribution and add a red vertical line at the observed statistic

```{r}
hist(null_dist, 
     main = "Null distribution",
     xlab = "Mean absolute deviation of group means")

#3.c# add a red vertical line at the observed statistic
abline(v = obs_stat, col = "red")
```

5. Calculate the p-value from the test

```{r}
#4# P.Value
p_value <- pnull(obs_stat, null_dist, lower.tail = FALSE)
p_value

### pv= 0.0433
```

6. State your conclusion

At significance level 0.05, we can **reject the null** and conclude the **alternative**, that there is difference between the three vaccines in terms of the amounts of antibodies they produce
