---
title: "Practice Session  6"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---

# Part 1: Hypothesis Testing Terms and Definitions

Generally speaking, hypothesis testing is a way for researchers to answer questions in a meaningful way. The methodology leverages statistical reasoning and probability to either **reject** or **fail to reject** a statement about the population as a whole. Hypothesis testing in statistics often focuses on questions about population parameters (e.g., $\mu$, $\pi$). We do not write hypotheses in terms of the sample statistics (e.g., $\bar x$, $\hat p$). This aligns with the general theme of the course, which focuses on using samples to better understand the population. 


## Practice 1:

Please match the following terms with the appropriate definition.\ 

**Terms:**

1. Null Hypothesis

2. Alternative Hypothesis

3. P-value

**Definitions:**

a. The probability of observing results at least as extreme as what we observed, assuming the null hypothesis is true.

b. A statement of interest that represents the *status quo*, or that there is no effect or difference.

c. A statement of interest that proposes there is an effect or difference.





# Part 2: Hypothesis Testing for a Single Proportion

## Practice 2.1 : Preferred Water Brand


A national sales bureau believes that 30% of consumers prefer Fiji water as their first choice for water. Specifically, they choose it over Aquafina, Sam's Choice, and tap water. The Fiji company claims that this proportion is greater than 30%. Using the `WaterTaste` data set from the Lock5Data library, run a hypothesis test to check this claim. Look at the `help` page for the data set to see which variable is appropriate to use.

```{r}
library(SDS1000)
library(Lock5Data)
data(WaterTaste)
```

1. **First**: State the null and alternative hypothesis in words



2. **Second**: Calculate the observed statistic of interest 

```{r}
# your code here #
```


3. **Third**: Create a null distribution in R using `do_it()` and `rflip()`. Plot the null distribution, and add a vertical red line at your point estimate.


```{r}
# your code here #
```


4. **Fourth**:  Calculate the p-value (probability of obtaining a result as or more extreme than what we observed) using the `pnull()` function.

```{r}
# your code here #
```


5. **Fifth**: Report the p-value. Make a decision on whether the results are statistically significant, and state your conclusion.


$\\$


## Practice 2.2: AP Multiple Choice


Does the answer choice "C" occur **less** frequently than expected? Answer this question with a hypothesis test. Use the `APMultipleChoice` data set from the `Lock5Data` library. Make sure to follow all 5 steps of hypothesis testing. *Hint:* when defining the null hypothesis, consider how likely each answer choice
would be if answers were selected at random.

```{r}
library(SDS1000)
library(Lock5Data)
data(APMultipleChoice)
```


1. **First**: State the null and alternative hypothesis in words


2. **Second**: Calculate the observed statistic of interest

```{r}
# your code here #
```


3. **Third**: Create a null distribution in R using `do_it()` and `rflip()`. Plot the null distribution, and add a vertical red line at the observed number of questions that had 'C' as the answer. *Note*: there were 400 multiple-choice questions in the data set.


```{r}
# your code here #
```



4. **Fourth**: Calculate the p-value (probability of obtaining a result as or more extreme than what we observed) using the `pnull()` function. 

```{r}
# your code here #
```



5. **Fifth**: Report the p-value. Make a decision on whether the results are statistically significant, and state your conclusion.



## Practice 2.3: Caffeine Taps

A sample of college students were asked to tap their fingers at a rapid rate. The sample was then divided at random into two groups of ten students each. Each student drank the equivalent of about two cups of coffee, which included about 200 mg of caffeine for the students in one group but was decaffeinated coffee for the second group. After a two hour period, each student was tested to measure finger tapping rate (taps per minute). The goal of the experiment was to determine whether caffeine produces an increase in the average tap rate.


1. Write the `null hypothesis`  and `alternative hypothesis`  in words and in symbols.


2. Create a boxplot to describe tap rates for `Caffeine` versus `No Caffeine`.

```{r}
# your code here #
```


3. Find some favorite statistics to visualize the number of taps for the `Caffeine` and `No Caffeine` group. You might find the function: `mosaic::favstats` useful.
*Note*: you can search online for this function arguments.

```{r}
# your code here #
```
 
 
4. Subset the data `CaffeineTaps` to two groups: `Caffeine` and `NoCaffeine`.

```{r}
# your code here #
```


5. Compute the observed statistic (mean difference of tap number for the two groups).

```{r}
# your code here #
```


6. Create null hypothesis distribution

a. Shuffle the two groups of `Caffeine` and `No Caffeine` into two samples, and find the mean difference of the two shuffled samples.

```{r}
# your code here #
```

b. Create the Null hypothesis Distribution using `do_it()` function.  

```{r}
# your code here #
```

c. Plot a `histogram` of the null distribution and show the `line` of the `observed mean difference` using the `abline()` function.

```{r}
# your code here #
```


7. Calculate p-value 

```{r}
# your code here #
```


8. Make a decision and state your conclusion:


$\\$


# Part 3:  Extra Practices 


## Practice 3.1:  Exploring the `pnull()` Function

In this exercise, we will practice using some basic R coding to calculate a p-value without using the `pnull()` function. 

1. Consider null distribution and point estimate from the `AP Multiple Choice` exercise. Using the `<=` operator, write a single line of code that will show which values in the null distribution are **less than** the point estimate. Run this code and observe what the values look like. Save this result to a vector. *Hint:* don't over think this part. Just write the inequality as you would in math class.

```{r}
# your code here #
```


2. Using your vector from above, count how many values from the null distribution are **less than** the point estimate. Use the `sum()` function for this. *Hint:* the `==` operator can be used within the `sum()` function to compare values.

```{r}
# your code here #
```


3. Using your result from part 2.), find the  proportion of observations in the null distributions that are **less than** the point estimate. Compare to the number from `pnull()` earlier.

```{r}
# your code here #
```


## Practice 3.2: UFO Shapes

UFOs, or Unidentified Flying Objects have become commonplace in media and popular culture. Some witnesses claim that the objects have a cylindrical shape, while others believe they are more spherical. Load the data `ufo_data.csv` using the function ` read.csv`.

Run a hypothesis test to check if the claim that UFOs are spherical is different from $0.75$.

1. State the null and alternative hypothesis in words:

2. Calculate the observed statistic of interest 


```{r}
library(SDS1000)
library(Lock5Data)

# your code here #
```


3. Create a null distribution in R using `do_it()` and `rflip()`. Plot the null distribution, and add a vertical red line at your observed total number of spheres.

```{r}
# your code here #
```


4. Calculate the p-value (probability of obtaining a result as or more extreme than what we observed) using the `pnull()` function.

```{r}
# your code here #
```


5. Report the p-value. Make a decision on whether the results are statistically significant, and state your conclusion.





