---
title: "Class 4 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---




## Part 1: More practice analyzing quantitative data:  Internet pings 


When an individual visits a webpage a 'ping' is generated

Below is a random sample of ping counts from 7 people who pinged a website at least once:

   12, 45, 6, 4, 158, 10, 59 

Question: What is the mean and median ping count in this sample? 

Also, see if you can calculate these values without using the `mean()` and 
`median()` functions. Hint: the `sum()`, `length()`, and `sort()` functions may 
be useful.

```{r}

# Create the ping data as a vector 




# Calculate the mean ping count




# Calculate the median ping count





```




$\\$





## Part 2: Measures variability: the variance and standard deviation 


Let's continue to analyze the [Gapminder data set](https://www.youtube.com/watch?v=jbkSRLYSojo) 
which contains information about different countries. Recall, the key variables 
in the Gapminder data set include:

- `country`: The country of the observation
- `continent`: The continent of the observation
- `year`: The year of the observation
- `lifeExp`: Life expectancy at birth (in years)
- `pop`: Population of the country
- `gdpPercap`: Gross domestic product per capita (in dollars)

The code below loads the Gapminder data set and show the first 6 rows of the data. 



```{r load_gapminder_data}

load("gapminder_2007.rda")

# Show the first 5 rows of the data
head(gapminder_2007)


```





$\\$




### Part 2.1: Review - histograms, mean, and median


To start, please create a histogram of the GDP per capita. Then answer these
questions:

1. What shape does this histogram have? Is it symmetric, right-skewed, or left-skewed?

2. Are there any extreme outliers in the data? 

3. Based on the shape of the histogram, do you expect the mean or median value to be larger?

Finally, use R to calculate the mean and the median and add these values as
red and blue vertical lines to your plot to check your answers.


```{r histogram_life_expectancy}

# extract the gdp per capita to an object


# Create a histogram of GDP per capita







# Calculate the mean and median and add them as vertical lines to the histogram








```


**Answer:** 

1. 

2. 


3. 



$\\$







### Part 2.2: Calculating the variance and standard deviation


Let's now calculate the variance and standard deviation of the GDP per capita
across the countries in the data.

Let's now calculate the variance and standard deviation of the life expectancy data for the year 2007.

```{r variance_standard_deviation_life_expectancy}

# Calculate the variance of life expectancy




# Calculate the standard deviation of life expectancy




```





$\\$





## Part 3: z-scores

LeBron James is a one of the greatest basketball player of all time. The table
below shows some of his statistics in the 2011 NBA season, along with the league
average (mean) and standard deviation of these statistics. 

Using the data in the table, calculate the z-score for each statistic to determine
which statistic was most impressive relative to his peers. 


| statistic       | LeBron James | League mean | League sd |
|-----------------|--------------|-------------|-----------|
| Field goal %    |    0.510     |    0.464    |  0.053    |
| Points scored   |     2111     |     994     |    414	   |
| Assists         |     554      |     220     |    170    |
| Steals          |     124      |     68.2    |   31.5    |


```{r z_scores_lebron}








```



**Answer**







$\\$







## Part 4: Percentile rank

The percentile rank is the percentage of values in a data set less than a 
particular value x. 

We can calculate the percentile rank using the SDS1000 function `ptail(x, vals)`
to assess the proportion of data less than or equal to a particular value `x`.
Note, this function returns the proportion (rather than percentage) of values 
less than or equal to x. 



```{r ptail}

library(SDS1000)

ping_data_small <- c(59, 6, 158, 12, 45)


# proportion of data less than or equal to 15



# proportion of data greater than or equal to 15




```






$\\$







What proportion of countries have a GDP per capita greater than (or equal to) $30,000?


```{r ptail_gdp}


# proportion of countries that have a GDP per capita greater than $30,000




# What proportion of the data is less than or equal to the median?





```








