---
title: "Class 4 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---




Today we're going to analyze data from the [Gapminder data set](https://www.youtube.com/watch?v=jbkSRLYSojo) 
which contains information about different countries. We will focus on data from 
the year 2007 only, which is loaded below.

Key variables in the Gapminder data set include:

- `country`: The country of the observation
- `continent`: The continent of the observation
- `year`: The year of the observation
- `lifeExp`: Life expectancy at birth (in years)
- `pop`: Population of the country
- `gdpPercap`: Gross domestic product per capita (in dollars)

The code below loads the Gapminder data set and show the first 6 rows of the data. 



```{r load_gapminder_data}

load("gapminder_2007.rda")

# Show the first 5 rows of the data
head(gapminder_2007)


```




$\\$




# Part 1: More practice analyzing categorical data

For more practice analyzing categorical data, please do the following analysis.

1. Extract information about the continents variable from the `gapminder_2007`
data frame and store the result in a new object called `continents`.

2. Create a frequency table showing how many countries are in each continent.

3. Create a barplot of the frequency of countries in each continent.



```{r practice_categorical_data}

# 1. Extract information about the continents variable




# 2. Create a frequency table showing how many countries are in each continent





# 3. Create a barplot of the frequency of countries in each continent






```




$\\$






# Part 2: Analyzing a single quantitative variable


We will now explore analyzing a single quantitative variable. To do this we will
continue to look at the Gap Minder data but we will look at the life expectancy which
is a quantitative variable. The code below extract information on life expectancy
from the `gapminder_2007` data frame and stores it in a new object called `life_expectancy`.


```{r extract_life_expectancy}

# Extract a vector of life expectancy data

life_expectancy <- gapminder_2007$lifeExp


```





$\\$




### 2.1 Histograms

Let's now create a histogram of the life expectancy data for the year 2007. We
can do this using the `hist()` function along with the vector of life expectancy
data we created in part 2.1 which is stored in the object `life_expectancy`.

Questions:

1. What shape does this histogram have? Is it symmetric, right-skewed, or left-skewed?
2. Are there any extreme outliers in the data? 


```{r histogram_life_expectancy}

# Create a histogram of life expectancy






```


**Answer:** 

1. 

2.







$\\$






# Part 3: Descriptive statistics for the central tendency: the mean and median

We can use the `mean()` and `median()` functions to calculate the mean and median
of a sample of data. 

Let's calculate the mean and median of the life expectancy data for the year 2007.

Are these values what you would expect based on the skew of the data?

```{r mean_median_life_expectancy}

# Calculate the mean of life expectancy



# Calculate the median of life expectancy



```



$\\$



### 3.2 Showing the mean and median on a histogram


We can add lines to the histogram to show where the mean and median are located using
the abline`()` function.

Let's add lines for the mean and median to the histogram we created in part 2.1.


```{r add_mean_median_lines}

# Create a histogram of life expectancy




# Add a line for the mean



# Add a line for the median



# Add a legend




```



$\\$





### 3.3 Internet pings

When an individual visits a webpage a 'ping' is generated

Below is a random sample of ping counts from 7 people who pinged a website at least once:

   12, 45, 6, 4, 158, 10, 59 

Question: What is the mean and median ping count in this sample? 

Also, see if you can calculate these values without using the `mean()` and 
`median()` functions. Hint: the `sum()`, `length()`, and `sort()` functions may 
be useful.

```{r}

# Create the ping data as a vector 




# Calculate the mean ping count




# Calculate the median ping count





```




$\\$




# Part 4: Measures variability: the variance and standard deviation 

Let's now calculate the variance and standard deviation of the life expectancy data for the year 2007.

```{r variance_standard_deviation_life_expectancy}

# Calculate the variance of life expectancy




# Calculate the standard deviation of life expectancy




```









