---
title: "S&DS 1000 Class Summaries"
subtitle: "Hypothesis Tests and Confidence Intervals"
format:
  pdf:
    toc: true
    toc-depth: 2
    number-sections: false
    colorlinks: true
    geometry: margin=1in
    fontsize: 11pt
    include-in-header:
      text: |
        \usepackage{titlesec}
        \usepackage{xcolor}
        \definecolor{sectiongreen}{HTML}{00A86B}
        \definecolor{subsectionblue}{HTML}{2c5282}
        \titleformat{\section}{\LARGE\bfseries\color{sectiongreen}}{}{0em}{}[\vspace{0.3ex}\titlerule]
        \titlespacing{\section}{0pt}{4ex}{2ex}
        \titleformat{\subsection}{\Large\bfseries\color{subsectionblue}}{}{0em}{}
        \titlespacing{\subsection}{0pt}{3ex}{1ex}
        \titleformat{\subsubsection}{\normalsize\bfseries}{}{0em}{}
        \titlespacing{\subsubsection}{0pt}{2ex}{0.5ex}
execute:
  echo: true
  message: false
  warning: false
  fig-width: 6
  fig-height: 3.5
---

```{r}
#| label: setup
#| include: false
library(SDS1000)

# knitr-safe do_it() — avoids lazyeval rendering issues in knitr
do_it <- function(n) structure(list(num_times = n), class = "do_it_knitr")
`*.do_it_knitr` <- function(e1, e2) {
  expr <- match.call()[[3]]
  replicate(e1$num_times, eval(expr, envir = parent.frame()))
}
```

{{< pagebreak >}}

# Randomization Hypothesis Tests

This guide covers every randomization (simulation-based) hypothesis test from
S&DS 1000. Each test follows the same five steps used in class:

1. **State** the null and alternative hypotheses
2. **Calculate** the observed test statistic
3. **Create** a null distribution by simulation
4. **Calculate** the p-value
5. **Make a decision**

The core idea is the same for every test: if the null hypothesis were true, how
extreme would our observed result be? We answer this by repeatedly simulating
data under the null and seeing where our observed statistic falls.

::: {.callout-tip}
**Key SDS1000 functions used in this guide.**
`do_it(n) * { expr }` repeats an expression `n` times and collects the
results. `rflip(n, p)` simulates `n` coin flips with probability `p`.
`rroll(n, prob)` simulates rolling a die `n` times. `shuffle(x)` randomly
permutes a vector. `ptail(obs, null, lower.tail)` computes the p-value from
a null distribution.
:::

---

## One Proportion

**When to use:** One categorical variable with two outcomes; test whether the
population proportion π equals a specific value π₀. The null distribution is
created by simulating coin flips under the assumption that H₀ is true using
`rflip()`.

**Example (class 12):** Doris the dolphin and Buzz the dolphin were tested to
see whether Buzz could identify which lever Doris was pointing to. In 16
trials, Buzz got the correct answer 15 times. Is there evidence that Buzz is
doing better than random chance?

### Step 1: State the hypotheses

$$H_0: \pi = 0.50 \qquad H_A: \pi > 0.50$$

### Step 2: Calculate the observed statistic

```{r}
#| label: rt-one-prop-obs
n_trials <- 16
obs_stat <- 15   # number of correct answers
obs_stat
```

### Step 3: Create the null distribution

Under H₀, each trial is like a fair coin flip. Use `rflip()` to simulate
10,000 experiments:

```{r}
#| label: rt-one-prop-null
set.seed(5731)

null_dist <- do_it(10000) * {
  rflip(n_trials, 0.50)
}

hist(null_dist, breaks = 50,
     main = "Null Distribution — Number of Correct Answers",
     xlab = "Number Correct (out of 16)",
     col = "steelblue", border = "white")
abline(v = obs_stat, col = "red", lwd = 2)
```

### Step 4: Calculate the p-value

```{r}
#| label: rt-one-prop-pval
p_value <- ptail(obs_stat, null_dist, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision

Since p = `r format(round(p_value, 4), scientific = FALSE)` < 0.05, we
**reject** H₀ and conclude that Buzz appears to be doing better than random
chance.

---

**Second example (class 13):** A lie detector correctly identified 31 out of
48 participants as lying. Is this evidence the detector is more than 60%
accurate?

$$H_0: \pi = 0.60 \qquad H_A: \pi > 0.60$$

```{r}
#| label: rt-one-prop-lie
set.seed(9021)

n        <- 48
obs_phat <- 31 / 48

null_dist <- do_it(10000) * {
  rflip(n, 0.60) / n
}

hist(null_dist, breaks = 60,
     main = "Null Distribution — Proportion Correct",
     xlab = "Proportion Correct",
     col = "steelblue", border = "white")
abline(v = obs_phat, col = "red", lwd = 2)

p_value <- ptail(obs_phat, null_dist, lower.tail = FALSE)
p_value
```

Since p = `r round(p_value, 3)` > 0.05, we **fail to reject** H₀. There is
not sufficient evidence that the lie detector exceeds 60% accuracy.

---

## Two Proportions

**When to use:** A binary outcome measured in two independent groups; test
whether the two population proportions are equal.

**Key idea:** Under H₀ (π₁ = π₂), both groups share the same true proportion.
We estimate this **overall proportion** by pooling the two samples:

$$\hat{p}_\text{overall} = \frac{x_1 + x_2}{n_1 + n_2}$$

We then simulate each group independently using this pooled rate with
`rflip()`, mimicking what we would observe if H₀ were true.

**Example:** In a clinical trial, 245 patients received a new vaccine and 255
received a placebo. Of the vaccinated group 32 got the flu; of the placebo
group 56 got the flu. Is there evidence that the vaccine reduces the flu rate?

### Step 1: State the hypotheses

Let group 1 = vaccine, group 2 = placebo.
$$H_0: \pi_1 = \pi_2 \qquad H_A: \pi_1 < \pi_2$$

### Step 2: Calculate the observed statistic

```{r}
#| label: rt-two-prop-obs
x1 <- 32;  n1 <- 245   # vaccinated
x2 <- 56;  n2 <- 255   # placebo

obs_phat1 <- x1 / n1
obs_phat2 <- x2 / n2
obs_stat  <- obs_phat1 - obs_phat2
obs_stat
```

### Step 3: Create the null distribution

```{r}
#| label: rt-two-prop-null
set.seed(4412)

overall_prop <- (x1 + x2) / (n1 + n2)

null_dist <- do_it(10000) * {
  sim_phat1 <- rflip(n1, overall_prop) / n1
  sim_phat2 <- rflip(n2, overall_prop) / n2
  sim_phat1 - sim_phat2
}

hist(null_dist, breaks = 80,
     main = "Null Distribution — Difference in Proportions",
     xlab = expression(hat(p)[1] - hat(p)[2]),
     col = "steelblue", border = "white")
abline(v = obs_stat, col = "red", lwd = 2)
```

### Step 4: Calculate the p-value

```{r}
#| label: rt-two-prop-pval
p_value <- ptail(obs_stat, null_dist, lower.tail = TRUE)
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 4)` < 0.05, we **reject** H₀ and conclude that
the vaccine is associated with a significantly lower flu rate.

---

## More Than Two Proportions

**When to use:** One categorical variable with three or more outcomes; test
whether the population proportions match a set of expected values. The
observed statistic is the chi-squared statistic; the null distribution is
created by simulating random samples using `rroll()`.

**Example (class 23):** A Yale professor examined the birth months of 198
randomly selected Yale students. Are students equally likely to be born in any
month?

### Step 1: State the hypotheses

$$H_0: \pi_{Jan} = \pi_{Feb} = \cdots = \pi_{Dec} = \tfrac{1}{12}$$
$$H_A: \text{at least one monthly proportion differs from } \tfrac{1}{12}$$

### Step 2: Calculate the observed statistic

```{r}
#| label: rt-multi-prop-obs
observed_counts <- c(14, 11, 21, 17, 15, 13, 19, 16, 18, 22, 14, 18)
names(observed_counts) <- month.abb

expected_counts <- rep(sum(observed_counts) / 12, 12)

obs_stat <- sum((observed_counts - expected_counts)^2 / expected_counts)
obs_stat
```

### Step 3: Create the null distribution

```{r}
#| label: rt-multi-prop-null
set.seed(3341)

n_students <- sum(observed_counts)

null_dist <- do_it(10000) * {
  sim_counts <- rroll(n_students, prob = rep(1/12, 12))
  sum((sim_counts - expected_counts)^2 / expected_counts)
}

hist(null_dist, breaks = 60,
     main = "Null Distribution — Chi-squared Statistic",
     xlab = "Chi-squared value",
     col = "steelblue", border = "white")
abline(v = obs_stat, col = "red", lwd = 2)
```

### Step 4: Calculate the p-value

```{r}
#| label: rt-multi-prop-pval
p_value <- ptail(obs_stat, null_dist, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 3)` > 0.05, we **fail to reject** H₀. There is
no evidence that Yale students' birth months deviate from a uniform
distribution.

---

## One Mean

**When to use:** A single quantitative variable; test whether the population
mean μ equals a specific value μ₀. We use a **bootstrap-based procedure**:
shift the data so its mean equals μ₀, then resample from that shifted data.

**Key idea:**
```r
data_H0 <- my_data - mean(my_data) + mu_0
```

**Example:** It is widely assumed that the average human body temperature is
98.6°F. A sample of 50 body temperatures had a mean of 98.25°F. Is there
evidence that the true mean differs from 98.6°F?

### Step 1: State the hypotheses

$$H_0: \mu = 98.6 \qquad H_A: \mu \ne 98.6$$

### Step 2: Calculate the observed statistic

```{r}
#| label: rt-one-mean-obs
set.seed(9910)
body_temps <- rnorm(50, mean = 98.25, sd = 0.73)

obs_mean <- mean(body_temps)
obs_mean
```

### Step 3: Create the null distribution

```{r}
#| label: rt-one-mean-null
set.seed(3345)

mu_0    <- 98.6
data_H0 <- body_temps - mean(body_temps) + mu_0
n       <- length(body_temps)

null_dist <- do_it(10000) * {
  mean(sample(data_H0, n, replace = TRUE))
}

hist(null_dist, breaks = 80,
     main = "Null Distribution — Mean Body Temperature",
     xlab = "Mean Temperature (°F)",
     col = "steelblue", border = "white")
abline(v = obs_mean, col = "red", lwd = 2)
```

### Step 4: Calculate the p-value

Two-sided test — count values at least as far from μ₀ as the observed mean:

```{r}
#| label: rt-one-mean-pval
margin <- abs(obs_mean - mu_0)

p_value <- ptail(mu_0 - margin, null_dist, lower.tail = TRUE) +
           ptail(mu_0 + margin, null_dist, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 4)` < 0.05, we **reject** H₀ and conclude that
there is evidence the true mean body temperature differs from 98.6°F.

---

## Two Means

**When to use:** A quantitative variable measured in two independent groups;
test whether the two group means are equal. The null distribution is created
by using `shuffle()` to randomly reassign group labels.

**Example (class 17):** A randomized controlled trial gave calcium supplements
to 10 men (treatment) and a placebo to 11 men (control). The outcome is the
decrease in blood pressure. Is there evidence that calcium produces a greater
decrease?

### Step 1: State the hypotheses

$$H_0: \mu_\text{treat} = \mu_\text{control} \qquad H_A: \mu_\text{treat} > \mu_\text{control}$$

### Step 2: Calculate the observed statistic

```{r}
#| label: rt-two-means-obs
treat   <- c( 7, -4, 18, 17, -3, -5,  1, 10, 11, -2)
control <- c(-1, 12, -1, -3,  3, -5,  5,  2, -11, -1, -3)

obs_stat <- mean(treat) - mean(control)
obs_stat
```

### Step 3: Create the null distribution

```{r}
#| label: rt-two-means-null
set.seed(6174)

combined_data <- c(treat, control)

null_dist <- do_it(10000) * {
  shuff         <- shuffle(combined_data)
  shuff_treat   <- shuff[1:10]
  shuff_control <- shuff[11:21]
  mean(shuff_treat) - mean(shuff_control)
}

hist(null_dist, breaks = 100,
     main = "Null Distribution — Difference in Means",
     xlab = "Difference in mean blood pressure decrease (mmHg)",
     col = "steelblue", border = "white")
abline(v = obs_stat, col = "red", lwd = 2)
```

### Step 4: Calculate the p-value

```{r}
#| label: rt-two-means-pval
p_value <- ptail(obs_stat, null_dist, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 3)` is just above 0.05, we **fail to reject** H₀
at the α = 0.05 level. The evidence for a calcium effect is suggestive but not
conclusive.

---

## More Than Two Means

**When to use:** A quantitative variable measured across three or more
independent groups; test whether all group means are equal. The observed
statistic is the **mean absolute deviation (MAD)** of group means. The null
distribution is created by using `shuffle()` on the group labels.

**Example (class 17/18):** Hope College students were timed completing a
Sudoku-like puzzle, grouped into four majors. Is there a difference in mean
completion time?

### Step 1: State the hypotheses

$$H_0: \mu_{as} = \mu_{ns} = \mu_{ss} = \mu_{ah}$$
$$H_A: \text{at least one group mean differs}$$

### Step 2: Calculate the observed statistic

```{r}
#| label: rt-multi-means-obs
set.seed(4455)
completion_times <- c(rnorm(10, mean = 22, sd = 5),
                      rnorm(10, mean = 20, sd = 5),
                      rnorm(10, mean = 26, sd = 5),
                      rnorm(10, mean = 21, sd = 5))
majors <- rep(c("Applied Sci", "Natural Sci", "Social Sci", "Arts/Hum"),
              each = 10)

boxplot(completion_times ~ majors,
        main = "Completion Time by Major",
        ylab = "Time (seconds)",
        col  = c("lightblue", "lightgreen", "lightpink", "lightyellow"))

obs_stat <- get_MAD_stat(completion_times, majors)
obs_stat
```

### Step 3: Create the null distribution

```{r}
#| label: rt-multi-means-null
set.seed(8812)

null_dist <- do_it(10000) * {
  get_MAD_stat(completion_times, shuffle(majors))
}

hist(null_dist, breaks = 100,
     main = "Null Distribution — MAD Statistic",
     xlab = "Mean absolute deviation of group means",
     col = "steelblue", border = "white")
abline(v = obs_stat, col = "red", lwd = 2)
```

### Step 4: Calculate the p-value

```{r}
#| label: rt-multi-means-pval
p_value <- ptail(obs_stat, null_dist, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 3)`, we
`r ifelse(p_value < 0.05, "**reject**", "**fail to reject**")` H₀ at the
α = 0.05 level.

::: {.callout-tip}
**What is the MAD statistic?** `get_MAD_stat()` computes the mean of all
pairwise absolute differences between group means:
$$MAD = \frac{1}{\binom{k}{2}} \sum_{i < j} |\bar{x}_i - \bar{x}_j|$$
It is sensitive to any difference between group means, making it a natural
choice for a general alternative hypothesis. Any statistic can be used —
for example, `get_F_stat()` would give the ANOVA F-statistic.
:::

---

## Correlation

**When to use:** Two quantitative variables; test whether the population
correlation ρ equals zero. The null distribution is created by using
`shuffle()` on one variable — this breaks any real relationship while
preserving the marginal distributions.

**Example (class 18):** A dataset of 77 breakfast cereals recorded sugar
content and calorie content. Is there evidence of a positive correlation?

### Step 1: State the hypotheses

$$H_0: \rho = 0 \qquad H_A: \rho > 0$$

### Step 2: Calculate the observed statistic

```{r}
#| label: rt-corr-obs
set.seed(1123)
n        <- 77
sugar    <- runif(n, 0, 15)
calories <- 90 + 3.5 * sugar + rnorm(n, sd = 12)

plot(sugar, calories,
     main = "Calories vs Sugar in Cereals",
     xlab = "Sugar (grams)", ylab = "Calories")

obs_stat <- cor(sugar, calories)
obs_stat
```

### Step 3: Create the null distribution

```{r}
#| label: rt-corr-null
set.seed(5671)

null_dist <- do_it(10000) * {
  cor(shuffle(sugar), calories)
}

hist(null_dist, breaks = 80,
     main = "Null Distribution — Correlation",
     xlab = "Correlation",
     col = "steelblue", border = "white")
abline(v = obs_stat, col = "red", lwd = 2)
```

### Step 4: Calculate the p-value

```{r}
#| label: rt-corr-pval
p_value <- ptail(obs_stat, null_dist, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision

Since p ≈ `r round(p_value, 4)`, we **reject** H₀ and conclude that there is
evidence of a positive correlation between sugar and calorie content in cereals.

---

## Simple Linear Regression

**When to use:** Two quantitative variables where you want to test whether the
slope of a linear regression model is significantly different from zero. The
null distribution is created by using `shuffle()` on the response variable.

**Example (class 25):** Data on cigarette consumption per capita and lung
cancer rates across U.S. states. Is there evidence of a positive slope?

### Step 1: State the hypotheses

$$H_0: \beta_1 = 0 \qquad H_A: \beta_1 > 0$$

### Step 2: Calculate the observed statistic

```{r}
#| label: rt-reg-obs
set.seed(2947)
cigs_per_capita <- runif(44, min = 10, max = 45)
cancer_rate     <- 2 + 0.005 * cigs_per_capita * 1000 + rnorm(44, sd = 5)

plot(cigs_per_capita, cancer_rate,
     xlab = "Cigarettes per Capita (hundreds)",
     ylab = "Lung Cancer Rate per 100,000",
     main = "Lung Cancer Rate vs. Cigarette Consumption")

lung_lm   <- lm(cancer_rate ~ cigs_per_capita)
obs_slope <- coef(lung_lm)["cigs_per_capita"]
obs_slope
```

### Step 3: Create the null distribution

```{r}
#| label: rt-reg-null
set.seed(3318)

null_dist <- do_it(10000) * {
  null_lm <- lm(shuffle(cancer_rate) ~ cigs_per_capita)
  coef(null_lm)["cigs_per_capita"]
}

hist(null_dist, breaks = 80,
     main = "Null Distribution — Regression Slope",
     xlab = "Slope",
     col = "steelblue", border = "white")
abline(v = obs_slope, col = "red", lwd = 2)
```

### Step 4: Calculate the p-value

```{r}
#| label: rt-reg-pval
p_value <- ptail(obs_slope, null_dist, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision

Since p ≈ `r round(p_value, 4)`, we **reject** H₀ and conclude there is
evidence of a positive relationship between cigarette consumption and lung
cancer rate.

---

## Quick Reference

| Test | Observed statistic | How to simulate the null | p-value |
|---|---|---|---|
| One proportion | Count or proportion | `do_it(n) * { rflip(n, pi_0) }` | `ptail(obs, null, lower.tail = ...)` |
| Two proportions | $\hat{p}_1 - \hat{p}_2$ | `do_it(n) * { rflip(n1,p_pool)/n1 - rflip(n2,p_pool)/n2 }` | `ptail(obs, null, lower.tail = ...)` |
| Multiple proportions | $\sum(O-E)^2/E$ | `do_it(n) * { rroll(n, expected_p) }` | `ptail(obs, null, lower.tail = FALSE)` |
| One mean | $\bar{x}$ | Shift data to μ₀, then resample | two-sided sum of tails |
| Two means | $\bar{x}_1 - \bar{x}_2$ | `do_it(n) * { shuff <- shuffle(combined); ... }` | `ptail(obs, null, lower.tail = ...)` |
| More than two means | MAD: `get_MAD_stat()` | `do_it(n) * { get_MAD_stat(y, shuffle(groups)) }` | `ptail(obs, null, lower.tail = FALSE)` |
| Correlation | $r =$ `cor(x, y)` | `do_it(n) * { cor(shuffle(x), y) }` | `ptail(obs, null, lower.tail = ...)` |
| Regression slope | $\hat{\beta}_1$ from `lm()` | `do_it(n) * { coef(lm(shuffle(y) ~ x))[2] }` | `ptail(obs, null, lower.tail = ...)` |

{{< pagebreak >}}

# Randomization Confidence Intervals

A **bootstrap confidence interval** estimates the uncertainty of a statistic by
repeatedly resampling the observed data *with replacement*. Because it relies on
the data itself rather than a mathematical formula, it works for almost any
statistic and requires very few distributional assumptions.

Every bootstrap CI in this guide follows the same three steps:

1. **Create a bootstrap distribution** — use `do_it()` to resample with replacement
   many times, computing the statistic each time.
2. **Calculate the bootstrap SE** — `SE_boot <- sd(boot_dist)`.
3. **Construct the CI** — `obs_stat ± cnorm(C) × SE_boot`.

::: {.callout-tip}
**Key SDS1000 functions.** `do_it(n) * { expr }` repeats an expression `n`
times. `sample(data, n, replace = TRUE)` resamples with replacement.
`get_proportion(v, "cat")` computes a proportion. `resample_pairs(v1, v2)`
resamples two paired vectors together. `cnorm(C)` gives the normal critical
value for confidence level C.
:::

---

## One Proportion

**When to use:** A single categorical variable with two outcomes; estimate the
population proportion π.

**Example (class 11):** A survey asked 2625 people whether they agreed with
"There is only one true love for each person." 1812 disagreed. Compute a 90%
bootstrap confidence interval for the proportion who disagreed.

### Step 1: Compute the observed statistic

```{r}
#| label: rci-one-prop-obs
believes_in_true_love <- c(rep("FALSE", 1812), rep("TRUE", 2625 - 1812))

obs_prop <- get_proportion(believes_in_true_love, "FALSE")
obs_prop
```

### Step 2: Create the bootstrap distribution

```{r}
#| label: rci-one-prop-boot
set.seed(2292)

n <- length(believes_in_true_love)

boot_dist <- do_it(10000) * {
  boot_sample <- sample(believes_in_true_love, n, replace = TRUE)
  get_proportion(boot_sample, "FALSE")
}

hist(boot_dist, breaks = 30,
     main = "Bootstrap Distribution — Proportion Who Disagreed",
     xlab = "Proportion",
     col = "steelblue", border = "white")
```

### Step 3: Compute the SE and construct the CI

```{r}
#| label: rci-one-prop-ci
SE_boot <- sd(boot_dist)

CI <- c(obs_prop - cnorm(0.90) * SE_boot,
        obs_prop + cnorm(0.90) * SE_boot)
CI
```

We are 90% confident that the true proportion of people who disagree that
there is only one true love for each person is between
`r round(CI[1], 3)` and `r round(CI[2], 3)`.

---

## One Mean

**When to use:** A single quantitative variable; estimate the population mean μ.

**Example (class 11/12):** A sample of 50 body temperatures from Mackowiak
et al. (1992). Compute a 95% bootstrap confidence interval for the mean
body temperature.

### Step 1: Compute the observed statistic

```{r}
#| label: rci-one-mean-obs
set.seed(9910)
body_temps <- rnorm(50, mean = 98.25, sd = 0.73)

obs_mean <- mean(body_temps)
obs_mean
```

### Step 2: Create the bootstrap distribution

```{r}
#| label: rci-one-mean-boot
set.seed(4471)

n <- length(body_temps)

boot_dist <- do_it(10000) * {
  mean(sample(body_temps, n, replace = TRUE))
}

hist(boot_dist, breaks = 80,
     main = "Bootstrap Distribution — Mean Body Temperature",
     xlab = "Mean Temperature (°F)",
     col = "steelblue", border = "white")
```

### Step 3: Compute the SE and construct the CI

```{r}
#| label: rci-one-mean-ci
SE_boot <- sd(boot_dist)

CI_95 <- c(obs_mean - cnorm(0.95) * SE_boot,
           obs_mean + cnorm(0.95) * SE_boot)
CI_80 <- c(obs_mean - cnorm(0.80) * SE_boot,
           obs_mean + cnorm(0.80) * SE_boot)
CI_99 <- c(obs_mean - cnorm(0.99) * SE_boot,
           obs_mean + cnorm(0.99) * SE_boot)

CI_80; CI_95; CI_99
```

We are 95% confident that the mean human body temperature is between
`r round(CI_95[1], 2)`°F and `r round(CI_95[2], 2)`°F.

::: {.callout-tip}
**Wider CI = more confident.** A 99% CI is wider than a 95% CI, which is wider
than an 80% CI. There is always a trade-off: greater confidence requires
sacrificing precision.
:::

---

## Difference of Two Means

**When to use:** A quantitative variable measured in two independent groups;
estimate $\mu_1 - \mu_2$. Resample each group separately with replacement,
keeping group sizes fixed.

**Example:** Using the calcium supplement study from class 17, compute a 95%
bootstrap confidence interval for the difference in mean blood pressure
decrease (treatment − control).

### Step 1: Compute the observed statistic

```{r}
#| label: rci-two-means-obs
treat   <- c( 7, -4, 18, 17, -3, -5,  1, 10, 11, -2)
control <- c(-1, 12, -1, -3,  3, -5,  5,  2, -11, -1, -3)

obs_diff <- mean(treat) - mean(control)
obs_diff
```

### Step 2: Create the bootstrap distribution

```{r}
#| label: rci-two-means-boot
set.seed(6174)

n_treat   <- length(treat)
n_control <- length(control)

boot_dist <- do_it(10000) * {
  boot_treat   <- sample(treat,   n_treat,   replace = TRUE)
  boot_control <- sample(control, n_control, replace = TRUE)
  mean(boot_treat) - mean(boot_control)
}

hist(boot_dist, breaks = 80,
     main = "Bootstrap Distribution — Difference in Means",
     xlab = "Difference in mean blood pressure decrease (mmHg)",
     col = "steelblue", border = "white")
```

### Step 3: Compute the SE and construct the CI

```{r}
#| label: rci-two-means-ci
SE_boot <- sd(boot_dist)

CI <- c(obs_diff - cnorm(0.95) * SE_boot,
        obs_diff + cnorm(0.95) * SE_boot)
CI
```

The 95% CI includes zero, consistent with the hypothesis test result. However,
the interval also extends well above zero, meaning a meaningful calcium effect
remains plausible.

---

## Correlation

**When to use:** Two paired quantitative variables; estimate the population
correlation ρ. Because x and y are paired, use `resample_pairs()` so that
every resampled (x, y) pair stays matched.

**Example (class 18):** Sugar content and calorie content for a sample of 77
breakfast cereals. Compute a 95% bootstrap confidence interval for the
correlation.

### Step 1: Compute the observed statistic

```{r}
#| label: rci-corr-obs
set.seed(1123)
n        <- 77
sugar    <- runif(n, 0, 15)
calories <- 90 + 3.5 * sugar + rnorm(n, sd = 12)

obs_cor <- cor(sugar, calories)
obs_cor
```

### Step 2: Create the bootstrap distribution

```{r}
#| label: rci-corr-boot
set.seed(8833)

boot_dist <- do_it(10000) * {
  resampled          <- resample_pairs(sugar, calories)
  cor(resampled$vector1, resampled$vector2)
}

hist(boot_dist, breaks = 80,
     main = "Bootstrap Distribution — Correlation",
     xlab = "Correlation",
     col = "steelblue", border = "white")
```

### Step 3: Compute the SE and construct the CI

```{r}
#| label: rci-corr-ci
SE_boot <- sd(boot_dist)

CI <- c(obs_cor - cnorm(0.95) * SE_boot,
        obs_cor + cnorm(0.95) * SE_boot)
CI
```

We are 95% confident that the true correlation between sugar and calorie
content in cereals is between `r round(CI[1], 3)` and `r round(CI[2], 3)`.

::: {.callout-tip}
**Why `resample_pairs()` and not `sample(..., replace = TRUE)`?** For paired
data, we must resample observations together to preserve the relationship
between x and y. Resampling x and y independently would destroy the very
correlation we are trying to measure.
:::

---

## Regression Slope

**When to use:** Two paired quantitative variables; estimate the slope β₁ of
a simple linear regression model. Use `resample_pairs()` to keep pairs intact.

**Example (class 25):** Cigarette consumption per capita and lung cancer rates
across 44 U.S. states. Compute a 95% bootstrap confidence interval for the
slope.

### Step 1: Compute the observed statistic

```{r}
#| label: rci-reg-obs
set.seed(2947)
cigs_per_capita <- runif(44, min = 10, max = 45)
cancer_rate     <- 2 + 0.005 * cigs_per_capita * 1000 + rnorm(44, sd = 5)

lung_lm   <- lm(cancer_rate ~ cigs_per_capita)
obs_slope <- coef(lung_lm)["cigs_per_capita"]
obs_slope
```

### Step 2: Create the bootstrap distribution

```{r}
#| label: rci-reg-boot
set.seed(3109)

boot_dist <- do_it(10000) * {
  resampled        <- resample_pairs(cigs_per_capita, cancer_rate)
  resampled_cigs   <- resampled$vector1
  resampled_cancer <- resampled$vector2
  coef(lm(resampled_cancer ~ resampled_cigs))["resampled_cigs"]
}

hist(boot_dist, breaks = 80,
     main = "Bootstrap Distribution — Regression Slope",
     xlab = "Slope",
     col = "steelblue", border = "white")
```

### Step 3: Compute the SE and construct the CI

```{r}
#| label: rci-reg-ci
SE_boot <- sd(boot_dist)

CI <- c(obs_slope - cnorm(0.95) * SE_boot,
        obs_slope + cnorm(0.95) * SE_boot)
CI
```

We are 95% confident that for each additional hundred cigarettes smoked per
capita, the lung cancer rate increases by between `r round(CI[1], 4)` and
`r round(CI[2], 4)` cases per 100,000 people.

---

## Quick Reference

| Parameter | Resample method | CI formula |
|---|---|---|
| One proportion | `sample(data, n, replace=TRUE)` → `get_proportion()` | `obs ± cnorm(C) * sd(boot_dist)` |
| One mean | `sample(data, n, replace=TRUE)` → `mean()` | `obs ± cnorm(C) * sd(boot_dist)` |
| Diff. of two means | `sample(group, n, replace=TRUE)` for each group | `obs ± cnorm(C) * sd(boot_dist)` |
| Correlation | `resample_pairs(x, y)` → `cor()` | `obs ± cnorm(C) * sd(boot_dist)` |
| Regression slope | `resample_pairs(x, y)` → `coef(lm(...))` | `obs ± cnorm(C) * sd(boot_dist)` |

{{< pagebreak >}}

# Parametric Hypothesis Tests

This guide covers every parametric hypothesis test from S&DS 1000. Each test
follows the same five steps used in class:

1. **State** the null and alternative hypotheses
2. **Calculate** the observed test statistic (and check conditions)
3. **Visualize** the null distribution and mark the observed statistic
4. **Calculate** the p-value
5. **Make a decision**

::: {.callout-warning}
**A note on assumptions.** Parametric tests rely on mathematical
approximations that are only valid when certain conditions are met (e.g.,
large enough sample sizes, approximate normality). Always check conditions
before interpreting a p-value. The randomization methods above are more
robust when conditions are uncertain.
:::

---

## One Proportion

**When to use:** One categorical variable with two outcomes; test whether the
population proportion π equals a specific value π₀.

**Conditions:** $n\pi_0 \geq 10$ and $n(1 - \pi_0) \geq 10$

**Test statistic:**
$$z = \frac{\hat{p} - \pi_0}{SE_0}, \qquad SE_0 = \sqrt{\frac{\pi_0(1 - \pi_0)}{n}}$$

**Example (class 20):** From 1982–1994 there were 128 penalty shots in the
World Cup. Goalkeepers correctly guessed the kick direction 41% of the time.
Test whether goalies guessed correctly **less than** 50% of the time.

### Step 1: State the hypotheses

$$H_0: \pi = 0.50 \qquad H_A: \pi < 0.50$$

### Steps 2–4: Calculate, visualize, p-value

```{r}
#| label: pt-one-prop
pi_0  <- 0.50
n     <- 128
p_hat <- 0.41

# Check conditions
n * pi_0; n * (1 - pi_0)

SE     <- sqrt(pi_0 * (1 - pi_0) / n)
z_stat <- (p_hat - pi_0) / SE
z_stat

plot_norm(col = "blue",
          main = "Null Distribution — Z-statistic",
          xlab = "Z-score", ylab = "Density")
abline(v = z_stat, col = "red", lwd = 2)

p_value <- pnorm(z_stat, mean = 0, sd = 1, lower.tail = TRUE)
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 4)` < 0.05, we **reject** $H_0$ and conclude
that goalkeepers guess the correct direction less than 50% of the time.

---

## Two Proportions

**When to use:** A binary outcome measured in two independent groups; test
whether the two population proportions are equal.

**Conditions:** At least 10 successes and 10 failures in each group.

**Test statistic** (using the unpooled standard error):
$$z = \frac{\hat{p}_1 - \hat{p}_2}{SE}, \qquad
SE = \sqrt{\frac{\hat{p}_1(1-\hat{p}_1)}{n_1} + \frac{\hat{p}_2(1-\hat{p}_2)}{n_2}}$$

**Example:** In a clinical trial, 245 patients received a new vaccine and 255
received a placebo. Of the vaccinated group 32 got the flu; of the placebo
group 56 got the flu. Test whether the vaccine **reduces** the proportion who
get the flu.

### Step 1: State the hypotheses

$$H_0: \pi_1 = \pi_2 \qquad H_A: \pi_1 < \pi_2$$

### Steps 2–4: Calculate, visualize, p-value

```{r}
#| label: pt-two-prop
x1 <- 32;  n1 <- 245   # vaccinated
x2 <- 56;  n2 <- 255   # placebo

p_hat1 <- x1 / n1
p_hat2 <- x2 / n2

# Check conditions
c(x1, n1 - x1, x2, n2 - x2)

SE     <- sqrt((p_hat1*(1-p_hat1)/n1) + (p_hat2*(1-p_hat2)/n2))
z_stat <- (p_hat1 - p_hat2) / SE
z_stat

plot_norm(col = "blue",
          main = "Null Distribution — Z-statistic",
          xlab = "Z-score", ylab = "Density")
abline(v = z_stat, col = "red", lwd = 2)

p_value <- pnorm(z_stat, lower.tail = TRUE)
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 4)` < 0.05, we **reject** $H_0$ and conclude
that the vaccine significantly reduces the proportion of flu cases.

---

## More Than Two Proportions

**When to use:** One categorical variable with three or more outcomes; test
whether the observed counts match a set of expected proportions.

**Conditions:** Expected count in each cell $\geq 5$

**Test statistic:**
$$\chi^2 = \sum_{i=1}^{k} \frac{(O_i - E_i)^2}{E_i}, \qquad df = k - 1$$

**Example (class 23/24):** A Yale professor gathered birth months for 198
randomly selected Yale students. Are students equally likely to be born in
any month?

### Step 1: State the hypotheses

$$H_0: \pi_{Jan} = \pi_{Feb} = \cdots = \pi_{Dec} = \tfrac{1}{12}$$
$$H_A: \text{at least one monthly proportion differs from } \tfrac{1}{12}$$

### Steps 2–4: Calculate, visualize, p-value

```{r}
#| label: pt-chisq
observed_counts <- c(14, 11, 21, 17, 15, 13, 19, 16, 18, 22, 14, 18)
names(observed_counts) <- month.abb
expected_counts <- rep(sum(observed_counts) / 12, 12)

expected_counts[1]   # all ~16.5 — conditions met

chi_sq_stat <- sum((observed_counts - expected_counts)^2 / expected_counts)
chi_sq_stat

df <- length(observed_counts) - 1

plot_chisq(df = df, col = "blue",
           main = "Null Distribution — Chi-squared Statistic",
           xlab = "Chi-squared value", ylab = "Density")
abline(v = chi_sq_stat, col = "red", lwd = 2)

p_value <- pchisq(chi_sq_stat, df = df, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 3)` > 0.05, we **fail to reject** $H_0$. There
is not sufficient evidence that Yale students' birth months deviate from a
uniform distribution.

---

## One Mean

**When to use:** One quantitative variable; test whether the population mean μ
equals a specific value μ₀.

**Conditions:** $n \geq 30$, or the population is approximately normal.

**Test statistic:**
$$t = \frac{\bar{x} - \mu_0}{SE}, \qquad SE = \frac{s}{\sqrt{n}}, \qquad df = n - 1$$

**Example (class 21):** Air Force cadets tested 42 bags of Chips Ahoy! cookies
and found a mean of 1261.6 chips, sd = 117.6. Test whether the mean exceeds
1000.

### Step 1: State the hypotheses

$$H_0: \mu = 1000 \qquad H_A: \mu > 1000$$

### Steps 2–4: Calculate, visualize, p-value

```{r}
#| label: pt-one-mean
n     <- 42
x_bar <- 1261.6
s     <- 117.6
mu_0  <- 1000

SE     <- s / sqrt(n)
t_stat <- (x_bar - mu_0) / SE
t_stat

df <- n - 1
plot_t(df, to = 15, col = "blue",
       main = "Null Distribution — T-statistic",
       xlab = "T-score", ylab = "Density")
abline(v = t_stat, col = "red", lwd = 2)

p_value <- pt(t_stat, df = df, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision

Since p ≈ 0, we **reject** $H_0$ and conclude the average chip count exceeds
1000.

---

## Two Means: Independent Samples

**When to use:** A quantitative variable measured in two independent groups;
test whether the group means are equal.

**Conditions:** Each group has $n \geq 30$, or both populations are
approximately normal.

**Test statistic (Welch's t-test):**
$$t = \frac{(\bar{x}_1 - \bar{x}_2)}{SE_{diff}}, \qquad
SE_{diff} = \sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}, \qquad
df = \min(n_1 - 1,\; n_2 - 1)$$

**Example (class 22):** A study of 2295 American men found that right-handers
(n = 2027) earned \$13.10/hr and left-handers (n = 268) earned \$13.40/hr,
both with sd = \$7.90. Test whether there is a difference in earnings.

### Step 1: State the hypotheses

$$H_0: \mu_1 = \mu_2 \qquad H_A: \mu_1 \ne \mu_2$$

### Steps 2–4: Calculate, visualize, p-value

```{r}
#| label: pt-two-means-indep
n1 <- 2027;  x_bar1 <- 13.10;  s1 <- 7.90
n2 <- 268;   x_bar2 <- 13.40;  s2 <- 7.90

SE_diff <- sqrt((s1^2/n1) + (s2^2/n2))
t_stat  <- (x_bar1 - x_bar2) / SE_diff
t_stat

df <- min(n1 - 1, n2 - 1)
plot_t(df, col = "blue",
       main = "Null Distribution — T-statistic",
       xlab = "T-score", ylab = "Density")
abline(v =  t_stat, col = "red", lwd = 2)
abline(v = -t_stat, col = "red", lwd = 2)

p_value <- 2 * pt(-abs(t_stat), df = df)
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 3)` > 0.05, we **fail to reject** $H_0$. There
is not sufficient evidence of a difference in hourly earnings between right-
and left-handed men.

---

## Two Means: Paired Samples

**When to use:** Observations come in natural pairs (e.g., before/after on the
same subject). Compute within-pair differences and run a one-sample t-test.

**Test statistic:**
$$t = \frac{\bar{d}}{SE_d}, \qquad SE_d = \frac{s_d}{\sqrt{n}}, \qquad df = n - 1$$

**Example (class 22):** Exam scores for 10 students on Exam 1 and Exam 2.
Test whether students scored higher on Exam 2 on average.

### Step 1: State the hypotheses

$$H_0: \mu_d = 0 \qquad H_A: \mu_d > 0$$

### Steps 2–4: Calculate, visualize, p-value

```{r}
#| label: pt-two-means-paired
exam1 <- c(72, 95, 56, 87, 80, 98, 74, 85, 77, 62)
exam2 <- c(78, 96, 72, 89, 80, 95, 86, 87, 82, 72)

d      <- exam2 - exam1
n      <- length(d)
SE_d   <- sd(d) / sqrt(n)
t_stat <- mean(d) / SE_d
t_stat

df <- n - 1
plot_t(df, col = "blue",
       main = "Null Distribution — T-statistic (Paired)",
       xlab = "T-score", ylab = "Density")
abline(v = t_stat, col = "red", lwd = 2)

p_value <- pt(t_stat, df = df, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 4)` < 0.05, we **reject** $H_0$ and conclude
that students scored significantly higher on Exam 2.

---

## More Than Two Means: ANOVA

**When to use:** A quantitative variable measured across three or more
independent groups; test whether all group means are equal.

**Conditions:** Approximately normal residuals; roughly equal variance across
groups; independent observations.

**Test statistic:**
$$F = \frac{MSG}{MSE}, \qquad df_1 = k - 1, \quad df_2 = N - k$$

**Example (class 24):** Hope College students were timed completing a
Sudoku-like puzzle. Majors were grouped into four categories. Is mean
completion time the same across majors?

### Step 1: State the hypotheses

$$H_0: \mu_{as} = \mu_{ns} = \mu_{ss} = \mu_{ah} \qquad
H_A: \text{at least one mean differs}$$

### Steps 2–4: Calculate, visualize, p-value

```{r}
#| label: pt-anova
set.seed(4455)
completion_times <- c(rnorm(10, mean = 22, sd = 5),
                      rnorm(10, mean = 20, sd = 5),
                      rnorm(10, mean = 26, sd = 5),
                      rnorm(10, mean = 21, sd = 5))
majors <- factor(rep(c("Applied Sci", "Natural Sci", "Social Sci", "Arts/Hum"),
                     each = 10))

boxplot(completion_times ~ majors,
        main = "Completion Time by Major",
        ylab = "Time (seconds)",
        col  = c("lightblue", "lightgreen", "lightpink", "lightyellow"))

f_stat <- get_F_stat(completion_times, majors)
df1 <- length(unique(majors)) - 1
df2 <- length(completion_times) - length(unique(majors))
f_stat

plot_f(df1, df2, col = "blue",
       main = "Null Distribution — F-statistic",
       xlab = "F value", ylab = "Density")
abline(v = f_stat, col = "red", lwd = 2)

p_value <- pf(f_stat, df1, df2, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision

```{r}
#| label: pt-anova-decision
#| echo: false
dec <- ifelse(p_value < 0.05,
              "reject H\u2080 — at least one group mean differs",
              "fail to reject H\u2080 — no evidence of a difference")
```

p = `r round(p_value, 3)`: we `r dec`.

---

## Correlation

**When to use:** Two quantitative variables; test whether the linear
association between them is significantly different from zero.

**Conditions:** Both variables approximately normal, or $n$ is large; the
relationship is linear.

**Test statistic:**
$$t = \frac{r\sqrt{n-2}}{\sqrt{1-r^2}}, \qquad df = n - 2$$

**Example (class 25):** Data on cigarette consumption per capita and lung
cancer rates across 44 U.S. states. Is there evidence of a positive linear
association?

### Step 1: State the hypotheses

$$H_0: \rho = 0 \qquad H_A: \rho > 0$$

### Steps 2–4: Calculate, visualize, p-value

```{r}
#| label: pt-correlation
set.seed(2947)
cigs_per_capita <- runif(44, min = 10, max = 45)
cancer_rate     <- 2 + 0.005 * cigs_per_capita * 1000 + rnorm(44, sd = 5)

plot(cigs_per_capita, cancer_rate,
     xlab = "Cigarettes per Capita (hundreds)",
     ylab = "Lung Cancer Rate per 100,000",
     main = "Lung Cancer Rate vs. Cigarette Consumption")

n <- length(cigs_per_capita)
r <- cor(cigs_per_capita, cancer_rate)
r

t_stat <- r * sqrt(n - 2) / sqrt(1 - r^2)
t_stat

df <- n - 2
plot_t(df, col = "blue",
       main = "Null Distribution — T-statistic",
       xlab = "T-score", ylab = "Density")
abline(v = t_stat, col = "red", lwd = 2)

p_value <- pt(t_stat, df = df, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 4)`, we **reject** $H_0$ and conclude that there
is a significant positive linear association between cigarette consumption and
lung cancer rate.

::: {.callout-tip}
**Shortcut:** `cor.test(x, y, alternative = "greater")` performs all of steps
2–4 in one call and returns the same t-statistic, df, and p-value.
:::

---

## Simple Linear Regression

**When to use:** Two quantitative variables where you want to model y as a
linear function of x **and** test whether the slope differs from zero.
Testing H₀: β₁ = 0 is equivalent to testing H₀: ρ = 0 — they yield
identical t-statistics and p-values.

**Conditions:** Linear relationship; approximately normal residuals; constant
variance; independent observations.

**Test statistic for the slope:**
$$t = \frac{\hat{\beta}_1}{SE_{\hat{\beta}_1}}, \qquad df = n - 2$$

**Example (class 25):** Using the same cigarette/lung cancer data, test
whether the slope is significantly positive.

### Step 1: State the hypotheses

$$H_0: \beta_1 = 0 \qquad H_A: \beta_1 > 0$$

### Steps 2–4: Fit the model, visualize, p-value

```{r}
#| label: pt-regression
lung_lm <- lm(cancer_rate ~ cigs_per_capita)
summary(lung_lm)
```

Reading the `Coefficients` table:

| Column | Meaning |
|---|---|
| `Estimate` | Fitted intercept (β̂₀) and slope (β̂₁) |
| `Std. Error` | Standard error of each estimate |
| `t value` | t-statistic testing H₀: coefficient = 0 |
| `Pr(>|t|)` | Two-sided p-value |

```{r}
#| label: pt-regression-viz
fit_sum <- summary(lung_lm)
t_stat  <- fit_sum$coefficients["cigs_per_capita", "t value"]
df      <- lung_lm$df.residual

plot_t(df, col = "blue",
       main = "Null Distribution — T-statistic for Slope",
       xlab = "T-score", ylab = "Density")
abline(v = t_stat, col = "red", lwd = 2)

p_two_sided <- fit_sum$coefficients["cigs_per_capita", "Pr(>|t|)"]
p_value     <- p_two_sided / 2
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 4)` < 0.05, we **reject** $H_0$ and conclude
that there is a significant positive relationship between cigarette
consumption and lung cancer rate.

### Confidence interval for the slope

```{r}
#| label: pt-regression-ci
confint(lung_lm, level = 0.90)
```

---

## Quick Reference

| Test | H₀ | Statistic | df | p-value function |
|---|---|---|---|---|
| One proportion | $\pi = \pi_0$ | $z = \frac{\hat{p}-\pi_0}{SE_0}$ | — | `pnorm(z, lower.tail=...)` |
| Two proportions | $\pi_1 = \pi_2$ | $z = \frac{\hat{p}_1-\hat{p}_2}{SE}$ | — | `pnorm(z, lower.tail=...)` |
| Multiple proportions | $\pi_i = \pi_0$ | $\chi^2 = \sum(O-E)^2/E$ | $k-1$ | `pchisq(x, df, lower.tail=FALSE)` |
| One mean | $\mu = \mu_0$ | $t = \frac{\bar{x}-\mu_0}{s/\sqrt{n}}$ | $n-1$ | `pt(t, df, lower.tail=...)` |
| Two means (indep.) | $\mu_1 = \mu_2$ | $t = \frac{\bar{x}_1-\bar{x}_2}{SE_{diff}}$ | $\min(n_1,n_2)-1$ | `pt(t, df, lower.tail=...)` |
| Two means (paired) | $\mu_d = 0$ | $t = \frac{\bar{d}}{s_d/\sqrt{n}}$ | $n-1$ | `pt(t, df, lower.tail=...)` |
| Multiple means | $\mu_1=\cdots=\mu_k$ | $F = MSG/MSE$ | $k-1$, $N-k$ | `pf(f, df1, df2, lower.tail=FALSE)` |
| Correlation | $\rho = 0$ | $t = \frac{r\sqrt{n-2}}{\sqrt{1-r^2}}$ | $n-2$ | `pt(t, df, lower.tail=...)` |
| Regression slope | $\beta_1 = 0$ | $t = \hat{\beta}_1/SE$ | $n-2$ | `summary(lm(...))` |

{{< pagebreak >}}

# Parametric Confidence Intervals

A **confidence interval** gives a range of plausible values for an unknown
population parameter. A parametric CI uses a mathematical formula based on the
sampling distribution of the estimator. Every CI in this guide takes the form:

$$\text{point estimate} \pm \text{critical value} \times SE$$

The **critical value** comes from a normal or t-distribution and is found using
`cnorm(C)` (normal) or `ct(C, df)` (t) for a confidence level C. The
**standard error (SE)** measures the typical variability of the estimator.

---

## One Proportion

**When to use:** A single categorical variable with two outcomes; estimate the
population proportion π.

**Conditions:** $n\hat{p} \geq 10$ and $n(1-\hat{p}) \geq 10$

**Formula:**
$$\hat{p} \pm z^* \cdot \hat{SE}, \qquad \hat{SE} = \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$

**Example (class 20):** A survey asked 2625 people if they agreed that
"There is only one true love for each person." 1812 of the respondents
disagreed. Compute a 90% confidence interval for the proportion who disagreed.

```{r}
#| label: pci-one-prop
n     <- 2625
x     <- 1812
p_hat <- x / n
p_hat

# Check conditions
n * p_hat; n * (1 - p_hat)

SE     <- sqrt(p_hat * (1 - p_hat) / n)
z_star <- cnorm(0.90)

CI <- c(p_hat - z_star * SE, p_hat + z_star * SE)
CI
```

We are 90% confident that the true proportion who disagree is between
`r round(CI[1], 3)` and `r round(CI[2], 3)`.

---

## One Mean

**When to use:** A single quantitative variable; estimate the population mean μ.

**Conditions:** $n \geq 30$, or the population is approximately normal.

**Formula:**
$$\bar{x} \pm t^* \cdot SE, \qquad SE = \frac{s}{\sqrt{n}}, \qquad df = n-1$$

**Example (class 21):** A study recorded outdoor hunting activity of n = 55
domestic cats. The mean number of kills per week was 2.4 with a standard
deviation of 1.51. Compute a 99% confidence interval for the mean.

```{r}
#| label: pci-one-mean
n     <- 55
x_bar <- 2.4
s     <- 1.51

SE     <- s / sqrt(n)
df     <- n - 1
t_star <- ct(0.99, df)
t_star

CI <- c(x_bar - t_star * SE, x_bar + t_star * SE)
CI
```

We are 99% confident that the mean number of kills per week is between
`r round(CI[1], 3)` and `r round(CI[2], 3)`.

::: {.callout-tip}
**When to use t vs z.** For a single mean, always use the t-distribution
(`ct()`). The t-distribution has heavier tails than the normal, which
accounts for the additional uncertainty from estimating σ with s. For a
proportion, use the normal distribution (`cnorm()`).
:::

---

## Difference of Two Means

**When to use:** A quantitative variable measured in two independent groups;
estimate $\mu_1 - \mu_2$.

**Conditions:** Each group has $n \geq 30$ or both populations are
approximately normal.

**Formula (Welch's):**
$$(\bar{x}_1 - \bar{x}_2) \pm t^* \cdot SE_{diff}, \qquad
SE_{diff} = \sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}, \qquad
df = \min(n_1-1,\, n_2-1)$$

**Example (class 22):** A study recorded dietary fiber intake. Compute a 95%
confidence interval for the difference in mean daily fiber intake (grams)
between males and females.

```{r}
#| label: pci-two-means
set.seed(7823)
fiber_males   <- round(rnorm(155, mean = 23.5, sd = 14.2))
fiber_females <- round(rnorm(662, mean = 20.1, sd = 11.8))

n1 <- length(fiber_males);    x_bar1 <- mean(fiber_males);    s1 <- sd(fiber_males)
n2 <- length(fiber_females);  x_bar2 <- mean(fiber_females);  s2 <- sd(fiber_females)

paste("Males:   n =", n1, " mean =", round(x_bar1,2), " sd =", round(s1,2))
paste("Females: n =", n2, " mean =", round(x_bar2,2), " sd =", round(s2,2))

SE_diff <- sqrt((s1^2/n1) + (s2^2/n2))
df      <- min(n1 - 1, n2 - 1)
t_star  <- ct(0.95, df)

obs_diff <- x_bar1 - x_bar2
CI <- c(obs_diff - t_star * SE_diff, obs_diff + t_star * SE_diff)
CI
```

The 95% CI includes zero, which suggests we do not have strong evidence of a
difference in mean fiber intake between males and females.

---

## Correlation

**When to use:** Two quantitative variables; construct a confidence interval
for the population correlation ρ using **Fisher's z-transformation**.

$$z = \tanh^{-1}(r), \quad SE_z = \frac{1}{\sqrt{n-3}}, \quad
CI_z = z \pm z^* \cdot SE_z, \quad CI_r = \tanh(CI_z)$$

**Example (class 18/25):** Sugar content and calorie content for a sample of
77 breakfast cereals. Compute a 95% confidence interval for the correlation.

```{r}
#| label: pci-corr
set.seed(1123)
n        <- 77
sugar    <- runif(n, 0, 15)
calories <- 90 + 3.5 * sugar + rnorm(n, sd = 12)

corr_result <- cor.test(sugar, calories, conf.level = 0.95)
corr_result$estimate
corr_result$conf.int
```

We are 95% confident that the true correlation is between
`r round(corr_result$conf.int[1], 3)` and
`r round(corr_result$conf.int[2], 3)`.

**Manual calculation using Fisher's z-transformation:**

```{r}
#| label: pci-corr-manual
r    <- cor(sugar, calories)
z    <- atanh(r)
SE_z <- 1 / sqrt(n - 3)

z_star <- cnorm(0.95)
ci_z   <- c(z - z_star * SE_z, z + z_star * SE_z)
tanh(ci_z)
```

---

## Regression Slope

**When to use:** Two quantitative variables; estimate how much y changes on
average for a one-unit increase in x. Use `confint()` on the fitted model,
which uses `qt()` internally with $df = n - 2$.

**Example (class 25):** Cigarette consumption per capita and lung cancer rates
across 44 U.S. states. Compute a 90% confidence interval for the slope.

```{r}
#| label: pci-reg-slope
set.seed(2947)
cigs_per_capita <- runif(44, min = 10, max = 45)
cancer_rate     <- 2 + 0.005 * cigs_per_capita * 1000 + rnorm(44, sd = 5)

lung_lm <- lm(cancer_rate ~ cigs_per_capita)

slope <- coef(lung_lm)["cigs_per_capita"]
slope

confint(lung_lm, level = 0.90)
```

We are 90% confident that for each additional hundred cigarettes smoked per
capita, the lung cancer rate increases by between
`r round(confint(lung_lm, level=0.90)["cigs_per_capita",1], 4)` and
`r round(confint(lung_lm, level=0.90)["cigs_per_capita",2], 4)` cases per
100,000 people.

::: {.callout-tip}
**Manual calculation:** You can also compute the CI by hand using
`summary(lung_lm)`. The SE of the slope is in the `Std. Error` column, and
the critical value is `ct(C, df = n-2)`.
:::

---

## Quick Reference

| Parameter | Formula | Critical value | Function |
|---|---|---|---|
| One proportion π | $\hat{p} \pm z^* \sqrt{\hat{p}(1-\hat{p})/n}$ | `cnorm(C)` | `cnorm()` |
| One mean μ | $\bar{x} \pm t^* \cdot s/\sqrt{n}$ | `ct(C, n-1)` | `ct()` |
| Difference μ₁−μ₂ | $(\bar{x}_1-\bar{x}_2) \pm t^* \cdot SE_{diff}$ | `ct(C, min(n₁,n₂)-1)` | `ct()` |
| Correlation ρ | `cor.test(x, y, conf.level=C)$conf.int` | Fisher z-transform | — |
| Regression slope β₁ | `confint(lm(...), level=C)` | `ct(C, n-2)` | — |
