---
title: "Practice Session 9 "
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---


# Introduction 

The **Central Limit Theorem** for Sample Means indicates that if the sample size $n$ is large, the distribution of sample means from a population with mean $\mu$ and standard deviation $\sigma$ is approximately normally distributed with mean $\mu$ and standard deviation $\sigma / \sqrt{n}$.

Unfortunately, we find two difficulties when trying to use this result in practice. First, we generally don't know the standard deviation $\sigma$. This means that we need to use the standard deviation from the sample, $s$, when estimating the standard error of the sample means.

$$
SE=\frac{s}{\sqrt{n}}
$$

The second difficulty is that when we use the $SE = \frac{s}{\sqrt{n}}$ to estimate the standard error of the sampling distribution of the means, the distribution of our test statistic is no longer normal. The resulting distribution that our test statistic follows is known as a **t-distribution** with **degrees of freedom** $df = n-1$. 

$\\$

## 1. Single Mean

Let $\bar{x}$ be the sample mean, and $s$ be the sample standard deviation. Let $t^*$ be the critical value from the t-distribution with $df=n-1$ that corresponds to our confidence level and/or significance level. Suppose that the distribution of the population is approximately normal and/or the sample size is sufficiently large.

* **CI for a One Mean**

The confidence interval for the population mean, $\mu$, is given by:

$$
\bar{x} \pm t^* \cdot \frac{s}{\sqrt{n}}
$$

* **Hypothesis Test for One Mean: T-Test**

To test $H_{\mathrm{0}}: \mu=\mu_{\mathrm{0}}$ vs $H_a: \mu \neq \mu_{\mathrm{0}}$ (or a one-tail alternative), we compute the following test statistic:

$$
t=\frac{\text { Statistic }- \text { Null value }}{S E}=\frac{\bar{x}-\mu_0}{s / \sqrt{n}}
$$

where the p-value of the test is computed using the appropriate tail(s) of a t-distribution with $n-1$ degrees of freedom.

$\\$

# 2. Inference for Difference of Two Means

Consider two populations. Assume each population is approximately normally distributed, or that both of our samples are sufficiently large. Let $\bar{x}_1$ and $\bar{x}_2$ be the sample means means, $s_1$ and $s_2$ be the sample standard deviations, and $n_1$ and $n_2$ be the sample sizes, respectively. 

 * **Confidence Interval for Difference of Two Means**
 
The confidence interval for the difference of two means, $\mu_1 - \mu_2$, is given by:

$$
\left(\bar{x}_1-\bar{x}_2\right) \pm t^* \cdot \sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}
$$

where **$t^*$** is an endpoint chosen from a t-distribution with df = $\min (n_1-1, n_2-1)$.

* **Hypothesis Test for Difference of Two Means (Two-Sample t-Test )**

To test $H_{\mathrm{O}}: \mu_1=\mu_2$ vs $H_a: \mu_1 \neq \mu_2$ (or one-tail test), we use two-sample t -statistic: 

$$
t=\frac{\text { Statistic }- \text { Null value }}{S E}=\frac{\left(\bar{x}_1-\bar{x}_2\right)-0}{\sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}}
$$
we use a t -distribution to find the p-value  with df = min ( $n_1-1$, $n_2-1$ ). 


$\\$


# 3. Inference for Difference of Means with **Paired Data**

* **CI for Difference of Means with Paired Data** 

To estimate the difference in means based on paired data, we first subtract to compute the difference for each data pair and compute the mean $\bar{x}_d$, the standard deviation $s_d$, and the sample size $n_d$ for the sample differences. As before, we need the differences to be approximately normally distributed (or the sample size to be sufficiently large).

The confidence interval for the difference in means is given by:

$$
\text { Statistic } \pm t^* \cdot S E=\bar{x}_d \pm t^* \frac{s_d}{\sqrt{n_d}}
$$

where $t^*$ is a percentile from a t-distribution with $df = n_d - 1$.


* **Hypothesis Test for Difference of Means with Paired Data** 

To test $H_{\mathrm{o}}: \mu_d=\mathrm{o}$ vs $H_a: \mu_d \neq \mathrm{o}$ (or one-tail test) we use the t-test statistic:

$$
t=\frac{\text { Statistic }- \text { Null value }}{S E}=\frac{\bar{x}_d-0}{s_d / \sqrt{n_d}}
$$

If the differences are reasonably normally distributed (or the sample size is sufficiently large), we can use a t-distribution with $n_d-\mathbf{1}$ degrees of freedom to compute a p-value for this test statistic.


$\\$



## Exercise 1: Single Sample T-Test


According to a local statistics instructor, the average score on his first exam of the semester (`Exam1`) is historically around 80%. The data is available in the `StatsGrades` from the `Lock5Data` library. 

Run a `one sample t-test` to see if this year's students performed significantly better than the historic average.

```{r}
library(SDS1000)
library(Lock5Data)
data(StatGrades)
```


a. Visualize the data using a boxplot and a histogram. Do you see any outliers?

```{r}
# Your code here
```


b. Step 1: State the null and alternative hypotheses using symbols.





c. Step 2: Calculate the sample statistic.

```{r}
# Your code here
```




d. Step 3: Calculate the t-test statistic

```{r}
# Your code here
```



e. Step 4: Calculate the p-value of the test 

```{r}
# Your code here
```


f. Step 5: State your conclusion.




g.  Use the `dt()` function and `plot()` to visualize the t-distribution. The degrees of freedom will be the same as for the test. Add a vertical red line to the plot at the observed sample mean.

```{r}
# Your code here
```



$\\$



# Exercise 2: Two Sample t-test

The same local statistics instructor believes that students perform  differently on the two exams, `Exam1` , and  `Exam2`.  Again, the data is available in the `StatsGrades`  from the `Lock5Data` library. Visualize the data of `Exam2`.


```{r}
library(SDS1000)
library(Lock5Data)
data(StatGrades)
```


```{r}
# Your code here
```


a. Is this research question could be answered with t-test for two independent samples, or, t- test for paired data? 

  
  

b. Step1: State the null and alternative hypotheses using symbols.





c. Step 2: Calculate the statistic of interest. 


```{r}
# Your code here
```



d. Step 3: Calculate the t-statistic

```{r}
# Your code here
```



e. Step 4: Calculate the p-value of the test

```{r}
# Your code here
```


f. Step 5: State your conclusion.




g. Use the `dt()` function and `plot()` to visualize the t-distribution. The degrees of freedom will be the same as for the test. Add a vertical line to the plot at the observed statistic.

$\\$

## Excersice 3 :  Mental Imaging and Muscle Fatigue

A new study explores how actual muscle fatigue affects mental imaging. Participants were asked to either perform `Actual` pointing motions or to `Mental` imagining doing the motions. Eight participants were assigned to each group (mental vs actual), and the time in seconds to complete the motions is given in the data set `MentalMuscle`. The participant's ability to perform motions (mental or actual) was measured pre- and post-fatigue. Participants developed muscle fatigue by holding a heavy weight outwards for as long as they could. After becoming fatigued, they were asked to repeat the previous mental or actual motions. Use a 5% significance level for the tests.

*Note*: Data is approximated from this research paper.
Demougeot, L., and Papaxanthis, C., *Muscle Fatigue Affects Mental Simulation of Action*, The Journal of Neuroscience, 2011;31(29): 10712–20. [Link](https://www.jneurosci.org/content/31/29/10712).

```{r}
library(SDS1000)
library(Lock5Data)
data(MentalMuscle)

# clean the data from "NA"
mental_muscule <- na.omit(MentalMuscle)
```
 
1. Test to see whether there is a significant difference in mean times between mentally doing the actions and actually doing the actions before any muscle fatigue (pre-fatigue). 

a. Is this t-test for paired difference in means, or a t-test for two means for two independent samples. 



b. Step 1:  State the null and alternative hypotheses.





c. Step 2: Calculate the statistic of interest. 

```{r}
# Your code here
```

d. Step 3: Calculate the t-statistic

```{r}
# Your code here
```



e. Step 4: Calculate the p-value of the test

```{r}
# Your code here
```


f. Step 5: Conclusion



g. Use the `dt()` function and `plot()` to visualize the t-distribution. The degrees of freedom will be the same as for the test. Add a vertical line to the plot at the observed statistic.

```{r}
# Your code here
```

$\\$


2. Now test to see whether people who mentally perform the motions are faster, on average, at `mentally imaging` the actions `post-fatigue` than `pre-fatigue`.
Is this t-test for paired difference in means, or a t-test difference in two means for two independent samples. 




a. Step 1:  State the null and alternative hypotheses 


 

b. Step 2:  Calculated the observed statistic

```{r}
# Your code here
```


c. Step 3:  Calculate the test statistic 

```{r}
# Your code here
```



d. Step 4: Find the p-value

```{r}
# Your code here
```


e. Step 5: Conclusion




f. Use the `dt()` function and `plot()` to visualize the t-distribution. Add a vertical line to the plot at the observed statistic.

```{r}
# Your code here
```

$\\$




# Extra Exercises 

## Exercise 4: 

How many more (or fewer) free throw attempts do the Warriors tend to get (on average) at home games compared to their road games. 

```{r}
library(SDS1000)
library(Lock5Data)
data(GSWarriors2019)
```
 
a. Find a 95% confidence interval for the difference in mean free throw attempts.

This question is asking about an estimate and confidence interval for the difference in means, $\mu_H-\mu_A$, where now $\mu$ is the mean number of free throw attempts in a game.



$\\$
