---
title: "Practice Session 9 Answers"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---


# Introduction 

The **Central Limit Theorem** for Sample Means indicates that if the sample size $n$ is large, the distribution of sample means from a population with mean $\mu$ and standard deviation $\sigma$ is approximately normally distributed with mean $\mu$ and standard deviation $\sigma / \sqrt{n}$.

Unfortunately, we find two difficulties when trying to use this result in practice. First, we generally don't know the standard deviation $\sigma$. This means that we need to use the standard deviation from the sample, $s$, when estimating the standard error of the sample means.

$$
SE=\frac{s}{\sqrt{n}}
$$

The second difficulty is that when we use the $SE = \frac{s}{\sqrt{n}}$ to estimate the standard error of the sampling distribution of the means, the distribution of our test statistic is no longer normal. The resulting distribution that our test statistic follows is known as a **t-distribution** with **degrees of freedom** $df = n-1$. 

$\\$

## 1. Single Mean

Let $\bar{x}$ be the sample mean, and $s$ be the sample standard deviation. Let $t^*$ be the critical value from the t-distribution with $df=n-1$ that corresponds to our confidence level and/or significance level. Suppose that the distribution of the population is approximately normal and/or the sample size is sufficiently large.

* **CI for a One Mean**

The confidence interval for the population mean, $\mu$, is given by:

$$
\bar{x} \pm t^* \cdot \frac{s}{\sqrt{n}}
$$

* **Hypothesis Test for One Mean: T-Test**

To test $H_{\mathrm{0}}: \mu=\mu_{\mathrm{0}}$ vs $H_a: \mu \neq \mu_{\mathrm{0}}$ (or a one-tail alternative), we compute the following test statistic:

$$
t=\frac{\text { Statistic }- \text { Null value }}{S E}=\frac{\bar{x}-\mu_0}{s / \sqrt{n}}
$$

where the p-value of the test is computed using the appropriate tail(s) of a t-distribution with $n-1$ degrees of freedom.

$\\$

# 2. Inference for Difference of Two Means

Consider two populations. Assume each population is approximately normally distributed, or that both of our samples are sufficiently large. Let $\bar{x}_1$ and $\bar{x}_2$ be the sample means means, $s_1$ and $s_2$ be the sample standard deviations, and $n_1$ and $n_2$ be the sample sizes, respectively. 

 * **Confidence Interval for Difference of Two Means**
 
The confidence interval for the difference of two means, $\mu_1 - \mu_2$, is given by:

$$
\left(\bar{x}_1-\bar{x}_2\right) \pm t^* \cdot \sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}
$$

where **$t^*$** is an endpoint chosen from a t-distribution with df = $\min (n_1-1, n_2-1)$.

* **Hypothesis Test for Difference of Two Means (Two-Sample t-Test )**

To test $H_{\mathrm{O}}: \mu_1=\mu_2$ vs $H_a: \mu_1 \neq \mu_2$ (or one-tail test), we use two-sample t -statistic: 

$$
t=\frac{\text { Statistic }- \text { Null value }}{S E}=\frac{\left(\bar{x}_1-\bar{x}_2\right)-0}{\sqrt{\frac{s_1^2}{n_1}+\frac{s_2^2}{n_2}}}
$$
we use a t -distribution to find the p-value  with df = min ( $n_1-1$, $n_2-1$ ). 


$\\$


# 3. Inference for Difference of Means with **Paired Data**

* **CI for Difference of Means with Paired Data** 

To estimate the difference in means based on paired data, we first subtract to compute the difference for each data pair and compute the mean $\bar{x}_d$, the standard deviation $s_d$, and the sample size $n_d$ for the sample differences. As before, we need the differences to be approximately normally distributed (or the sample size to be sufficiently large).

The confidence interval for the difference in means is given by:

$$
\text { Statistic } \pm t^* \cdot S E=\bar{x}_d \pm t^* \frac{s_d}{\sqrt{n_d}}
$$

where $t^*$ is a percentile from a t-distribution with $df = n_d - 1$.


* **Hypothesis Test for Difference of Means with Paired Data** 

To test $H_{\mathrm{o}}: \mu_d=\mathrm{o}$ vs $H_a: \mu_d \neq \mathrm{o}$ (or one-tail test) we use the t-test statistic:

$$
t=\frac{\text { Statistic }- \text { Null value }}{S E}=\frac{\bar{x}_d-0}{s_d / \sqrt{n_d}}
$$

If the differences are reasonably normally distributed (or the sample size is sufficiently large), we can use a t-distribution with $n_d-\mathbf{1}$ degrees of freedom to compute a p-value for this test statistic.


$\\$



## Exercise 1: Single Sample T-Test


According to a local statistics instructor, the average score on his first exam of the semester (`Exam1`) is historically around 80%. The data is available in the `StatsGrades` from the `Lock5Data` library. 

Run a `one sample t-test` to see if this year's students performed significantly better than the historic average.

```{r}
library(SDS1000)
library(Lock5Data)
data(StatGrades)
```


a. Visualize the data using a boxplot and a histogram. Do you see any outliers?

```{r}
par(mfrow = c(1, 2))
exam1 <- StatGrades$Exam1

boxplot(exam1,
        main= " boxplt of Exam 1",
        ylab = "Exam 1 Scores")

hist(exam1, xlab = "Exam 1 Scores")
```

There appears to be one outlier. We might want to investigate this score further.




b. Step 1: State the null and alternative hypotheses using symbols.

* $H_0: \mu = 80$
* $H_A: \mu \neq 80$




c. Step 2: Calculate the sample statistic.

```{r}

# compute sample mean
mean_exam1 <-  mean(exam1) 

```




d. Step 3: Calculate the t-test statistic

```{r}
# first compute standard error

# get sample size (n = 50)
n1 <- length(exam1) 

# sd of the exam1 
sd_exam1 <-  sd(exam1) 
se_exam1 <-  sd_exam1 / sqrt(n1) 

# t-test statistic
t_stat_exam1 <- (mean_exam1 - 80) / se_exam1
t_stat_exam1

```




e. Step 4: Calculate the p-value of the test 

```{r}
# df = 50 - 1 = 49
pt(t_stat_exam1, df = n1 - 1, lower.tail = F)
```



f. Step 5: State your conclusion.

Since our p-value is greater than 0.05, we will **fail to reject** the null hypothesis. We therefore do not have evidence that student's exam scores are significantly higher than 80.





g.  Use the `dt()` function and `plot()` to visualize the t-distribution. The degrees of freedom will be the same as for the test. Add a vertical red line to the plot at the observed sample mean.

```{r}
# Create sequence of t-values
t_values <-  seq(-4,4, length.out = 1000)
dt_values <-  dt(t_values, df = 49) # convert to density values

# plot and make line plot with type = "l"
plot(t_values, dt_values, type = "l") 
abline(v = t_stat_exam1, col = "blue") # add vertical red line
```

It seems that the sample mean wasn't very far from the null value of 80, which supports the lack of significance in our test.



$\\$



# Exercise 2: Two Sample t-test

The same local statistics instructor believes that students perform  differently on the two exams, `Exam1` , and  `Exam2`.  Again, the data is available in the `StatsGrades`  from the `Lock5Data` library. Visualize the data of `Exam2`.


```{r}
library(SDS1000)
library(Lock5Data)
data(StatGrades)
```

```{r}
par(mfrow = c(1, 2))
exam2 <- StatGrades$Exam2

diff_ex1_ex2 <- exam2-exam1
boxplot(exam2-exam1,
        main= " boxplt of Exam 2-Exam2",
        ylab = "Exam 2- Exam1 Scores")
hist(exam2, xlab = "Exam 2-Exam1  Scores")

```


a. Is this research question could be answered with t-test for two independent samples, or, t- test for paired data? 


This is t- test for paired data.
  
  
  

b. Step1: State the null and alternative hypotheses using symbols.

Let  $D$ be the true difference between the exam 1 and exam 2 scores. We will perform the following two-sided test:

-   $H_0: D = 0$
-   $H_A: D \neq 0$




c. Step 2: Calculate the statistic of interest. 

```{r}
# compute mean difference 
exam1<- StatGrades$Exam1
exam2<- StatGrades$Exam2

diff_exams <- exam2 - exam1
mean_diff <-  mean(diff_exams)
```




d. Step 3: Calculate the t-statistic

```{r}

# find the sample size 
n_diff <-  length(diff_exams)

# find the SE of the difference 
se_diff <-  sd(diff_exams) / sqrt(n_diff)

# find the t-test statistics for the difference between the two exams
t_stat_diff <-  mean_diff / se_diff

```




e. Step 4: Calculate the p-value of the test

Since we are doing a two-sided test, we need to multiply the p-value by 2.

```{r}
pv  <- 2 * pt(t_stat_diff, df = 50 - 1, lower.tail = F)
```



f. Step 5: State your conclusion.

Since our p-value is less than 0.05, we will reject the null hypothesis. We therefore have evidence that students performed differently on the two exams.

$\\$

## Excersice 3 :  Mental Imaging and Muscle Fatigue

A new study explores how actual muscle fatigue affects mental imaging. Participants were asked to either perform `Actual` pointing motions or to `Mentally` imagine doing the motions. Eight participants were assigned to each group (mental vs actual), and the time in seconds to complete the motions is given in the data set `MentalMuscle`. The participant's ability to perform motions (mental or actual) was measured pre- and post-fatigue. Participants developed muscle fatigue by holding a heavy weight outwards for as long as they could. After becoming fatigued, they were asked to repeat the previous mental or actual motions. Use a 5% significance level for the tests.

*Note*: Data is approximated from this research paper.
Demougeot, L., and Papaxanthis, C., *Muscle Fatigue Affects Mental Simulation of Action*, The Journal of Neuroscience, 2011;31(29): 10712–20. [Link](https://www.jneurosci.org/content/31/29/10712).

```{r}

library(SDS1000)
library(Lock5Data)
data(MentalMuscle)

 # clean the data from "NA"
mental_muscule <- na.omit(MentalMuscle)
```
 
1. Test to see whether there is a significant difference in mean times between mentally doing the actions and actually doing the actions before any muscle fatigue (pre-fatigue). 

a. Is this t-test for paired difference in means, or a t-test for two means for two independent samples. 

* This is a t-test for difference in two means for two independent samples, the `Actual` motions vs the `Mental`  imagining`. 


1.b Step 1:  State the null and alternative hypotheses.

 * To compare the mean mental and actual times before fatigue we do a two-sample t-test for difference in means. Letting $\mu_M$ represent the mean time for someone mentally imaging the actions before any muscle fatigue and $\mu_A$ represent the mean time for someone actually performing the actions before any muscle fatigue, our hypotheses are:

$$
\begin{array}{ll}
H_0: & \mu_M- \mu_A =0  \\
H_a: & \mu_M - \mu_A \neq 0
\end{array}
$$

1.c Step 2: Calculate the statistic of interest. 

```{r}
pre_fatigue_mental <- mental_muscule$PreFatigue[mental_muscule$Action == "Mental"]
mean(pre_fatigue_mental)

pre_fatigue_actual <- mental_muscule$PreFatigue[mental_muscule$Action == "Actual"]
mean( pre_fatigue_actual)

obs_stat <- mean(pre_fatigue_mental) - mean( pre_fatigue_actual)
obs_stat
```


d. Step 3: Calculate the t-statistic

```{r}

# find the samples sizes 
n_m <-  length(pre_fatigue_mental)
n_a <-  length(pre_fatigue_actual)


# find the sd for he two samples of the difference 
sd_m <-  sd(pre_fatigue_mental)
sd_a <- sd(pre_fatigue_actual)


# find the t-test statistics for the difference between the two exams
t_stat_pre_mental_actual <-  (obs_stat- 0) / sqrt (( sd_m^2/n_m )+ (sd_a^2/n_a))

```


$$
t=\frac  {(\bar{x}_M-\bar{x}_A
)- \mu_0}{\sqrt{\frac{s_M^2}{n_M}+\frac{s_A^2}{n_A}}}=\frac{(7.34-7.16)- 0}{\sqrt{\frac{1.22^2}{8}+\frac{0.70^2}{8}}}=0.3520491
$$


e.  Step 4: Calculate the p-value of the test


```{r}
pv <- 2 *pt ( 0.3520491, df= 8-1,  lower.tail =F)
pv
```

This is a two-tail test, so the p-value is twice the area above 0.3520 in a t -distribution with $df = 7$. 

f- Step 5: Conclusion

We see that the p-value is $0.73$. This is a very large p-value, so we do not reject $H_0$. There is no convincing evidence at all of a difference between the two groups before muscle fatigue.

```{r}
# Create sequence of t-values
t_values <-  seq(-4,4, length.out = 1000)
dt_values <-  dt(t_values, df = 7) # convert to density values

# plot and make line plot with type = "l"
plot(t_values, dt_values, type = "l") 
abline(v= c(t_stat_pre_mental_actual, - t_stat_pre_mental_actual) , col = "green2") # add vertical red line
```

2. Now test to see whether people who mentally perform the motions are faster, on average, at `mentally imaging` the actions `post-fatigue` than `pre-fatigue`.
Is this t-test for paired difference in means, or a t-test difference in two means for two independent samples. 


**Summarized Answers***

* Since the same 8 people under the the `mental` group  perform the movements twice, once before muscle fatigue and once after. To compare the pre-fatigue and post-fatigue means for those doing mental movements we use a paired difference in means test. 

We compute the differences, `D = PostFatigue − PreFatigue`, using the data for the people doing the `mental imaging`.


a. Step 1:  State the null and alternative hypotheses
The hypotheses are:

$$
\begin{array}{ll}
H_0: & \mu_D = 0 \\
H_a: & \mu_D < 0
\end{array}
$$

Notice that if we had subtracted the other direction (PreFatigue - PostFatigue), the alternative would be in the other direction and all the differences would have the opposite sign, but the results would be identical. 

b. Step 2:  Calculated the observed statistic

```{r}

# compute mean difference 
pre_fatigue_mental <- mental_muscule$PreFatigue[mental_muscule$Action == "Mental"]
post_fatigue_mental <- mental_muscule$PostFatigue[mental_muscule$Action == "Mental"]

diff_mental_post_pres <- post_fatigue_mental - pre_fatigue_mental
diff_mental_post_pres
obs_diff_mental <- mean(diff_mental_post_pres)
obs_diff_mental  

```



c. Step 3:  Calculate the test statistic 

```{r}

n <- 8 
# first find SD of the difference between the post and pre in mental data
sd_diff_mental <- sd(diff_mental_post_pres )
sd_diff_mental
t_test_diff_mental <- obs_diff_mental / (sd_diff_mental/sqrt(n))
t_test_diff_mental 
```


$$
t=\frac{\bar{x}_D}{s_D / \sqrt{n_D}}=\frac{-1.2375}{1.538958 / \sqrt{8}}=-2.2743
$$
d. Step 4: Find the p-value


This is a lower-tail test, so the p-value is the area below -2.28 in a t -distribution with $d f=7$. 
We could find the p-value is 0.028 .

```{r}

pv <- pt ( -2.2743, df= 7,  lower.tail =T)
pv
```


e. Step 5: Conclusion

At a $5 \%$ significance level, we reject $H_0$ and conclude that people are faster, on average, at mentally imaging physical motions when they have muscle fatigue.


This is the new finding from the study: When people have muscle fatigue, they speed up their mental imaging - presumable because they just want to get done! It is likely that this makes the mental imagery less effective.


```{r}
# Create sequence of t-values
t_values <-  seq(-4,4, length.out = 1000)
dt_values <-  dt(t_values, df = 7) # convert to density values

# plot and make line plot with type = "l"
plot(t_values, dt_values, type = "l") 
abline(v= t_test_diff_mental , col = "yellow4") # add vertical red line
```


$\\$




# Extra Exercises 

## Exercise 4: 

How many more (or fewer) free throw attempts do the Warriors tend to get (on average) at home games compared to their road games. 

```{r}
library(SDS1000)
library(Lock5Data)
data(GSWarriors2019)
```
 
a. Find a 95% confidence interval for the difference in mean free throw attempts.

This question is asking about an estimate and confidence interval for the difference in means, $\mu_H-\mu_A$, where now $\mu$ is the mean number of free throw attempts in a game.

The sample sizes are both more than 30 and the $FTA$ amounts have no outliers, so we use a t-distribution with $41-1=40$ degrees of freedom to find $t^*=2.021$ for $95 \%$ confidence. We compute the confidence interval using

$$
\begin{aligned}
\left(\bar{x}_H-\bar{x}_A\right) & \pm t^* \cdot \sqrt{\frac{s_H^2}{n_H}+\frac{s_A^2}{n_A}} \\
(20.61-20.17) & \pm 2.021 \cdot \sqrt{\frac{5.95^2}{41}+\frac{7.86^2}{41}} \\
0.44 & \pm 3.11 \\
-2.67 & \text { to } 3.55
\end{aligned}
$$


Based on these results we are $95 \%$ sure that the Warriors average between 2.67 fewer and 3.55 more free throw attempts (per game) when playing games at home versus games on the road.


$\\$
