---
title: "Class 17 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---





## Part 1: Hypothesis tests for more than two means


A group of Hope College students wanted to see if there was an association
between a student’s major and the time it takes to complete a small Sudoku-like
puzzle


They grouped majors into four categories

* Applied science (as)
* Natural science (ns)
* Social science  (ss)
* Arts/humanities  (ah)

Let's run a hypothesis test to see if there is an association between a student's
major and the time it takes to complete the puzzle.



$\\$



### Step 1: State the null and alternative hypotheses



Words:

* Null hypothesis: 

* Alternative hypothesis: 



Symbols:  







$\\$




### Step 2a: Visualize the data

```{r sudoku_viz}

# load the data
sudoku_data <- read.table("MajorPuzzle.txt", header = TRUE)

completion_times <- sudoku_data$time
majors <- sudoku_data$major


# visualize the data using side-by-side box plots






```



**Question**: From looking at the visualization you created, does it appear that
the mean time to complete the puzzle is different for at least one of the four
groups?


**Answer**: 








$\\$




### Step 2b: Calculate the observed statistic


```{r sudoku_obs_stat}

library(SDS1000)


# calculate the observed statistic





```





$\\$





### Step 3: Create the null distribution


```{r sudoku_null_distribution}

library(SDS1000)


# Create the null distribution












# plot the null distribution







```




$\\$




### Step 4: Calculate the p-value


```{r sudoku_p_value}






```




$\\$




### Step 5: Make decision









$\\$








## Part 2: Hypothesis tests for correlation 


Let's now look at hypothesis tests for correlation by examining whether cereals
with more sugar tend to have more calories.




### Step 1: State the null and alternative hypotheses



Words:

* Null hypothesis: 

* Alternative hypothesis: 





Symbols:  








$\\$




### Step 2a: Visualize the data

```{r cereal_viz}

# load the data
load("cereal.Rda")


sugar <- cereal$Sugars
calories <- cereal$Calories




# visualize the data using a scatter plot






```



**Question**: From looking at the visualization you created, does it appear that
there is a positive correlation between sugar content and calorie content in cereals?



**Answer**: 





$\\$




### Step 2b: Calculate the observed statistic


```{r cereal_obs_stat}

# calculate the observed statistic





```



$\\$



### Step 3: Create the null distribution


```{r cereal_null_distribution}

library(SDS1000)


# create the null distribution












# plot the null distribution






```




$\\$




### Step 4: Calculate the p-value


```{r cereal_p_value}





```




$\\$





### Step 5: Make decision








$\\$






