---
title: "Class 17 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---





## Part 1: Hypothesis tests for more than two means


A group of Hope College students wanted to see if there was an association
between a student’s major and the time it takes to complete a small Sudoku-like
puzzle


They grouped majors into four categories

* Applied science (as)
* Natural science (ns)
* Social science  (ss)
* Arts/humanities  (ah)

Let's run a hypothesis test to see if there is an association between a student's
major and the time it takes to complete the puzzle.



$\\$



### Step 1: State the null and alternative hypotheses



Words:

* Null hypothesis: The mean time to complete the puzzle is the same for all four groups  

* Alternative hypothesis: The mean time to complete the puzzle is not the same 
for all four groups  


Symbols:  

* $H_0: \mu_{as} = \mu_{ns} = \mu_{ss} = \mu_{ah}$   
* $H_A$: At least one of the means is different  




$\\$




### Step 2a: Visualize the data

```{r sudoku_viz}

# load the data
sudoku_data <- read.table("MajorPuzzle.txt", header = TRUE)

completion_times <- sudoku_data$time
majors <- sudoku_data$major

# visualize the data using side-by-side box plots
boxplot(completion_times ~ majors, 
        main = "Time to complete puzzle by major",
        xlab = "Major",
        ylab = "Time (seconds)",
        col = c("lightblue", "lightgreen", "lightpink", "lightyellow"))

```



**Question**: From looking at the visualization you created, does it appear that
the mean time to complete the puzzle is different for at least one of the four
groups?


**Answer**: From my visualization it appears that the mean time to complete the
puzzle is different for at least one of the four groups. The box plot is showing
that the median time for the natural science group is lower than the median times
for the other three groups. However, there is also a lot of overlap in the times
for the four groups. So, it is hard to tell just by looking at the box plot if
the mean time to complete the puzzle is actually different for at least one of
the four groups.




$\\$




### Step 2b: Calculate the observed statistic


```{r sudoku_obs_stat}

library(SDS1000)


# calculate the observed statistic

obs_stat <- get_MAD_stat(completion_times, majors)

obs_stat

```



$\\$



### Step 3: Create the null distribution


```{r sudoku_null_distribution}

library(SDS1000)


# Create the null distribution

null_dist <- do_it(10000) * {

  # shuffle the completion times
  shuffled_majors <- shuffle(majors)

  # calculate the simulated statistic
  get_MAD_stat(completion_times, shuffled_majors)

}

# plot the null distribution
hist(null_dist, breaks = 200,
     main = "Null distribution",
     xlab = "Mean absolute deviation of group means")


# add a red vertical line at the observed statistic
abline(v = obs_stat, col = "red")

```




$\\$




### Step 4: Calculate the p-value


```{r sudoku_p_value}

p_value <- pnull(obs_stat, null_dist, lower.tail = FALSE)

p_value

```




$\\$




### Step 5: Make decision

Since the p-value is greater than the typical cutoff of 0.05, we fail to reject
the null hypothesis. Thus we can't conclude that the mean time to complete the
puzzle is different for the different majors.


$\\$








## Part 2: Hypothesis tests for correlation 


Let's now look at hypothesis tests for correlation by examining whether cereals
with more sugar tend to have more calories.




### Step 1: State the null and alternative hypotheses



Words:

* Null hypothesis: There is no correlation between sugar content and calorie content in cereals

* Alternative hypothesis: There is a positive correlation between sugar content and calorie content in cereals.  


Symbols:  

* $H_0: \rho = 0$
* $H_A$: $\rho > 0$



$\\$




### Step 2a: Visualize the data

```{r cereal_viz}

# load the data
load("cereal.Rda")


sugar <- cereal$Sugars
calories <- cereal$Calories


# visualize the data using a scatter plot
plot(sugar, calories, 
     main = "Calories vs Sugar in Cereals",
     xlab = "Sugar (grams)",
     ylab = "Calories")



```



**Question**: From looking at the visualization you created, does it appear that
there is a positive correlation between sugar content and calorie content in cereals?




**Answer**: From my visualization it appears that there is a positive correlation between
 sugar content and calorie content in cereals. The scatter plot shows an upward trend,
indicating that as the sugar content increases, the calorie content also tends to increase.





$\\$




### Step 2b: Calculate the observed statistic


```{r cereal_obs_stat}

# calculate the observed statistic

obs_stat <- cor(sugar, calories)

obs_stat


```



$\\$



### Step 3: Create the null distribution


```{r cereal_null_distribution}

library(SDS1000)


null_dist <- do_it(10000) * {

  # shuffle the sugar content
  shuffled_sugar <- shuffle(sugar)

  # calculate the simulated statistic
  cor(shuffled_sugar, calories)

}


# plot the null distribution
hist(null_dist, breaks = 200,
     main = "Null distribution",
     xlab = "Correlation between sugar and calories")

# add a red vertical line at the observed statistic
abline(v = obs_stat, col = "red")


```




$\\$




### Step 4: Calculate the p-value


```{r cereal_p_value}

p_value <- pnull(obs_stat, null_dist, lower.tail = FALSE)

p_value

```




$\\$




### Step 5: Make decision

Since the p-value is much less than the typical cutoff of 0.05, we reject the
null hypothesis, and we conclude that there is a positive correlation between
sugar content and calorie content in cereals.




$\\$






