---
title: "Homework 5"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---




$\\$



In the following homework, you will gain more experience examining sampling
distributions, confidence intervals, and constructing confidence intervals using
the bootstrap. Please submit a compiled pdf with your answers to the exercises
to Gradescope *by 11pm on Sunday October 11th*.
**Points will be taken off if you do not mark the pages for each problem on Gradescope.**

A list of functions we have used in class can be found on Canvas. You might find
the following symbols useful:  $\mu$, $\bar{x}$, $\pi$, $\hat{p}$, $\rho$,
$\hat{y}$, $\pm$, and $\approx$, as well as the functions `cnorm()`, `do_it()`
and `sample()`. For full credit on problems involving R, be sure to label all
your figures and to "show your work" by making sure all values that are answers
to questions are printed/visible in your Quarto pdf.

As always, if you need help with any of the homework assignments, please attend
the TA office hours which are listed on Canvas and/or ask questions on Ed
Discussions forum. Also, if you have completed the homework, please help others
out by answering questions on Ed Discussions. Finally, reviewing the class
slides and videos will be helpful if you run into difficulties.

Note: The "seed" to generate random numbers is set at the top of this Quarto
document. This will allow you to get the same answers every time you render the
document, and thus to answer questions that involve randomness in a consistent
way (i.e., you can read out results generated by random functions from the
rendered pdf and then comment on these results). In order to make it easier to
grade the homework, please do not change this seed, and do **not** call the
set.seed function anywhere else in the Quarto document. Also, to obtain
consistent values that you write in the answer sections (where applicable),
please render your file to a pdf first, and then write in the numbers from the
rendered pdf into the answer sections. If you run code chunks one at a time in
RStudio, the random numbers you see can differ from those in the rendered pdf,
so always report the numbers from the rendered pdf.






<!-- This R chunk sets some parameters that will be used in the rest of the document. Please ignore it and do not modify it. -->

```{r message=FALSE, warning=FALSE, tidy=TRUE, echo=FALSE}
    
  options(scipen=999)
  set.seed(100)
  library(SDS1000)
  
```




$\\$





## Part 1: Confidence intervals



Let's start by doing a few exercises to make sure you have a firm understanding
of confidence intervals.



$\\$



**Exercise 1.1 (6 points)** Which of the following statements are true about 
**90% confidence intervals**? Indicate all statements that are true by copying 
the full sentence(s) to the answer section: 

a. 90 percent of the sample data is contained in the intervals produced by this method.

b. 90 percent of the population data is contained in the intervals produced by this method.

c. 90 percent of the intervals produced by this method contain the parameter of interest.

d. 90 percent of the intervals produced by this method contain the statistic of interest.


**Answers:**







$\\$






**Exercise 1.2 (10 points)**: In the class we discussed that the formula for 
computing a 95% confidence interval for the mean $\mu$ can be written as: 
$\bar{x} ~ \pm 2 \cdot SE$. Please describe *as clearly as you can*, the logic 
of why an interval of this form is able to capture the parameter $\mu$ 95% of
the time. You can assume that the sampling distribution is normal and that there
is no bias.

If you are unsure of your answer, rewatching some of the class videos could be
useful. If you think it will help your explanation, you are welcome to create
drawings and embed them into your Quarto document.


**Answer:**






$\\$




**Exercise 1.3 (8 points)**: Please write the formula that can be used to compute 
an 88% confidence interval for a proportion. You can assume the standard error,
SE, is known and that the sampling distribution is normal. You are welcome to
use the R chunk below as well to help come up with the formula.


```{r CI_88}

library(SDS1000)







```

**Answer:**


 







$\\$







## Part 2: Exploring sampling distributions 



In the following exercises you will use a web application to explore sampling 
and bootstrap distributions. You will also explore standard errors and confidence 
intervals computed from these distributions. The application to use is located at:
[https://emeyers.github.io/SDS1000/apps/sampling_bootstrap_app/](https://emeyers.github.io/SDS1000/apps/sampling_bootstrap_app/).
The application runs R directly in your web browser, so it may take a few
seconds to load the first time you open it. The app generates its own random
samples, so the values you see will change each time it creates a new sample
and are not controlled by the seed in this Quarto document. This is expected,
and answers that use values from the app will differ from student to student.


To complete these exercises, open up the web application in another window. When
the application is open you will see the following three plots:

a.	The plot on the *upper left* shows the distribution of data in a population.
The shape of the population distribution can be changed using the population
distribution dropdown box, where the choices are: right skewed, bimodal and
normal.

b.	The plot on the *lower left* shows a histogram of the data from *one single sample* 
of size n. The size of the sample can be changed using the sample size input box.

c. The plot on the *upper right* shows an (approximate) sampling distribution
(histogram) of statistics based on 5,000 samples (i.e., each statistic in the
histogram was computed from one of the 5,000 samples). The type of statistic
that goes into this distribution can be changed using the 'statistic' drop down
box. Options for the statistics you can use are: the mean, the median and the
standard deviation.



When answering the questions below, please make sure you really understand the
relationship between these different distributions. One strategy to understand
these relationships is to predict what relationships should hold in the
questions that are being asked prior to looking at the app. Then use the app to
test if your understanding is correct. The type of questions that are asked here
are likely to be similar to the types of questions that might be on an exam, so
it will definitely be worth your time to make sure you understand these
relationships now (while you still have access to the app to check your
understanding!).






$\\$






**Exercise 2.1 (6 points)**:  Using the *mean* statistic and a sample size of 
*n = 100*, describe the shape of the *sampling distribution* for the: 

a. Right skewed population distribution 
b. Bimodal population distribution 
c. Normal population distribution 


Answer in the conclusion section below: does the shape of the *sampling distribution* 
change a lot depending on the shape of the *population distribution*? 


**Answers:**

a. shape of the sampling distribution is:  

b. shape of the sampling distribution is:  

c. shape of the sampling distribution is:  


Conclusion: 





$\\$







  
**Exercise 2.2 (6 points)**: Using a sample size of *n = 100* and a population 
distribution that is *right skewed*, describe the shape of the sampling
distribution for each of the following statistics:

a.  mean ($\bar{x}$)
b.	standard deviation (s) 
c.	median  


Answer in the conclusion section below: does the shape of the distribution
change a lot depending on the statistic chosen?


**Answers:**

a. shape of the sampling distribution is:  

b. shape of the sampling distribution is:  

c. shape of the sampling distribution is:  


Conclusion: 







$\\$









**Exercise 2.3 (6 points)**:  Keeping population distribution *right skewed* and 
the statistic being the sample mean $\bar{x}$. Compare the sample sizes of:

a.  n = 20
b.	n = 80 
c.	n = 320


In the conclusion section below describe the shape of the sampling distribution
for these different sample sizes and what the standard errors (SE) are. Also,
describe whether this is what you would expect and why. Finally, how does the
SE change each time the sample size is multiplied by 4?



**Answers:**

a. shape is:      , SE = 

b. shape is:      , SE = 

c. shape is:      , SE = 


Conclusion: 







$\\$







## Part 3: Explore bootstrap distributions 



Let's continue to use our web application, but now let's explore bootstrap
distributions.

For the questions in this part, check the box that says 'Display Bootstrap
Distribution'. You should notice a new plot in the lower right that shows the
bootstrap distribution that is created from the one sample in the lower left
plot.


$\\$


**Exercise 3.1 (8 points)**:  The centers of these distributions are given by red 
lines on the plots, and the center values noted at the top of the 4 plots. To
get a sense of the relationship between the center values between the plots, set
the statistic to the mean and the sample size to n = 100 (using any population
distribution), and click the 'Draw a new sample' button several times to
generate new samples and new bootstrap distributions. Then answer the following
questions:


a.	How does the population mean ($\mu$) relate to the center of the sampling
distribution ($E[\bar{x}]$)? Is there bias here?

b.	How does the center of the sample ($\bar{x}$) relate to the population
mean ($\mu$)? Is this expected? Briefly explain why.

c.	How does the center of the bootstrap distribution ($E[\bar{x}*]$) relate to
the center of the sample ($\bar{x}$)?

d.	How does the center of the bootstrap distribution relate to the center of the
population ($\mu$)?




**Answers:**

a. 


b. 


c. 


d.






$\\$









**Exercise 3.2 (10 points)**:  How does the standard error and confidence
interval of the sampling distribution compare to the standard error and
confidence interval created from the bootstrap distribution? To explore this,
set the distribution to normal and the statistic to the mean, and run the
following steps twice. The first row of the table below shows my results from
one run; fill in the second and third rows with the results from your two runs.


a.	Make sure the sample size is set to n = 100, and then click the 'Draw a new
sample' button. This will create a new sample and a new bootstrap distribution.

b.	Fill in a row of the table below with the value of $\bar{x}$ from the sample,
the SE from the sampling distribution, and the $SE*$ from the bootstrap
distribution. Then, using $\bar{x}$, compute 95% confidence intervals from the
sampling distribution (using SE) and from the bootstrap distribution (using
$SE*$) with the formulas $CI = \bar{x} \pm 2 \cdot SE$ and
$CI_b = \bar{x} \pm 2 \cdot SE*$, and add these intervals to the table.


Do all these intervals capture the population parameter? 
Also, across your runs, compare the SE with $SE*$ and the two confidence
intervals. How similar are they, and what differences do you notice?




**Answers:**


| $\bar{x}$  |   SE   |        CI      |   SE*   |  CI-bootstrap | 
|------------|--------|----------------|---------|---------------|
|    5.07    |  0.10  |  [4.87  5.27]  |   0.13  | [4.81  5.33]  |
|            |        |                |         |               |
|            |        |                |         |               |


Write your answers to these questions here:









$\\$






**Exercise 3.3 (10 points)**:  Repeat exercise 3.2 but using the right skewed
population. As in exercise 3.2, the first row of the table below shows my
results from one run; fill in the second and third rows with the results from
your two runs.

a.	Make sure the sample size is set to n = 100, and then click the 'Draw a new
sample' button. This will create a new sample and a new bootstrap distribution.

b.	Fill in a row of the table below with the value of $\bar{x}$ from the sample,
the SE from the sampling distribution, and the $SE*$ from the bootstrap
distribution. Then, using $\bar{x}$, compute 95% confidence intervals from the
sampling distribution (using SE) and from the bootstrap distribution (using
$SE*$) with the formulas $CI = \bar{x} \pm 2 \cdot SE$ and
$CI_b = \bar{x} \pm 2 \cdot SE*$, and add these intervals to the table.


Do all these intervals capture the population parameter? 
Also, across your runs, compare the SE with $SE*$ and the two confidence
intervals. How similar are they, and what differences do you notice?




**Answers:**



| $\bar{x}$  |   SE   |        CI      |   SE*   |  CI-bootstrap | 
|------------|--------|----------------|---------|---------------|
|    1.91    |  0.14  |  [1.63  2.19]  |   0.13  | [1.65  2.17]  |
|            |        |                |         |               |
|            |        |                |         |               |




Write your answers to these questions here:









$\\$








## Part 4: Calculating confidence intervals using the bootstrap in R



In this last set of exercises you will calculate a confidence interval for the
mean price of Hondas that were sold on [Edmunds.com](http://edmunds.com) between
December 2013 and February 2015.





$\\$






**Exercise 4.1 (8 points)**:  Let's start by looking at the price of Hondas that 
were sold on Edmunds.com. The code below loads the data on the prices of cars
and extracts a sample of prices that different Hondas were sold for (you can
treat these cars as a random sample of all the Hondas that were sold on
Edmunds.com during this time). Please do the following:

1. Calculate the size of the sample and store the answer in an object called `n_sample_size`.

2. Calculate the mean price that Hondas were sold for and store the result in
an object called `the_stat`. 

3. Plot a histogram of the car prices and don't forget to label your axes!

In the answer section, report the value for the sample size and the mean price
Hondas were sold for and use the appropriate symbol.



```{r honda_hist, message=FALSE, warning=FALSE}

# load the car price data
load("cars_small.Rda")


# extract the sample of Honda prices
price_sample <- subset(car_data$price, car_data$brand == "Honda")



# get the sample size and store it in an object called n_sample_size




# get the mean price and store it in an object called the_stat 



# plot a histogram of the Honda prices




```



**Answer:** 







$\\$




    
**Exercise 4.2 (14 points)**: Now use the `do_it()` function to create a bootstrap 
distribution by doing the following steps:

1. Sample with replacement `n_sample_size` car prices from the `price_sample` vector. 

2. Calculate the mean of the bootstrap sample. 

3. Repeat this process 10,000 times and store the results in a vector called
`bootstrap_dist`.

4. Plot a histogram of the bootstrap distribution

5. Calculate the bootstrap standard error ($SE*$), store the value in an
object called `SE_boot`, and then print out the bootstrap standard error.

6. Calculate a 95% confidence interval using the formula:  $CI_b = \bar{x} \pm  2 \cdot SE*$ and
report what the interval is in the answer section below.


```{r honda_boot, message=FALSE, warning=FALSE}


# Create the bootstrap distribution following steps 1-3









# Plot a histogram of the bootstrap distribution






# Calculate SE* and print out its value





# Calculate a 95% confidence interval using SE*






```



**Answer:** 






$\\$









**Exercise 4.3 (5 points)**  If you were to create a confidence interval for the
mean price of Hyundais using the bootstrap method on the Edmunds data, you would
get a 95% confidence interval of [\$19,505, \$19,778] (you can verify this
yourself by replacing the line above that gets the car prices with
`price_sample <- subset(car_data$price, car_data$brand == "Hyundai")`). Based
on the confidence
interval you calculated for Hondas above, and the confidence interval for
Hyundais I just gave you, is it likely that the population mean price for Hondas
($\mu_{honda}$) is the same as the population mean price for Hyundais
($\mu_{hyundai}$)?  Why or why not?



**Answer:** 









$\\$












## Reflection (3 points)


Please reflect on how the homework went by going to Canvas, going to the Quizzes
link, and clicking on Reflection on homework 5.







