---
title: "Class 22 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---






# Part 1: Parametric confidence interval for a single mean



A study by Nierenberg et al (1989) investigated the relationship between dietary
factors, and plasma concentrations of carotenoids.

Let’s use the data they collected to create a 98% confidence interval for the
number of grams fiber US adults get in a day.



$$CI ~ = ~ \bar{x} ~ \pm ~ t^* \cdot \frac{s}{\sqrt{n}}$$



```{r confidence_interval_mean}

# Given data
nutrition_df <- read.csv("NutritionStudy.csv")
fiber <- nutrition_df$Fiber


# Get basic descriptive statistics

           


# Degrees of freedom



# Calculate the standard error (SE)





# Find the critical t-value for the desired confidence level





# Calculate the margin of error (ME)





# Construct the confidence interval






```




$\\$





# Part 2: Parametric hypothesis test for a single mean


In the mid-1990s a Nabisco marking campaign claimed that there were at least
1000 chips in every bag of Chips Ahoy! cookies.

A group of Air Force cadets tested this claim by dissolving the cookies from 42
bags in water and counting the number of chips.

They found the average number of chips per bag was 1261.6, with a standard
deviation of 117.6 chips

Test whether the average (mean) number of chips per bag is greater than 1000.  

Do the results confirm Nabisco’s claim?  



$\\$

Step 1: State the null and alternative hypotheses


In words: 

- Null hypothesis: 
- Alternative hypothesis: 


In symbols:  

- $H_0:$  
- $H_A:$  



```{r hypothesis_test_mean}



# Step 2: Calculate the t-statistic

n <- 42
x_bar <- 1261.6
s <- 117.6
mu_0 <- 1000






# Step 3: Visualize the null distribution






# Step 4: Calculate the p-value





```



Step 5: Make a decision




Does this confirm Nabisco's claim? 








$\\$




# Part 3: Confidence interval for the difference of two means

A study by Nierenberg et al (1989) investigated the relationship between dietary
factors, and plasma concentrations of carotenoids.

Let’s continue to use the data they collected, but this time let's create a 95%
confidence interval for the differences in the number of grams of fiber eaten in
a day between males and females.


$$CI ~ = ~ (\bar{x}_1 - \bar{x}_2) ~ \pm ~ t^* \cdot \sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}$$

```{r confidence_interval_difference_means}

# Given data
nutrition_df <- read.csv("NutritionStudy.csv")

fiber_males  <- subset(nutrition_df, Sex == "Male")$Fiber
fiber_females <- subset(nutrition_df, Sex == "Female")$Fiber


# Get basic descriptive statistics for males

          

# Get basic descriptive statistics for females

          


# Degrees of freedom (using the smaller of n1-1 and n2-1)



# Calculate the standard error (SE)



# Find the critical t-value for the desired confidence level




# Calculate the margin of error (ME)




# Construct the confidence interval



```


Question: Does it seem that there is a difference in the amount of fiber males
and females eat in a day?






$\\$




# Part 4: Hypothesis test for the difference of two means


A study randomly sampled 2295 American men 
- 2027 men were right-handed, 268 men were left-handed
- Right-handers earned \$13.10/hr, left-handers earned \$13.40/hr
- The standard deviation for both groups was \$7.90

Test the hypothesis that there is a difference in earnings between right and
left-handed men.

```{r hypothesis_test_difference_means}

# Given data
n1 <- 2027
x_bar1 <- 13.10
s1 <- 7.90

n2 <- 268
x_bar2 <- 13.40
s2 <- 7.90



# Step 2: Calculate the t-statistic





# Step 3: Visualize the null distribution







# Step 4: Calculate the p-value





```


Step 5: Make a decision






$\\$




# Part 5: Paired samples t-test


A professor was interested in assessing if students scored higher on average on
the second exam based on a sample of data from 10 students. 

Let's run a paired samples t-test to see if there is a significant difference in
exam scores between the first and second exams.

Step 1:







```{r paired_samples_t_test}

# Given data
exam1 <- c(72, 95, 56, 87, 80, 98, 74, 85, 77, 62)
exam2 <- c(78, 96, 72, 89, 80, 95, 86, 87, 82, 72)


# Calculate the differences



# Step 2: Calculate the t-statistic


  


# Step 4: Calculate the p-value





```



Step 5: Make a decision





Try this at home: Run an independent samples t-test that assumes the two samples
of exam scores are come from different randomly selected students. Do you come
to the same conclusion if you ran this hypothesis test? 



$\\$



