---
title: "Class 19 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---







## Part 1: Generating random data from a normal distribution


IQ scores are normally distributed with a mean of 100 and a standard deviation of 15.

Let's generate/simulate some random IQ scores and visualize the distribution.



```{r normal_data}

set.seed(123)  # for reproducibility
n <- 1000

# Generate random IQ scores
iq_scores <- rnorm(n, mean = 100, sd = 15)


# Visualize the distribution
hist(iq_scores, 
     breaks = 30, 
     main = "Histogram of Simulated IQ Scores", 
     xlab = "IQ Score")


```





$\\$





## Part 2: Visualzing the distribution of IQ scores with a density plot


We can visualize the distribution of IQ scores using a density plot.


```{r density_plot}

# Get a range of x values
x_values <- seq(50, 150, length.out = 1000)

                
# Calculate the density values for the normal distribution
density_values <- dnorm(x_values, mean = 100, sd = 15)


# Plot the density
plot(x_values, density_values,
     type = "l",
     main = "Density Plot of IQ Scores",
     xlab = "IQ Score",
     ylab = "Density")

```


We can also overlay the density plot on the histogram of the simulated IQ scores by normalizing the histogram so that the area sums to 1.


```{r overlay_density}

# Plot the histogram with density
hist(iq_scores,
     breaks = 50,
     probability = TRUE,  # normalize the histogram
     main = "Histogram of Simulated IQ Scores with Density Overlay",
     xlab = "IQ Score")


# Overlay the density plot - lines() adds to the previous plot
lines(x_values, density_values, col = "red")

```






$\\$





## Part 3: Calculating probabilities using the normal distribution

We can calculate probabilities using the normal distribution. For example, what is the probability that a randomly selected individual has an IQ score less than 120?


```{r calculate_probability}

# Calculate the probability
probability <- pnorm(120, mean = 100, sd = 15)

probability


# we can viaulize this using the mosaic package xpnorm function
# mosaic::xpnorm(120, mean = 100, sd = 15)


```


We can also get probability ranges by subtracting cumulative probabilities.

What is the probability that a randomly selected individual has an IQ score between 85 and 115?


```{r calculate_probability_range}

# Calculate the probability

probability_range <- pnorm(115, mean = 100, sd = 15) - pnorm(85, mean = 100, sd = 15)

probability_range




# we can viaulize this on a density plot 
plot(x_values, density_values,
     type = "l",
     xlab = "IQ Score",
     ylab = "Density")

abline(v = c(85, 115), col = "red")
abline(h = 0, col = "black")


```





$\\$






## Part 4: The standard normal distribution

The standard normal distribution is a special case of the normal distribution 
where the mean is 0 and the standard deviation is 1.




$\\$




### Part 4.1: Converting data from an arbitrary normal distribution to a standard normal distribution

We can convert any normal distribution to the standard normal distribution using
the z-score formula: $$ z = \frac{x - \mu}{\sigma} $$

Where: 
- `z` is the z-score
- `x` is the value from the original normal distribution
- $\mu$ is the mean of the original normal distribution
- $sigma$ is the standard deviation of the original normal distribution

Let's say we have an IQ score of 112. What is the corresponding z-score?

```{r z_score}


# Given IQ score
x <- 112


# Mean and standard deviation of the IQ distribution
mu <- 100
sigma <- 15


# Calculate the z-score
z <- (x - mu) / sigma
z


# Now, let's find the probability of having an IQ score greater than 130 using the z-score
probability_z <- pnorm(z, mean = 0, sd = 1, lower.tail = FALSE)


```



$\\$



### Part 4.2: Converting many data points to a standard normal distribution 

We can also convert many data points from an arbitrary normal distribution to a
standard normal distribution. 


```{r zscore_many_points}

set.seed(100)


# generate 1000 random IQ scores
many_iq_scores <- rnorm(1000, 100, 15)

# visualize these IQ scores
hist(many_iq_scores, breaks = 50,
     xlab = "IQ scores",
     main = "1000 random IQ scores")



# convert these scores to z-scores
iq_z_scores <- (many_iq_scores - 100)/15

# visualize these z-scores
hist(iq_z_scores, breaks = 50,
     xlab = "Z-scores",
     main = "IQ scores converted to z-scores")



```





$\\$





### Part 4.3: Converting from a standard normal distribution to an arbitrary normal distribution


We can also convert z-scores back to the original normal distribution using the formula:
$$ X = z \cdot \sigma + \mu $$


Let's say we have a z-score of 2.2. What is the corresponding IQ score?

```{r z_to_x}

# Given z-score
z_value <- 2.2

# Convert z-score back to IQ score

x_value <- z_value * sigma + mu
x_value


```




$\\$






