---
title: "Parametric Hypothesis Tests"
description: |
  A reference guide covering every parametric hypothesis test from S&DS 1000:
  one and two proportions, multiple proportions (chi-squared), one mean, two
  means (independent and paired), multiple means (ANOVA), correlation, and
  simple linear regression.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parametric Hypothesis Tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6, fig.height = 3.5)

library(SDS1000)

```



## Overview

This guide covers every parametric hypothesis test from S&DS 1000. Each test
follows the same five steps used in class:

1. **State** the null and alternative hypotheses
2. **Calculate** the observed test statistic (and check conditions)
3. **Visualize** the null distribution and mark the observed statistic
4. **Calculate** the p-value
5. **Make a decision**

<div class="guide-nav">
**In this article:**

- [One proportion](#one-proportion)
- [Two proportions](#two-proportions)
- [More than two proportions (chi-squared)](#more-than-two-proportions)
- [One mean](#one-mean)
- [Two means — independent](#two-means-independent-samples)
- [Two means — paired](#two-means-paired-samples)
- [More than two means (ANOVA)](#more-than-two-means-anova)
- [Correlation](#correlation)
- [Simple linear regression](#simple-linear-regression)
- [Quick reference table](#quick-reference)

<br>

**Also in Class Summaries:**

- [Randomization Hypothesis Tests](class-summary-randomization-tests.html)
- [Randomization Confidence Intervals](class-summary-randomization-cis.html)
- [Parametric Confidence Intervals](class-summary-parametric-cis.html)
</div>

<div class="warn-box">
**A note on assumptions.** Parametric tests rely on mathematical
approximations that are only valid when certain conditions are met (e.g.,
large enough sample sizes, approximate normality). Always check conditions
before interpreting a p-value. The randomization methods from Parts 1 and 2
are more robust when conditions are uncertain.
</div>

---

## One Proportion

**When to use:** One categorical variable with two outcomes; test whether the
population proportion π equals a specific value π₀.

**Conditions:** $n\pi_0 \geq 10$ and $n(1 - \pi_0) \geq 10$

**Test statistic:**
$$z = \frac{\hat{p} - \pi_0}{SE_0}, \qquad SE_0 = \sqrt{\frac{\pi_0(1 - \pi_0)}{n}}$$

**Example (class 20):** From 1982–1994 there were 128 penalty shots in the
World Cup. Goalkeepers correctly guessed the kick direction 41% of the time.
Test whether goalies guessed correctly **less than** 50% of the time.

### Step 1: State the hypotheses

$$H_0: \pi = 0.50 \qquad H_A: \pi < 0.50$$

### Steps 2–4: Calculate, visualize, p-value

```{r one_prop}
# Step 2: Calculate the z-statistic
pi_0  <- 0.50
n     <- 128
p_hat <- 0.41

# Check conditions
n * pi_0        # must be >= 10
n * (1 - pi_0)  # must be >= 10

SE     <- sqrt(pi_0 * (1 - pi_0) / n)
z_stat <- (p_hat - pi_0) / SE
z_stat


# Step 3: Visualize the null distribution
plot_norm(col = "blue",
          main = "Null Distribution — Z-statistic",
          xlab = "Z-score", ylab = "Density")
abline(v = z_stat, col = "red", lwd = 2)

# Step 4: p-value (lower tail, H_A: pi < pi_0)
p_value <- pnorm(z_stat, mean = 0, sd = 1, lower.tail = TRUE)
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 4)` < 0.05, we **reject** $H_0$ and conclude
that goalkeepers guess the correct direction less than 50% of the time.

---

## Two Proportions

**When to use:** A binary outcome measured in two independent groups; test
whether the two population proportions are equal.

**Conditions:** At least 10 successes and 10 failures in each group.

**Test statistic** (using the unpooled standard error):
$$z = \frac{\hat{p}_1 - \hat{p}_2}{SE}, \qquad
SE = \sqrt{\frac{\hat{p}_1(1-\hat{p}_1)}{n_1} + \frac{\hat{p}_2(1-\hat{p}_2)}{n_2}}$$

**Example:** In a clinical trial, 245 patients received a new vaccine and 255
received a placebo. Of the vaccinated group 32 got the flu; of the placebo
group 56 got the flu. Test whether the vaccine **reduces** the proportion who
get the flu.

### Step 1: State the hypotheses

Let group 1 = vaccine, group 2 = placebo.
$$H_0: \pi_1 = \pi_2 \qquad H_A: \pi_1 < \pi_2$$

### Steps 2–4: Calculate, visualize, p-value

```{r two_prop}
# Step 2: Calculate the z-statistic
x1 <- 32   # vaccinated
n1 <- 245;
x2 <- 56   # placebo
n2 <- 255;  

p_hat1 <- x1 / n1
p_hat2 <- x2 / n2

# Check conditions (successes and failures >= 10 in each group)
c(x1, n1 - x1, x2, n2 - x2)

# Unpooled standard error
SE <- sqrt((p_hat1 * (1 - p_hat1) / n1) + (p_hat2 * (1 - p_hat2) / n2))

z_stat <- (p_hat1 - p_hat2) / SE
z_stat

# Step 3: Visualize the null distribution
plot_norm(col = "blue",
          main = "Null Distribution — Z-statistic",
          xlab = "Z-score", ylab = "Density")
abline(v = z_stat, col = "red", lwd = 2)

# Step 4: p-value (lower tail, H_A: pi_1 < pi_2)
p_value <- pnorm(z_stat, mean = 0, sd = 1, lower.tail = TRUE)
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 4)` < 0.05, we **reject** $H_0$ and conclude
that the vaccine significantly reduces the proportion of flu cases.

---

## More Than Two Proportions

**When to use:** One categorical variable with three or more outcomes; test
whether the observed counts match a set of expected proportions.

**Conditions:** Expected count in each cell $\geq 5$

**Test statistic:**
$$\chi^2 = \sum_{i=1}^{k} \frac{(O_i - E_i)^2}{E_i}, \qquad df = k - 1$$

**Example (class 23/24):** A Yale professor gathered birth months for 198
randomly selected Yale students. Are students equally likely to be born in
any month?

### Step 1: State the hypotheses

$$H_0: \pi_{Jan} = \pi_{Feb} = \cdots = \pi_{Dec} = \tfrac{1}{12}$$
$$H_A: \text{at least one monthly proportion differs from } \tfrac{1}{12}$$

### Steps 2–4: Calculate, visualize, p-value

```{r chisq_gof}
# Step 2: Calculate the chi-squared statistic
observed_counts <- c(14, 11, 21, 17, 15, 13, 19, 16, 18, 22, 14, 18)
names(observed_counts) <- month.abb

expected_counts <- rep(sum(observed_counts) / 12, 12)

# Check conditions: expected counts >= 5
expected_counts[1]   # all equal ~16.5 ✓

chi_sq_stat <- sum((observed_counts - expected_counts)^2 / expected_counts)
chi_sq_stat

# Step 3: Visualize the null distribution
df   <- length(observed_counts) - 1   # k - 1 = 11

plot_chisq(df = df, col = "blue",
           main = "Null Distribution — Chi-squared Statistic",
           xlab = "Chi-squared value", ylab = "Density")
abline(v = chi_sq_stat, col = "red", lwd = 2)

# Step 4: p-value (always upper tail for chi-squared)
p_value <- pchisq(chi_sq_stat, df = df, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 3)` > 0.05, we **fail to reject** $H_0$. There
is not sufficient evidence that Yale students' birth months deviate from a 
uniform distribution.

---

## One Mean

**When to use:** One quantitative variable; test whether the population mean μ
equals a specific value μ₀.

**Conditions:** $n \geq 30$, or the population is approximately normal.

**Test statistic:**
$$t = \frac{\bar{x} - \mu_0}{SE}, \qquad SE = \frac{s}{\sqrt{n}}, \qquad df = n - 1$$

**Example (class 21):** Air Force cadets tested 42 bags of Chips Ahoy! cookies
and found a mean of 1261.6 chips, sd = 117.6. Test whether the mean exceeds
1000.

### Step 1: State the hypotheses

$$H_0: \mu = 1000 \qquad H_A: \mu > 1000$$

### Steps 2–4: Calculate, visualize, p-value

```{r one_mean}
# Step 2: Calculate the t-statistic
n     <- 42
x_bar <- 1261.6
s     <- 117.6
mu_0  <- 1000

SE     <- s / sqrt(n)
t_stat <- (x_bar - mu_0) / SE
t_stat

# Step 3: Visualize the null distribution
df   <- n - 1
plot_t(df, to = 15, col = "blue",
       main = "Null Distribution — T-statistic",
       xlab = "T-score", ylab = "Density")
abline(v = t_stat, col = "red", lwd = 2)

# Step 4: p-value (upper tail, H_A: mu > mu_0)
p_value <- pt(t_stat, df = df, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision

Since p ≈ 0, we **reject** $H_0$ and conclude the average chip count exceeds
1000. Note this confirms the average exceeds 1000, not that every individual
bag does.

---

## Two Means: Independent Samples

**When to use:** A quantitative variable measured in two independent groups;
test whether the group means are equal.

**Conditions:** Each group has $n \geq 30$, or both populations are
approximately normal.

**Test statistic (Welch's t-test — does not assume equal variances):**
$$t = \frac{(\bar{x}_1 - \bar{x}_2)}{SE_{diff}}, \qquad
SE_{diff} = \sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}, \qquad
df = \min(n_1 - 1,\; n_2 - 1)$$

**Example (class 22):** A study of 2295 American men found that right-handers
(n = 2027) earned \$13.10/hr and left-handers (n = 268) earned \$13.40/hr,
both with sd = \$7.90. Test whether there is a difference in earnings.

### Step 1: State the hypotheses

$$H_0: \mu_1 = \mu_2 \qquad H_A: \mu_1 \ne \mu_2$$

### Steps 2–4: Calculate, visualize, p-value

```{r two_means_indep}
# Step 2: Calculate the t-statistic
n1 <- 2027;  x_bar1 <- 13.10;  s1 <- 7.90   # right-handed
n2 <- 268;   x_bar2 <- 13.40;  s2 <- 7.90   # left-handed

SE_diff <- sqrt((s1^2 / n1) + (s2^2 / n2))
t_stat  <- (x_bar1 - x_bar2) / SE_diff
t_stat

# Step 3: Visualize the null distribution
df   <- min(n1 - 1, n2 - 1)

plot_t(df, col = "blue",
       main = "Null Distribution — T-statistic",
       xlab = "T-score", ylab = "Density")
abline(v =  t_stat, col = "red", lwd = 2)
abline(v = -t_stat, col = "red", lwd = 2)   # two-sided: mark both tails

# Step 4: p-value (two-sided)
p_value <- 2 * pt(-abs(t_stat), df = df)
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 3)` > 0.05, we **fail to reject** $H_0$. There is
not sufficient evidence of a difference in hourly earnings between right- and
left-handed men.

---

## Two Means: Paired Samples

**When to use:** Observations come in natural pairs (e.g., before/after on the
same subject). Compute the within-pair differences and run a one-sample t-test
on them.

**Test statistic:**
$$t = \frac{\bar{d}}{SE_d}, \qquad SE_d = \frac{s_d}{\sqrt{n}}, \qquad df = n - 1$$

**Example (class 22):** Exam scores for 10 students on Exam 1 and Exam 2.
Test whether students scored higher on Exam 2 on average.

### Step 1: State the hypotheses

$$H_0: \mu_d = 0 \qquad H_A: \mu_d > 0$$

### Steps 2–4: Calculate, visualize, p-value

```{r two_means_paired}
# Step 2: Calculate the t-statistic on the differences
exam1 <- c(72, 95, 56, 87, 80, 98, 74, 85, 77, 62)
exam2 <- c(78, 96, 72, 89, 80, 95, 86, 87, 82, 72)

d      <- exam2 - exam1
n      <- length(d)
SE_d   <- sd(d) / sqrt(n)
t_stat <- mean(d) / SE_d
t_stat

# Step 3: Visualize the null distribution
df   <- n - 1

plot_t(df, col = "blue",
     main = "Null Distribution — T-statistic (Paired)",
     xlab = "T-score", ylab = "Density")
abline(v = t_stat, col = "red", lwd = 2)

# Step 4: p-value (upper tail, H_A: mu_d > 0)
p_value <- pt(t_stat, df = df, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 4)` < 0.05, we **reject** $H_0$ and conclude
that students scored significantly higher on Exam 2.

---

## More Than Two Means: ANOVA

**When to use:** A quantitative variable measured across three or more
independent groups; test whether all group means are equal.

**Conditions:** Approximately normal residuals; roughly equal variance across
groups; independent observations.

**Test statistic:**
$$F = \frac{MSG}{MSE}, \qquad df_1 = k - 1, \quad df_2 = N - k$$

A large F means between-group variation is large relative to within-group
variation. The SDS1000 function `get_F_stat()` returns the value of the F-statistic.

**Example (class 24):** Hope College students were timed completing a
Sudoku-like puzzle. Majors were grouped into four categories. Is mean
completion time the same across majors?

### Step 1: State the hypotheses

$$H_0: \mu_{as} = \mu_{ns} = \mu_{ss} = \mu_{ah} \qquad
H_A: \text{at least one mean differs}$$

### Steps 2–4: Calculate, visualize, p-value

```{r anova}
# Step 2: Load data and calculate the F-statistic
set.seed(4455)  # generate fake data
completion_times <- c(rnorm(10, mean = 22, sd = 5),
                      rnorm(10, mean = 20, sd = 5),
                      rnorm(10, mean = 26, sd = 5),
                      rnorm(10, mean = 21, sd = 5))
majors <- factor(rep(c("Applied Sci", "Natural Sci", "Social Sci", "Arts/Hum"),
                     each = 10))

# Visualize the data
boxplot(completion_times ~ majors,
        main  = "Completion Time by Major",
        ylab  = "Time (seconds)",
        col   = c("lightblue", "lightgreen", "lightpink", "lightyellow"))

# Fit the model and extract the F-statistic
f_stat <- get_F_stat(completion_times, majors)
df1 <- length(unique(majors)) - 1   # k - 1
df2 <- length(completion_times) - length(unique(majors))  # N - k
f_stat

# Step 3: Visualize the null distribution
plot_f(df1, df2, col = "blue",
     main = "Null Distribution — F-statistic",
     xlab = "F value", ylab = "Density")
abline(v = f_stat, col = "red", lwd = 2)

# Step 4: p-value (always upper tail for F)
p_value <- pf(f_stat, df1, df2, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision and follow-up

```{r anova_decision, echo=FALSE}
dec <- ifelse(p_value < 0.05,
              "reject H\u2080 — at least one group mean differs",
              "fail to reject H\u2080 — no evidence of a difference between group means")
```

p = `r round(p_value, 3)`: we `r dec`.


---

## Correlation

**When to use:** Two quantitative variables; test whether the linear
association (correlation) between them is significantly different from zero.

**Conditions:** Both variables are approximately normal, or $n$ is large
enough that the CLT applies; the relationship is linear.

**Test statistic:**
$$t = \frac{r\sqrt{n-2}}{\sqrt{1-r^2}}, \qquad df = n - 2$$

**Example (class 25):** Data on cigarette consumption per capita and lung
cancer rates across 44 U.S. states. Is there evidence of a positive linear
association?

### Step 1: State the hypotheses

$$H_0: \rho = 0 \qquad H_A: \rho > 0$$

### Steps 2–4: Calculate, visualize, p-value

```{r correlation}
# Data simulated to match the class 25 states_smoking dataset
set.seed(2947)
cigs_per_capita <- runif(44, min = 10, max = 45)  # create fake data
cancer_rate     <- 2 + 0.005 * cigs_per_capita * 1000 + rnorm(44, sd = 5)

# Visualize the data
plot(cigs_per_capita, cancer_rate,
     xlab = "Cigarettes per Capita (hundreds)",
     ylab = "Lung Cancer Rate per 100,000",
     main = "Lung Cancer Rate vs. Cigarette Consumption")

# Step 2: Calculate the t-statistic
n <- length(cigs_per_capita)
r <- cor(cigs_per_capita, cancer_rate)
r

t_stat <- r * sqrt(n - 2) / sqrt(1 - r^2)
t_stat

# Step 3: Visualize the null distribution
df   <- n - 2

plot_t(df, col = "blue",
     main = "Null Distribution — T-statistic",
     xlab = "T-score", ylab = "Density")
abline(v = t_stat, col = "red", lwd = 2)

# Step 4: p-value (upper tail, H_A: rho > 0)
p_value <- pt(t_stat, df = df, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 4)`, we **reject** $H_0$ and conclude that there
is a significant positive linear association between cigarette consumption and
lung cancer rate.

<div class="tip-box">
**Shortcut:** `cor.test(x, y, alternative = "greater")` performs all of steps
2–4 in one call and returns the same t-statistic, df, and p-value.
</div>

---

## Simple Linear Regression

**When to use:** Two quantitative variables where you want to model y as a
linear function of x **and** test whether the slope is significantly different
from zero. Testing H₀: β₁ = 0 is equivalent to testing H₀: ρ = 0 — they
yield identical t-statistics and p-values.

**Conditions:** Linear relationship; approximately normal residuals; constant
variance (homoscedasticity); independent observations.

**Test statistic for the slope:**
$$t = \frac{\hat{\beta}_1}{SE_{\hat{\beta}_1}}, \qquad df = n - 2$$

**Example (class 25):** Using the same cigarette/lung cancer data, fit a
regression model and test whether the slope is significantly positive.

### Step 1: State the hypotheses

$$H_0: \beta_1 = 0 \qquad H_A: \beta_1 > 0$$

### Steps 2–4: Fit the model, visualize, p-value

```{r regression}
# Fit the linear model (using same data from correlation section above)
lung_lm <- lm(cancer_rate ~ cigs_per_capita)

# Step 2: summary() reports the t-statistic and two-sided p-value for each coefficient
summary(lung_lm)
```

Reading the `Coefficients` table:

| Column | Meaning |
|---|---|
| `Estimate` | Fitted intercept (β̂₀) and slope (β̂₁) |
| `Std. Error` | Standard error of each estimate |
| `t value` | t-statistic testing H₀: coefficient = 0 |
| `Pr(>|t|)` | Two-sided p-value |

```{r regression_viz}
# Step 3: Visualize the null distribution and the observed t-statistic
# (The t-stat and df for the slope are extracted from the model summary)
fit_sum <- summary(lung_lm)
t_stat  <- fit_sum$coefficients["cigs_per_capita", "t value"]
df      <- lung_lm$df.residual   # n - 2

plot_t(df, col = "blue",
     main = "Null Distribution — T-statistic for Slope",
     xlab = "T-score", ylab = "Density")
abline(v = t_stat, col = "red", lwd = 2)

# Step 4: p-value (upper tail; summary() reports two-sided, so divide by 2)
p_two_sided <- fit_sum$coefficients["cigs_per_capita", "Pr(>|t|)"]
p_value     <- p_two_sided / 2   # one-sided upper-tail p-value
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 4)` < 0.05, we **reject** $H_0$ and conclude
that there is a significant positive relationship between cigarette consumption
and lung cancer rate.

### Confidence interval for the slope

`confint()` gives theory-based confidence intervals for each coefficient —
the parametric alternative to the bootstrap CI from class 25:

```{r regression_ci}
confint(lung_lm, level = 0.90)   # 90% CI, as used in class 25
```

---

## Quick Reference

| Test | H₀ | Statistic | df | p-value function |
|---|---|---|---|---|
| One proportion | $\pi = \pi_0$ | $z = \frac{\hat{p} - \pi_0}{SE_0}$ | — | `pnorm(z_stat, lower.tail = ...)` |
| Two proportions | $\pi_1 = \pi_2$ | $z = \frac{\hat{p}_1 - \hat{p}_2}{SE}$ | — | `pnorm(z_stat, lower.tail = ...)` |
| Multiple proportions | $\pi_1 = \cdots = \pi_k = \pi_0$ | $\chi^2 = \sum\frac{(O-E)^2}{E}$ | $k-1$ | `pchisq(chi_sq, df, lower.tail=FALSE)` |
| One mean | $\mu = \mu_0$ | $t = \frac{\bar{x}-\mu_0}{s/\sqrt{n}}$ | $n-1$ | `pt(t_stat, df, lower.tail = ...)` |
| Two means (indep.) | $\mu_1 = \mu_2$ | $t = \frac{\bar{x}_1-\bar{x}_2}{SE_{diff}}$ | $\min(n_1,n_2)-1$ | `pt(t_stat, df, lower.tail = ...)` |
| Two means (paired) | $\mu_d = 0$ | $t = \frac{\bar{d}}{s_d/\sqrt{n}}$ | $n-1$ | `pt(t_stat, df, lower.tail = ...)` |
| Multiple means | $\mu_1 = \cdots = \mu_k$ | $F = MSG/MSE$ | $k-1$, $N-k$ | `pf(f_stat, df1, df2, lower.tail=FALSE)` |
| Correlation | $\rho = 0$ | $t = \frac{r\sqrt{n-2}}{\sqrt{1-r^2}}$ | $n-2$ | `pt(t_stat, df, lower.tail = ...)` |
| Regression slope | $\beta_1 = 0$ | $t = \hat{\beta}_1 / SE_{\hat{\beta}_1}$ | $n-2$ | `summary(lm(...))` or `pt(t_stat, df, ...)` |
