---
title: "Parametric Confidence Intervals"
description: |
  A reference guide covering every parametric confidence interval from S&DS 1000:
  one proportion, one mean, difference of two means, and regression slope.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Parametric Confidence Intervals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6, fig.height = 3.5)
library(SDS1000)
```

<style>
h2 {
  margin-top:     3.5rem;
  padding-top:    0.6rem;
  padding-bottom: 0.4rem;
  border-bottom:  1px solid #dee2e6;
  color: #1a3a5c;
  font-size: 1.45rem;
}
h3 { margin-top: 2rem; color: #2c5282; font-size: 1.15rem; }
hr { margin: 3rem 0 2rem 0; border: none; border-top: 2px solid #e2e8f0; }
table {
  margin: 1.25rem 0 1.75rem 0;
  font-size: 0.93rem;
  width: 100%;
}
table thead tr { background-color: #dbeafe; color: #1e3a5f; }
table tbody tr:nth-child(even) { background-color: #f8fafc; }
.guide-nav {
  background-color: #f0f7ff; border: 1px solid #bfdbfe;
  border-radius: 6px; padding: 1.1rem 1.4rem;
  margin-bottom: 2rem; font-size: 0.95rem;
}
.guide-nav ul { margin-bottom: 0; padding-left: 1.2rem; }
.guide-nav li { margin-bottom: 0.3rem; }
.tip-box {
  background-color: #f0fdf4; border-left: 4px solid #22c55e;
  border-radius: 4px; padding: 0.75rem 1rem;
  margin: 1.25rem 0; font-size: 0.93rem;
}
</style>

## Overview

A **confidence interval** gives a range of plausible values for an unknown
population parameter. A parametric CI uses a mathematical formula based on the
sampling distribution of the estimator. Every CI in this guide takes the form:

$$\text{point estimate} \pm \text{critical value} \times SE$$

The **critical value** comes from a normal or t-distribution and is found using
`cnorm(C)` (normal) or `ct(C, df)` (t) for a confidence level C. The
**standard error (SE)** measures the typical variability of the estimator.

<div class="guide-nav">
**In this article:**

- [One proportion](#one-proportion)
- [One mean](#one-mean)
- [Difference of two means](#difference-of-two-means)
- [Regression slope](#regression-slope)
- [Quick reference](#quick-reference)

**Also in Class Summaries:**

- [Randomization Confidence Intervals](class-summary-randomization-cis.html)
- [Parametric Hypothesis Tests](transitioning-parametric-tests.html)
</div>

---

## One Proportion

**When to use:** You have a single categorical variable with two outcomes and
want to estimate the population proportion π.

**Conditions:** $n\hat{p} \geq 10$ and $n(1-\hat{p}) \geq 10$

**Formula:**
$$\hat{p} \pm z^* \cdot \hat{SE}, \qquad \hat{SE} = \sqrt{\frac{\hat{p}(1-\hat{p})}{n}}$$

The critical value $z^*$ = `cnorm(C)` for confidence level C.

**Example (class 20):** A survey asked 2625 people if they agreed that
"There is only one true love for each person." 1812 of the respondents
disagreed. Compute a 90% confidence interval for the proportion who disagreed.

```{r one_prop_ci}
# Data
n     <- 2625
x     <- 1812   # number who disagreed
p_hat <- x / n
p_hat

# Check conditions
n * p_hat        # must be >= 10
n * (1 - p_hat)  # must be >= 10

# Standard error
SE <- sqrt(p_hat * (1 - p_hat) / n)
SE

# Critical value for a 90% CI
z_star <- cnorm(0.90)
z_star

# Confidence interval
CI <- c(p_hat - z_star * SE, p_hat + z_star * SE)
CI
```

We are 90% confident that the true proportion of people who disagree that
there is only one true love for each person is between
`r round(CI[1], 3)` and `r round(CI[2], 3)`.

---

## One Mean

**When to use:** You have a single quantitative variable and want to estimate
the population mean μ.

**Conditions:** $n \geq 30$, or the population is approximately normal.

**Formula:**
$$\bar{x} \pm t^* \cdot SE, \qquad SE = \frac{s}{\sqrt{n}}, \qquad df = n-1$$

The critical value $t^*$ = `ct(C, df)` for confidence level C with $df = n-1$
degrees of freedom.

**Example (class 21):** A study by Loyd et al. (2013) used KittyCams to
record all outdoor hunting activity of n = 55 domestic cats. The mean number
of kills per week was 2.4 with a standard deviation of 1.51. Compute a 99%
confidence interval for the mean number of kills per week.

```{r one_mean_ci}
# Data
n     <- 55
x_bar <- 2.4
s     <- 1.51

# Standard error and degrees of freedom
SE <- s / sqrt(n)
df <- n - 1

# Critical value for a 99% CI
t_star <- ct(0.99, df)
t_star

# Confidence interval
ME <- t_star * SE
CI <- c(x_bar - ME, x_bar + ME)
CI
```

We are 99% confident that the mean number of kills per week for outdoor
hunting cats is between `r round(CI[1], 3)` and `r round(CI[2], 3)`.

<div class="tip-box">
**When to use t vs z.** For a single mean, always use the t-distribution
(`ct()`). The t-distribution has heavier tails than the normal, which
accounts for the additional uncertainty from estimating σ with s. For a
proportion, use the normal distribution (`cnorm()`).
</div>

---

## Difference of Two Means

**When to use:** You have a quantitative variable measured in two independent
groups and want to estimate the difference in population means $\mu_1 - \mu_2$.

**Conditions:** Each group has $n \geq 30$ or both populations are
approximately normal.

**Formula (Welch's — does not assume equal variances):**
$$(\bar{x}_1 - \bar{x}_2) \pm t^* \cdot SE_{diff}, \qquad
SE_{diff} = \sqrt{\frac{s_1^2}{n_1} + \frac{s_2^2}{n_2}}, \qquad
df = \min(n_1-1,\, n_2-1)$$

**Example (class 22):** A study by Nierenberg et al. (1989) recorded dietary
fiber intake. Compute a 95% confidence interval for the difference in mean
daily fiber intake (grams) between males and females.

```{r two_means_ci}
# Summary statistics (simulated to match class 22 Nutrition Study structure)
set.seed(7823)
fiber_males   <- round(rnorm(155, mean = 23.5, sd = 14.2))
fiber_females <- round(rnorm(662, mean = 20.1, sd = 11.8))

# Group statistics
n1    <- length(fiber_males);    x_bar1 <- mean(fiber_males);    s1 <- sd(fiber_males)
n2    <- length(fiber_females);  x_bar2 <- mean(fiber_females);  s2 <- sd(fiber_females)

cat("Males:   n =", n1, "  mean =", round(x_bar1, 2), "  sd =", round(s1, 2), "\n")
cat("Females: n =", n2, "  mean =", round(x_bar2, 2), "  sd =", round(s2, 2), "\n")

# Standard error of the difference
SE_diff <- sqrt((s1^2 / n1) + (s2^2 / n2))

# Degrees of freedom (conservative: smaller of n1-1 and n2-1)
df <- min(n1 - 1, n2 - 1)

# Critical value for a 95% CI
t_star <- ct(0.95, df)

# Confidence interval for the difference (males - females)
obs_diff <- x_bar1 - x_bar2
ME       <- t_star * SE_diff
CI       <- c(obs_diff - ME, obs_diff + ME)
CI
```

The 95% CI includes zero, which suggests we do not have strong evidence of a
difference in mean fiber intake between males and females.

---

## Regression Slope

**When to use:** Two quantitative variables; estimate how much y changes on
average for a one-unit increase in x.

**Formula:** The parametric CI for β₁ is computed directly from the fitted
regression model using `confint()`, which uses `qt()` internally with
$df = n - 2$.

**Example (class 25):** Cigarette consumption per capita and lung cancer rates
across 44 U.S. states. Compute a 90% confidence interval for the slope.

```{r reg_slope_ci}
set.seed(2947)
cigs_per_capita <- runif(44, min = 10, max = 45)
cancer_rate     <- 2 + 0.005 * cigs_per_capita * 1000 + rnorm(44, sd = 5)

# Fit the linear model
lung_lm <- lm(cancer_rate ~ cigs_per_capita)

# Point estimate of the slope
slope <- coef(lung_lm)["cigs_per_capita"]
slope

# 90% CI for all coefficients (matches class 25)
confint(lung_lm, level = 0.90)
```

We are 90% confident that for each additional hundred cigarettes smoked per
capita, the lung cancer rate increases by between
`r round(confint(lung_lm, level=0.90)["cigs_per_capita", 1], 4)` and
`r round(confint(lung_lm, level=0.90)["cigs_per_capita", 2], 4)` cases per
100,000 people.

<div class="tip-box">
**Manual calculation:** You can also compute the CI by hand using the
regression output from `summary(lung_lm)`. The SE of the slope is in the
`Std. Error` column, and the critical value is `qt(1 - (1-C)/2, df = n-2)`.
</div>

---

## Quick Reference

| Parameter | Formula | Critical value | SDS1000 function |
|---|---|---|---|
| One proportion π | $\hat{p} \pm z^* \sqrt{\hat{p}(1-\hat{p})/n}$ | `cnorm(C)` | `cnorm()` |
| One mean μ | $\bar{x} \pm t^* \cdot s/\sqrt{n}$ | `ct(C, n-1)` | `ct()` |
| Difference μ₁ − μ₂ | $(\bar{x}_1-\bar{x}_2) \pm t^* \cdot SE_{diff}$ | `ct(C, min(n₁,n₂)-1)` | `ct()` |
| Regression slope β₁ | from `confint(lm(...), level = C)` | `qt(1-(1-C)/2, n-2)` | — |
