---
title: "Randomization (Bootstrap) Confidence Intervals"
description: |
  A reference guide covering every bootstrap confidence interval from S&DS 1000:
  one proportion, one mean, difference of two means, and regression slope.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Randomization (Bootstrap) Confidence Intervals}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6, fig.height = 3.5)
library(SDS1000)

# Shadow do_it() to avoid lazyeval rendering issue in knitr.
# Behaviour is identical to the original.
do_it <- function(n) structure(list(num_times = n), class = "do_it_knitr")
`*.do_it_knitr` <- function(e1, e2) {
  expr <- match.call()[[3]]
  replicate(e1$num_times, eval(expr, envir = parent.frame()))
}
```



## Overview

A **bootstrap confidence interval** estimates the uncertainty of a statistic by
repeatedly resampling the observed data *with replacement*. Because it relies on
the data itself rather than a mathematical formula, it works for almost any
statistic and requires very few distributional assumptions.

Every bootstrap CI in this guide follows the same three steps:

1. **Create a bootstrap distribution** — use `do_it()` to resample with replacement
   many times, computing the statistic each time.
2. **Calculate the bootstrap SE** — `SE_boot <- sd(boot_dist)`.
3. **Construct the CI** — `obs_stat ± cnorm(C) × SE_boot`.

<div class="guide-nav">
**In this article:**

- [One proportion](#one-proportion)
- [One mean](#one-mean)
- [Difference of two means](#difference-of-two-means)
- [Correlation](#correlation)
- [Regression slope](#regression-slope)
- [Quick reference](#quick-reference)

<br>

**Also in Class Summaries:**

- [Randomization Hypothesis Tests](class-summary-randomization-tests.html)
- [Parametric Hypothesis Tests](class-summary-parametric-tests.html)
- [Parametric Confidence Intervals](class-summary-parametric-cis.html)
</div>

<div class="tip-box">
**Key SDS1000 functions.** `do_it(n) * { expr }` repeats an expression `n`
times. `sample(data, n, replace = TRUE)` resamples with replacement.
`get_proportion(v, "cat")` computes a proportion. `resample_pairs(v1, v2)`
resamples two paired vectors together. `cnorm(C)` gives the normal critical
value for confidence level C.
</div>

---

## One Proportion

**When to use:** A single categorical variable with two outcomes; estimate the
population proportion π.

**Example (class 11):** A survey asked 2625 people whether they agreed with
"There is only one true love for each person." 1812 disagreed. Compute a 90%
bootstrap confidence interval for the proportion who disagreed.

### Step 1: Compute the observed statistic

```{r one_prop_obs}
believes_in_true_love <- c(rep("FALSE", 1812), rep("TRUE", 2625 - 1812))

obs_prop <- get_proportion(believes_in_true_love, "FALSE")
obs_prop
```

### Step 2: Create the bootstrap distribution

Resample the data with replacement 10,000 times, computing the proportion
each time:

```{r one_prop_boot}
set.seed(2292)

n <- length(believes_in_true_love)

boot_dist <- do_it(10000) * {
  boot_sample <- sample(believes_in_true_love, n, replace = TRUE)
  get_proportion(boot_sample, "FALSE")
}

hist(boot_dist, breaks = 30,
     main = "Bootstrap Distribution — Proportion Who Disagreed",
     xlab = "Proportion",
     col = "steelblue", border = "white")
```

### Step 3: Compute the SE and construct the CI

```{r one_prop_ci}
SE_boot <- sd(boot_dist)

# 90% CI
CI <- c(obs_prop - cnorm(0.90) * SE_boot,
        obs_prop + cnorm(0.90) * SE_boot)
CI
```

We are 90% confident that the true proportion of people who disagree that
there is only one true love for each person is between
`r round(CI[1], 3)` and `r round(CI[2], 3)`.

---

## One Mean

**When to use:** A single quantitative variable; estimate the population mean μ.
The bootstrap is especially useful when the data is skewed or the sample is
small, since it does not require a normality assumption.

**Example (class 11/12):** A sample of 50 body temperatures from Mackowiak
et al. (1992). Compute a 95% bootstrap confidence interval for the mean
body temperature.

### Step 1: Compute the observed statistic

```{r one_mean_obs}
set.seed(9910)
# Simulated to match class 11 Lock5Data body temperature sample
body_temps <- rnorm(50, mean = 98.25, sd = 0.73)

obs_mean <- mean(body_temps)
obs_mean
```

### Step 2: Create the bootstrap distribution

```{r one_mean_boot}
set.seed(4471)

n <- length(body_temps)

boot_dist <- do_it(10000) * {
  boot_sample <- sample(body_temps, n, replace = TRUE)
  mean(boot_sample)
}

hist(boot_dist, breaks = 80,
     main = "Bootstrap Distribution — Mean Body Temperature",
     xlab = "Mean Temperature (°F)",
     col = "steelblue", border = "white")
```

### Step 3: Compute the SE and construct the CI

```{r one_mean_ci}
SE_boot <- sd(boot_dist)

# 95% CI
CI_95 <- c(obs_mean - cnorm(0.95) * SE_boot,
           obs_mean + cnorm(0.95) * SE_boot)

# 80% CI
CI_80 <- c(obs_mean - cnorm(0.80) * SE_boot,
           obs_mean + cnorm(0.80) * SE_boot)

# 99% CI
CI_99 <- c(obs_mean - cnorm(0.99) * SE_boot,
           obs_mean + cnorm(0.99) * SE_boot)

CI_80
CI_95
CI_99
```

Notice how wider confidence levels produce wider intervals. We are 95% confident
that the mean human body temperature is between `r round(CI_95[1], 2)`°F and
`r round(CI_95[2], 2)`°F.

<div class="tip-box">
**Wider CI = more confident.** A 99% CI is wider than a 95% CI, which is wider
than an 80% CI. There is always a trade-off: greater confidence requires
sacrificing precision.
</div>

---

## Difference of Two Means

**When to use:** A quantitative variable measured in two independent groups;
estimate $\mu_1 - \mu_2$. Resample each group separately with replacement,
keeping group sizes fixed.

**Example:** Using the calcium supplement study from class 17, compute a 95%
bootstrap confidence interval for the difference in mean blood pressure
decrease (treatment − control).

### Step 1: Compute the observed statistic

```{r two_means_obs}
treat   <- c( 7, -4, 18, 17, -3, -5,  1, 10, 11, -2)
control <- c(-1, 12, -1, -3,  3, -5,  5,  2, -11, -1, -3)

obs_diff <- mean(treat) - mean(control)
obs_diff
```

### Step 2: Create the bootstrap distribution

Resample each group independently with replacement:

```{r two_means_boot}
set.seed(6174)

n_treat   <- length(treat)
n_control <- length(control)


boot_dist <- do_it(10000) * {
  boot_treat   <- sample(treat,   n_treat,   replace = TRUE)
  boot_control <- sample(control, n_control, replace = TRUE)
  mean(boot_treat) - mean(boot_control)
}

hist(boot_dist, breaks = 80,
     main = "Bootstrap Distribution — Difference in Means",
     xlab = "Difference in mean blood pressure decrease (mmHg)",
     col = "steelblue", border = "white")
```

### Step 3: Compute the SE and construct the CI

```{r two_means_ci}
SE_boot <- sd(boot_dist)

CI <- c(obs_diff - cnorm(0.95) * SE_boot,
        obs_diff + cnorm(0.95) * SE_boot)
CI
```

The 95% CI includes zero, consistent with the hypothesis test result. However,
the interval also extends well above zero, meaning a meaningful calcium effect
remains plausible.

---

## Correlation

**When to use:** Two paired quantitative variables; estimate the population
correlation ρ between them. Because x and y are paired, we must resample
both together using `resample_pairs()` so that every resampled (x, y) pair
stays matched — this preserves the pairing structure.

**Example (class 18):** Sugar content and calorie content for a sample of 77
breakfast cereals. Compute a 95% bootstrap confidence interval for the
correlation between sugar and calories.

### Step 1: Compute the observed statistic

```{r corr_obs}
set.seed(1123)
n        <- 77
sugar    <- runif(n, 0, 15)   # fake simulated data
calories <- 90 + 3.5 * sugar + rnorm(n, sd = 12)

obs_cor <- cor(sugar, calories)
obs_cor
```

### Step 2: Create the bootstrap distribution

Use `resample_pairs()` to resample both variables together, then recompute
`cor()` each time:

```{r corr_boot}
set.seed(8833)

boot_dist <- do_it(10000) * {
  resampled          <- resample_pairs(sugar, calories)
  resampled_sugar    <- resampled$vector1
  resampled_calories <- resampled$vector2
  cor(resampled_sugar, resampled_calories)
}

hist(boot_dist, breaks = 80,
     main = "Bootstrap Distribution — Correlation",
     xlab = "Correlation",
     col = "steelblue", border = "white")
```

### Step 3: Compute the SE and construct the CI

```{r corr_ci}
SE_boot <- sd(boot_dist)

CI <- c(obs_cor - cnorm(0.95) * SE_boot,
        obs_cor + cnorm(0.95) * SE_boot)
CI
```

We are 95% confident that the true correlation between sugar and calorie
content in cereals is between `r round(CI[1], 3)` and `r round(CI[2], 3)`.
Since the interval lies entirely above zero, the data provide good evidence
of a positive association.

<div class="tip-box">
**Why `resample_pairs()` and not `sample(..., replace = TRUE)`?** For paired
data, we must resample observations (rows) together to preserve the
relationship between x and y. Resampling x and y independently would
destroy the very correlation we are trying to measure.
</div>

---

## Regression Slope

**When to use:** Two paired quantitative variables; estimate the slope β₁ of
a simple linear regression model. Because x and y are paired, resample both
together using `resample_pairs()` — exactly the same approach as for the regression slope. 

**Example (class 25):** Cigarette consumption per capita and lung cancer rates
across 44 U.S. states. Compute a 95% bootstrap confidence interval for the
regression slope.

### Step 1: Compute the observed statistic

```{r reg_obs}
set.seed(2947)
cigs_per_capita <- runif(44, min = 10, max = 45)  # create fake simulated data
cancer_rate     <- 2 + 0.005 * cigs_per_capita * 1000 + rnorm(44, sd = 5)

lung_lm  <- lm(cancer_rate ~ cigs_per_capita)
obs_slope <- coef(lung_lm)["cigs_per_capita"]
obs_slope
```

### Step 2: Create the bootstrap distribution

Use `resample_pairs()` to resample both variables together, keeping pairs
intact:

```{r reg_boot}
set.seed(3109)

boot_dist <- do_it(10000) * {
  resampled       <- resample_pairs(cigs_per_capita, cancer_rate)
  resampled_cigs  <- resampled$vector1
  resampled_cancer <- resampled$vector2
  coef(lm(resampled_cancer ~ resampled_cigs))["resampled_cigs"]
}

hist(boot_dist, breaks = 80,
     main = "Bootstrap Distribution — Regression Slope",
     xlab = "Slope",
     col = "steelblue", border = "white")
```

### Step 3: Compute the SE and construct the CI

```{r reg_ci}
SE_boot <- sd(boot_dist)

CI <- c(obs_slope - cnorm(0.95) * SE_boot,
        obs_slope + cnorm(0.95) * SE_boot)
CI
```

We are 95% confident that for each additional hundred cigarettes smoked per
capita, the lung cancer rate increases by between `r round(CI[1], 4)` and
`r round(CI[2], 4)` cases per 100,000 people.


---

## Quick Reference

| Parameter | Resample method | CI formula |
|---|---|---|
| One proportion | `sample(data, n, replace=TRUE)` → `get_proportion()` | `obs ± cnorm(C) * sd(boot_dist)` |
| One mean | `sample(data, n, replace=TRUE)` → `mean()` | `obs ± cnorm(C) * sd(boot_dist)` |
| Diff. of two means | `sample(group, n, replace=TRUE)` for each group | `obs ± cnorm(C) * sd(boot_dist)` |
| Correlation | `resample_pairs(x, y)` → `cor()` | `obs ± cnorm(C) * sd(boot_dist)` |
| Regression slope | `resample_pairs(x, y)` → `coef(lm(...))` | `obs ± cnorm(C) * sd(boot_dist)` |
