---
title: "Randomization Hypothesis Tests"
description: |
  A reference guide covering every randomization (simulation-based) hypothesis
  test from S&DS 1000: one proportion, multiple proportions, two means,
  more than two means, correlation, and simple linear regression.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Randomization Hypothesis Tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE,
                      fig.width = 6, fig.height = 3.5)
library(SDS1000)

# do_it() uses lazyeval::f_capture() which has issues in knitr rendering.
# Shadow do_it() with a version that returns a new class, then register
# a knitr-safe * method for it — behaviour is identical to the original.
do_it <- function(n) structure(list(num_times = n), class = "do_it_knitr")
`*.do_it_knitr` <- function(e1, e2) {
  expr <- match.call()[[3]]
  replicate(e1$num_times, eval(expr, envir = parent.frame()))
}
```

<style>
h2 {
  margin-top:     3.5rem;
  padding-top:    0.6rem;
  padding-bottom: 0.4rem;
  border-bottom:  1px solid #dee2e6;
  color: #1a3a5c;
  font-size: 1.45rem;
}

h3 {
  margin-top: 2rem;
  color: #2c5282;
  font-size: 1.15rem;
}

h4 {
  margin-top: 1.5rem;
  color: #2c5282;
  font-size: 1.05rem;
}

hr {
  margin:     3rem 0 2rem 0;
  border:     none;
  border-top: 2px solid #e2e8f0;
}

table {
  margin:    1.25rem 0 1.75rem 0;
  font-size: 0.93rem;
  width:     100%;
}

table thead tr {
  background-color: #dbeafe;
  color: #1e3a5f;
}

table tbody tr:nth-child(even) {
  background-color: #f8fafc;
}

.guide-nav {
  background-color: #f0f7ff;
  border:           1px solid #bfdbfe;
  border-radius:    6px;
  padding:          1.1rem 1.4rem;
  margin-bottom:    2rem;
  font-size:        0.95rem;
}

.guide-nav ul  { margin-bottom: 0; padding-left: 1.2rem; }
.guide-nav li  { margin-bottom: 0.3rem; }

.tip-box {
  background-color: #f0fdf4;
  border-left:      4px solid #22c55e;
  border-radius:    4px;
  padding:          0.75rem 1rem;
  margin:           1.25rem 0;
  font-size:        0.93rem;
}
</style>

## Overview

This guide covers every randomization (simulation-based) hypothesis test from
S&DS 1000. Each test follows the same five steps used in class:

1. **State** the null and alternative hypotheses
2. **Calculate** the observed test statistic
3. **Create** a null distribution by simulation
4. **Calculate** the p-value
5. **Make a decision**

The core idea is the same for every test: if the null hypothesis were true, how
extreme would our observed result be? We answer this by repeatedly simulating
data under the null and seeing where our observed statistic falls.

<div class="guide-nav">
**In this article:**

- [One proportion](#one-proportion)
- [More than two proportions](#more-than-two-proportions)
- [Two means](#two-means)
- [More than two means](#more-than-two-means)
- [Correlation](#correlation)
- [Simple linear regression](#simple-linear-regression)
- [Quick reference table](#quick-reference)

**Also in Class Summaries:**

- [Parametric Hypothesis Tests](transitioning-parametric-tests.html)
</div>

<div class="tip-box">
**Key SDS1000 functions used in this guide.**
`do_it(n) * { expr }` repeats an expression `n` times and collects the
results. `rflip(n, p)` simulates `n` coin flips with probability `p`.
`rroll(n, prob)` simulates rolling a die `n` times. `shuffle(x)` randomly
permutes a vector. `ptail(obs, null, lower.tail)` computes the p-value from
a null distribution.
</div>

---

## One Proportion

**When to use:** One categorical variable with two outcomes; test whether the
population proportion π equals a specific value π₀. The null distribution is
created by simulating coin flips under the assumption that H₀ is true using
`rflip()`.

**Example (class 12):** Doris the dolphin and Buzz the dolphin were tested to
see whether Buzz could identify which lever Doris was pointing to. In 16
trials, Buzz got the correct answer 15 times. Is there evidence that Buzz is
doing better than random chance?

### Step 1: State the hypotheses

$$H_0: \pi = 0.50 \qquad H_A: \pi > 0.50$$

### Step 2: Calculate the observed statistic

```{r one_prop_obs}
n_trials <- 16
obs_stat <- 15   # number of correct answers
obs_stat
```

### Step 3: Create the null distribution

Under H₀, each trial is like a fair coin flip. Use `rflip()` to simulate
10,000 experiments:

```{r one_prop_null}
set.seed(5731)

null_dist <- do_it(10000) * {
  rflip(n_trials, 0.50)
}

hist(null_dist, breaks = 50,
     main = "Null Distribution — Number of Correct Answers",
     xlab = "Number Correct (out of 16)",
     col = "steelblue", border = "white")
abline(v = obs_stat, col = "red", lwd = 2)
```

### Step 4: Calculate the p-value

```{r one_prop_pval}
p_value <- ptail(obs_stat, null_dist, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 4)` < 0.05, we **reject** H₀ and conclude that
Buzz appears to be doing better than random chance.

---

**Second example (class 13):** A lie detector correctly identified 31 out of
48 participants as lying. Is this evidence the detector is more than 60%
accurate?

$$H_0: \pi = 0.60 \qquad H_A: \pi > 0.60$$

```{r one_prop_lie}
set.seed(9021)

n        <- 48
obs_phat <- 31 / 48

null_dist <- do_it(10000) * {
  rflip(n, 0.60) / n
}

hist(null_dist, breaks = 60,
     main = "Null Distribution — Proportion Correct",
     xlab = "Proportion Correct",
     col = "steelblue", border = "white")
abline(v = obs_phat, col = "red", lwd = 2)

p_value <- ptail(obs_phat, null_dist, lower.tail = FALSE)
p_value
```

Since p = `r round(p_value, 3)` > 0.05, we **fail to reject** H₀. There is
not sufficient evidence that the lie detector exceeds 60% accuracy.

---

## More Than Two Proportions

**When to use:** One categorical variable with three or more outcomes; test
whether the population proportions match a set of expected values. The
observed statistic is the chi-squared statistic; the null distribution is
created by simulating random samples from the expected distribution using
`rroll()`.

**Example (class 23):** A Yale professor examined the birth months of 198
randomly selected Yale students. Are students equally likely to be born in any
month?

### Step 1: State the hypotheses

$$H_0: \pi_{Jan} = \pi_{Feb} = \cdots = \pi_{Dec} = \tfrac{1}{12}$$
$$H_A: \text{at least one monthly proportion differs from } \tfrac{1}{12}$$

### Step 2: Calculate the observed statistic

```{r multi_prop_obs}
observed_counts <- c(14, 11, 21, 17, 15, 13, 19, 16, 18, 22, 14, 18)
names(observed_counts) <- month.abb

expected_counts <- rep(sum(observed_counts) / 12, 12)

obs_stat <- sum((observed_counts - expected_counts)^2 / expected_counts)
obs_stat
```

### Step 3: Create the null distribution

Under H₀, birth months are equally likely. Use `rroll()` to simulate 10,000
samples of 198 students from a uniform distribution across 12 months:

```{r multi_prop_null}
set.seed(3341)

n_students      <- sum(observed_counts)
expected_counts <- expected_counts

null_dist <- do_it(10000) * {
  sim_counts <- rroll(n_students, prob = rep(1/12, 12))
  sum((sim_counts - expected_counts)^2 / expected_counts)
}

hist(null_dist, breaks = 60,
     main = "Null Distribution — Chi-squared Statistic",
     xlab = "Chi-squared value",
     col = "steelblue", border = "white")
abline(v = obs_stat, col = "red", lwd = 2)
```

### Step 4: Calculate the p-value

```{r multi_prop_pval}
p_value <- ptail(obs_stat, null_dist, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 3)` >> 0.05, we **fail to reject** H₀. There is
no evidence that Yale students' birth months deviate from a uniform
distribution.

---

## Two Means

**When to use:** A quantitative variable measured in two independent groups;
test whether the two group means are equal. The null distribution is created
by using `shuffle()` to randomly reassign group labels — under H₀, group
membership doesn't matter.

**Observed statistic:** Difference in group means: $\bar{x}_1 - \bar{x}_2$

**Example (class 17):** A randomized controlled trial (Lyle et al., 1987)
gave calcium supplements to 10 men (treatment) and a placebo to 11 men
(control). The outcome is the decrease in blood pressure. Is there evidence
that calcium produces a greater decrease?

### Step 1: State the hypotheses

$$H_0: \mu_\text{treat} = \mu_\text{control} \qquad H_A: \mu_\text{treat} > \mu_\text{control}$$

### Step 2: Calculate the observed statistic

```{r two_means_obs}
treat   <- c( 7, -4, 18, 17, -3, -5,  1, 10, 11, -2)
control <- c(-1, 12, -1, -3,  3, -5,  5,  2, -11, -1, -3)

obs_stat <- mean(treat) - mean(control)
obs_stat
```

### Step 3: Create the null distribution

Combine both groups into one vector and use `shuffle()` to randomly reassign
group labels 10,000 times:

```{r two_means_null}
set.seed(6174)

combined_data <- c(treat, control)

null_dist <- do_it(10000) * {
  shuff         <- shuffle(combined_data)
  shuff_treat   <- shuff[1:10]
  shuff_control <- shuff[11:21]
  mean(shuff_treat) - mean(shuff_control)
}

hist(null_dist, breaks = 100,
     main = "Null Distribution — Difference in Means",
     xlab = "Difference in mean blood pressure decrease (mmHg)",
     col = "steelblue", border = "white")
abline(v = obs_stat, col = "red", lwd = 2)
```

### Step 4: Calculate the p-value

```{r two_means_pval}
p_value <- ptail(obs_stat, null_dist, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 3)` is just above 0.05, we **fail to reject** H₀
at the α = 0.05 level. The evidence for a calcium effect is suggestive but not
conclusive.

---

## More Than Two Means

**When to use:** A quantitative variable measured across three or more
independent groups; test whether all group means are equal. The observed
statistic is the **mean absolute deviation (MAD)** of group means — the
average of all pairwise absolute differences between group means. The null
distribution is created by using `shuffle()` on the group labels.

**Example (class 17/18):** Hope College students were timed completing a
Sudoku-like puzzle, grouped into four majors. Is there a difference in mean
completion time?

### Step 1: State the hypotheses

$$H_0: \mu_{as} = \mu_{ns} = \mu_{ss} = \mu_{ah}$$
$$H_A: \text{at least one group mean differs}$$

### Step 2: Calculate the observed statistic

```{r multi_means_obs}
set.seed(4455)
completion_times <- c(rnorm(10, mean = 22, sd = 5),
                      rnorm(10, mean = 20, sd = 5),
                      rnorm(10, mean = 26, sd = 5),
                      rnorm(10, mean = 21, sd = 5))
majors <- rep(c("Applied Sci", "Natural Sci", "Social Sci", "Arts/Hum"),
              each = 10)

boxplot(completion_times ~ majors,
        main  = "Completion Time by Major",
        ylab  = "Time (seconds)",
        col   = c("lightblue", "lightgreen", "lightpink", "lightyellow"))

obs_stat <- get_MAD_stat(completion_times, majors)
obs_stat
```

### Step 3: Create the null distribution

Use `shuffle()` on the group labels to break any real association:

```{r multi_means_null}
set.seed(8812)

completion_times <- completion_times
majors           <- majors

null_dist <- do_it(10000) * {
  shuffled_majors <- shuffle(majors)
  get_MAD_stat(completion_times, shuffled_majors)
}

hist(null_dist, breaks = 100,
     main = "Null Distribution — MAD Statistic",
     xlab = "Mean absolute deviation of group means",
     col = "steelblue", border = "white")
abline(v = obs_stat, col = "red", lwd = 2)
```

### Step 4: Calculate the p-value

```{r multi_means_pval}
p_value <- ptail(obs_stat, null_dist, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision

Since p = `r round(p_value, 3)`, we `r ifelse(p_value < 0.05, "**reject**", "**fail to reject**")` H₀ at the α = 0.05 level.

<div class="tip-box">
**What is the MAD statistic?** `get_MAD_stat()` computes the mean of all
pairwise absolute differences between group means:
$$MAD = \frac{1}{\binom{k}{2}} \sum_{i < j} |\bar{x}_i - \bar{x}_j|$$
It is sensitive to any difference between group means, making it a natural
choice for a general alternative hypothesis.
</div>

---

## Correlation

**When to use:** Two quantitative variables; test whether the population
correlation ρ equals zero. The null distribution is created by using
`shuffle()` on one of the two variables — this breaks any real relationship
while preserving the marginal distributions.

**Example (class 18):** A dataset of 77 breakfast cereals recorded sugar
content and calorie content. Is there evidence of a positive correlation?

### Step 1: State the hypotheses

$$H_0: \rho = 0 \qquad H_A: \rho > 0$$

### Step 2: Calculate the observed statistic

```{r corr_obs}
set.seed(1123)
# Simulated to match class 18 cereal data structure
n        <- 77
sugar    <- runif(n, 0, 15)
calories <- 90 + 3.5 * sugar + rnorm(n, sd = 12)

plot(sugar, calories,
     main = "Calories vs Sugar in Cereals",
     xlab = "Sugar (grams)", ylab = "Calories")

obs_stat <- cor(sugar, calories)
obs_stat
```

### Step 3: Create the null distribution

Use `shuffle()` on the sugar values to destroy any correlation with calories:

```{r corr_null}
set.seed(5671)

sugar    <- sugar
calories <- calories

null_dist <- do_it(10000) * {
  cor(shuffle(sugar), calories)
}

hist(null_dist, breaks = 80,
     main = "Null Distribution — Correlation",
     xlab = "Correlation",
     col = "steelblue", border = "white")
abline(v = obs_stat, col = "red", lwd = 2)
```

### Step 4: Calculate the p-value

```{r corr_pval}
p_value <- ptail(obs_stat, null_dist, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision

Since p ≈ `r round(p_value, 4)`, we **reject** H₀ and conclude that there is
evidence of a positive correlation between sugar and calorie content in cereals.

---

## Simple Linear Regression

**When to use:** Two quantitative variables where you want to test whether the
slope of a linear regression model is significantly different from zero. The
null distribution is created by using `shuffle()` on the response variable —
under H₀ (no relationship), the pairing between x and y is arbitrary.

**Example (class 25):** Data on cigarette consumption per capita and lung
cancer rates across U.S. states. Is there evidence of a positive slope?

### Step 1: State the hypotheses

$$H_0: \beta_1 = 0 \qquad H_A: \beta_1 > 0$$

### Step 2: Calculate the observed statistic

```{r reg_obs}
set.seed(2947)
cigs_per_capita <- runif(44, min = 10, max = 45)
cancer_rate     <- 2 + 0.005 * cigs_per_capita * 1000 + rnorm(44, sd = 5)

plot(cigs_per_capita, cancer_rate,
     xlab = "Cigarettes per Capita (hundreds)",
     ylab = "Lung Cancer Rate per 100,000",
     main = "Lung Cancer Rate vs. Cigarette Consumption")

lung_lm   <- lm(cancer_rate ~ cigs_per_capita)
obs_slope <- coef(lung_lm)["cigs_per_capita"]
obs_slope
```

### Step 3: Create the null distribution

Use `shuffle()` on the cancer rate values — this breaks the association with
`cigs_per_capita` while keeping both variables' distributions intact:

```{r reg_null}
set.seed(3318)

cancer_rate     <- cancer_rate
cigs_per_capita <- cigs_per_capita

null_dist <- do_it(10000) * {
  shuffled_cancer <- shuffle(cancer_rate)
  null_lm <- lm(shuffled_cancer ~ cigs_per_capita)
  coef(null_lm)["cigs_per_capita"]
}

hist(null_dist, breaks = 80,
     main = "Null Distribution — Regression Slope",
     xlab = "Slope",
     col = "steelblue", border = "white")
abline(v = obs_slope, col = "red", lwd = 2)
```

### Step 4: Calculate the p-value

```{r reg_pval}
p_value <- ptail(obs_slope, null_dist, lower.tail = FALSE)
p_value
```

### Step 5: Make a decision

Since p ≈ `r round(p_value, 4)`, we **reject** H₀ and conclude there is
evidence of a positive relationship between cigarette consumption and lung
cancer rate.

---

## Quick Reference

| Test | Observed statistic | How to simulate the null | p-value |
|---|---|---|---|
| One proportion | Count or proportion | `do_it(n) * { rflip(n, pi_0) }` | `ptail(obs, null, lower.tail = FALSE)` |
| Multiple proportions | Chi-squared: $\sum(O-E)^2/E$ | `do_it(n) * { rroll(n, expected_p) }` | `ptail(obs, null, lower.tail = FALSE)` |
| Two means | $\bar{x}_1 - \bar{x}_2$ | `do_it(n) * { shuff <- shuffle(combined); ... }` | `ptail(obs, null, lower.tail = FALSE)` |
| More than two means | MAD: `get_MAD_stat()` | `do_it(n) * { get_MAD_stat(y, shuffle(groups)) }` | `ptail(obs, null, lower.tail = FALSE)` |
| Correlation | $r =$ `cor(x, y)` | `do_it(n) * { cor(shuffle(x), y) }` | `ptail(obs, null, lower.tail = FALSE)` |
| Regression slope | $\hat{\beta}_1$ from `lm()` | `do_it(n) * { coef(lm(shuffle(y) ~ x))[2] }` | `ptail(obs, null, lower.tail = FALSE)` |
