---
title: "Transitioning to Base R: Simulation and Data Summaries"
description: |
  Part 1 of 2. Covers do_it(), get_proportion(), rflip(), rroll(),
  shuffle(), and resample_pairs() and their base R equivalents.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Transitioning to Base R: Simulation and Data Summaries}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
library(SDS1000)
```



## Overview

The SDS1000 package was designed to make it easier to learn the core ideas of
statistics in R. Now that you have finished S&DS 1000, you are ready to analyze
data using standard R tools that are available to every R user — no special
package required.

This is **Part 1** of a two-part guide. It covers the simulation and data
summary functions. Part 2 covers inference and visualization.

<div class="guide-nav">
**In this article:**

- [Repeating a process — `do_it()` → `replicate()`](#repeating-a-process-do_it-replicate)
- [Calculating a proportion — `get_proportion()` → `mean()` or `table()`](#calculating-a-proportion-get_proportion-mean-or-table)
- [Simulating coin flips and die rolls — `rflip()` and `rroll()`](#simulating-coin-flips-and-die-rolls-rflip-and-rroll)
- [Permutation tests and bootstrapping — `shuffle()` and `resample_pairs()`](#permutation-tests-and-bootstrapping-shuffle-and-resample_pairs)

**Also in this series:**

- [Part 2: Inference and Visualization](transitioning-inference.html) —
  `ptail()`, `cnorm()`, `ct()`, distribution plots, and a complete worked example
- [Parametric Hypothesis Tests](class-summary-parametric-tests.html) —
  t-tests, chi-squared, ANOVA, correlation, and regression
- [Cheat Sheet](transitioning-cheatsheet.html) — printable one-page summary
</div>

---

## Repeating a process: `do_it()` → `replicate()`

### What `do_it()` does

In S&DS 1000, you used `do_it(n)` to repeat a block of code `n` times and
collect the results into a vector. For example, to build a **sampling
distribution** of the proportion of red sprinkles from samples of size 100:

```{r do_it_example, eval=FALSE}
# SDS1000 version
sampling_distribution <- do_it(10000) * {
  get_proportion(rsprinkles(100), "red")
}
```

You also used it to build a **bootstrap distribution**:

```{r do_it_bootstrap, eval=FALSE}
# SDS1000 version
bootstrap_dist <- do_it(10000) * {
  boot_sample <- sample(price_sample, length(price_sample), replace = TRUE)
  mean(boot_sample)
}
```

And to build a **null distribution** by shuffling:

```{r do_it_null, eval=FALSE}
# SDS1000 version
null_distribution <- do_it(10000) * {
  cor(shuffle(sugar), calories)
}
```

### The base R equivalent: `replicate()`

The base R function `replicate(n, expr)` does exactly the same thing: it
evaluates the expression `expr` a total of `n` times and returns the results
as a vector.

```{r replicate_pattern, eval=FALSE}
results <- replicate(n, {
  # your code here — the result of the last line is collected
})
```

Here are the same three examples rewritten with `replicate()`:

**Sampling distribution:**

```{r replicate_sampling, eval=FALSE}
sampling_distribution <- replicate(10000, {
  mean(rsprinkles(100) == "red")   # get_proportion() → mean(v == "cat")
})
```

**Bootstrap distribution:**

```{r replicate_bootstrap, eval=FALSE}
bootstrap_dist <- replicate(10000, {
  boot_sample <- sample(price_sample, length(price_sample), replace = TRUE)
  mean(boot_sample)
})
```

**Null distribution (shuffling):**

```{r replicate_null, eval=FALSE}
null_distribution <- replicate(10000, {
   cor(sample(sugar), calories)      # sample() with no replace = shuffle
})
```

### The `for` loop alternative

You can also use a `for` loop, which makes the mechanics more explicit:

```{r for_loop, eval=FALSE}
n_repetitions <- 10000
sampling_distribution <- numeric(n_repetitions)  # pre-allocate an empty vector

for (i in 1:n_repetitions) {
  sampling_distribution[i] <- mean(rsprinkles(100) == "red")
}
```

The `replicate()` approach is more concise; the `for` loop makes the indexing
transparent. Both produce identical results.

### A working example

```{r working_example}
set.seed(1783)

sampling_dist <- replicate(5000, {
  mean(rnorm(30, mean = 0, sd = 1))
})

hist(sampling_dist,
     main = "Sampling Distribution of the Mean (n = 30)",
     xlab = "Sample Mean",
     col = "steelblue", border = "white")
```

### Quick reference

| Task | SDS1000 | Base R |
|---|---|---|
| Repeat a process `n` times | `do_it(n) * { expr }` | `replicate(n, { expr })` |
| Repeat with explicit indexing | — | `for (i in 1:n) { result[i] <- expr }` |

---

## Calculating a proportion: `get_proportion()` → `mean()` or `table()`

### What `get_proportion()` does

`get_proportion(v, category_name)` returns the proportion of elements in
vector `v` that belong to a named category:

```{r get_proportion_example, eval=FALSE}
# SDS1000 version
p_hat <- get_proportion(one_sample, "red")
```

### Base R approach 1: `mean()` with a logical comparison (recommended)

Because `TRUE` and `FALSE` are stored as `1` and `0` in R, taking `mean()` of
a logical vector gives the proportion of `TRUE` values:

```{r mean_logical, eval=FALSE}
p_hat <- mean(one_sample == "red")
```

This form is especially useful inside `replicate()` since it fits on one line:

```{r replicate_with_mean, eval=FALSE}
sampling_distribution <- replicate(10000, {
  mean(rsprinkles(100) == "red")
})
```

### Base R approach 2: `table()` and `prop.table()`

Use these when you want to see all categories at once:

```{r table_approach}
one_sample <- rsprinkles(100)
table(one_sample)         # raw counts
```

```{r prop_table_approach}
prop.table(table(one_sample))   # proportions for every category
```

### Quick reference

| Task | SDS1000 | Base R |
|---|---|---|
| Proportion of one category | `get_proportion(v, "red")` | `mean(v == "red")` |
| Proportions of all categories | — | `prop.table(table(v))` |
| Counts of all categories | — | `table(v)` |

---

## Simulating coin flips and die rolls: `rflip()` and `rroll()`

### `rflip()` → `rbinom()`

`rflip(num_flips, prob)` simulates flipping a coin `num_flips` times and
returns the **count** of heads. Under the hood it is one call to `rbinom()`:

```{r rflip_def, eval=FALSE}
rflip(num_flips, prob)
rbinom(1, size = num_flips, prob = prob)   # equivalent
```

Simulating 100 fair coin flips (class 12 / 13 style):

```{r rflip_example}
set.seed(3341)
rbinom(1, size = 100, prob = 0.5)
```

To build a null distribution of head counts inside `replicate()`:

```{r rflip_null}
set.seed(7823)

null_dist <- replicate(10000, rbinom(1, size = 100, prob = 0.5))

hist(null_dist,
     main = "Null Distribution — Heads out of 100 Flips (p = 0.5)",
     xlab = "Number of Heads",
     col = "steelblue", border = "white")
```

To return a **proportion** instead of a count, divide by `num_flips`:

```{r rflip_proportion, eval=FALSE}
# rflip(48, prob = 0.6) / 48    ← SDS1000
rbinom(1, size = 48, prob = 0.6) / 48   # Base R
```

### `rroll()` → `rmultinom()`

`rroll(num_rolls, prob)` simulates rolling a die `num_rolls` times and
returns the **count** of each face. It wraps `rmultinom()`:

```{r rroll_def, eval=FALSE}
rroll(100)                                                    # SDS1000
as.vector(rmultinom(1, size = 100, prob = rep(1/6, 6)))      # Base R
```

`rmultinom()` returns a matrix, so `as.vector()` is needed for a plain vector.
For a 12-sided die as used in class 23:

```{r rroll_example}
set.seed(3319)
die_counts <- as.vector(rmultinom(1, size = 120, prob = rep(1/12, 12)))
names(die_counts) <- 1:12
die_counts
```

### Quick reference

| Task | SDS1000 | Base R |
|---|---|---|
| Count of heads from *n* flips | `rflip(n, prob)` | `rbinom(1, size = n, prob = prob)` |
| Proportion of heads from *n* flips | `rflip(n, prob) / n` | `rbinom(1, n, prob) / n` |
| Counts for each face of a die | `rroll(n, prob)` | `as.vector(rmultinom(1, size = n, prob = prob))` |

---

## Permutation tests and bootstrapping: `shuffle()` and `resample_pairs()`

These two functions are the building blocks of simulation-based inference.
`shuffle()` powers **permutation tests**; `resample_pairs()` powers
**bootstrap confidence intervals** for paired variables.

### `shuffle()` → `sample()`

`shuffle(v)` is a direct wrapper around `sample(v)` — both return a random
permutation of `v`:

```{r shuffle_def, eval=FALSE}
shuffle(v)   # SDS1000
sample(v)    # Base R — identical
```

In a permutation test, shuffling one variable breaks its relationship with
another, simulating the null hypothesis. For example, testing whether the correlation
between `x` and `y` is 0 at the population level we could use: 

```{r permutation_test}
set.seed(8812)

# generate fake data
n <- 100
x <- rnorm(n)
y <- 0.3 * x + rnorm(n)

obs_stat <- cor(x, y)

null_distribution <- replicate(10000, {
  cor(sample(x), y)   # sample(x) = shuffle(x)
})

p_value <- mean(abs(null_distribution) >= abs(obs_stat))

hist(null_distribution,
     main = "Null Distribution of Correlation",
     xlab = "Correlation", col = "steelblue", border = "white")
abline(v = obs_stat, col = "red", lwd = 2)
p_value
```

### `resample_pairs()` → `sample()` with `replace = TRUE`

`resample_pairs(vector1, vector2)` draws a bootstrap resample from two paired
vectors, keeping each pair together. The key is generating one set of indices
and applying them to both vectors:

```{r resample_pairs_base, eval=FALSE}
# SDS1000 version
resampled_data   <- resample_pairs(cigs_per_capita, cancer_rate)
resampled_cigs   <- resampled_data$vector1
resampled_cancer <- resampled_data$vector2

# Base R version
n                <- length(cigs_per_capita)
boot_inds        <- sample(n, replace = TRUE)
resampled_cigs   <- cigs_per_capita[boot_inds]
resampled_cancer <- cancer_rate[boot_inds]
```

A complete bootstrap for a regression slope in base R (class 25 style):

```{r bootstrap_slope}
set.seed(3047)

# generate fake data
cigs_per_capita <- runif(50, 10, 40)
cancer_rate     <- 2 + 0.8 * cigs_per_capita + rnorm(50, sd = 5)

obs_slope       <- coef(lm(cancer_rate ~ cigs_per_capita))[2]

boot_dist <- replicate(10000, {
  n         <- length(cigs_per_capita)
  boot_inds <- sample(n, replace = TRUE)
  coef(lm(cancer_rate[boot_inds] ~ cigs_per_capita[boot_inds]))[2]
})

hist(boot_dist,
     main = "Bootstrap Distribution of Regression Slope",
     xlab = "Slope", col = "steelblue", border = "white")
abline(v = obs_slope, col = "red", lwd = 2)

quantile(boot_dist, c(0.025, 0.975))
```

### Quick reference

| Task | SDS1000 | Base R |
|---|---|---|
| Randomly permute a vector | `shuffle(v)` | `sample(v)` |
| Bootstrap resample two paired vectors | `resample_pairs(v1, v2)` | `idx <- sample(n, replace=TRUE); v1[idx]; v2[idx]` |

<div class="tip-box">
**Continue to Part 2:** [Inference and Visualization](transitioning-inference.html) covers
`ptail()`, `cnorm()`, `ct()`, distribution plots, and a complete worked example
tying everything together.
</div>
