---
title: "Transitioning to Base R: Theory-Based Hypothesis Tests"
description: |
  Part 3 of the transitioning series. Covers t.test(), cor.test(), aov(),
  chisq.test(), and lm() as complete base R replacements for the SDS1000
  wrapper functions.
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Transitioning to Base R: Theory-Based Hypothesis Tests}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

<style>
h2 {
  margin-top:     3.5rem;
  padding-top:    0.6rem;
  padding-bottom: 0.4rem;
  border-bottom:  1px solid #dee2e6;
  color: #1a3a5c;
  font-size: 1.45rem;
}

h3 {
  margin-top: 2rem;
  color: #2c5282;
  font-size: 1.15rem;
}

hr {
  margin:     3rem 0 2rem 0;
  border:     none;
  border-top: 2px solid #e2e8f0;
}

table {
  margin:    1.25rem 0 1.75rem 0;
  font-size: 0.93rem;
  width:     100%;
}

table thead tr {
  background-color: #dbeafe;
  color: #1e3a5f;
}

table tbody tr:nth-child(even) {
  background-color: #f8fafc;
}

.guide-nav {
  background-color: #f0f7ff;
  border:           1px solid #bfdbfe;
  border-radius:    6px;
  padding:          1.1rem 1.4rem;
  margin-bottom:    2rem;
  font-size:        0.95rem;
}

.guide-nav ul  { margin-bottom: 0; padding-left: 1.2rem; }
.guide-nav li  { margin-bottom: 0.3rem; }

.tip-box {
  background-color: #f0fdf4;
  border-left:      4px solid #22c55e;
  border-radius:    4px;
  padding:          0.75rem 1rem;
  margin:           1.25rem 0;
  font-size:        0.93rem;
}

.warn-box {
  background-color: #fffbeb;
  border-left:      4px solid #f59e0b;
  border-radius:    4px;
  padding:          0.75rem 1rem;
  margin:           1.25rem 0;
  font-size:        0.93rem;
}
</style>

## Overview

In S&DS 1000, you built null distributions by simulation (using `do_it()` and
`shuffle()` / `rflip()`) and computed p-values by counting. That
**randomization** approach works for almost any statistic and requires very
few assumptions.

For many common situations — comparing means, testing correlations, comparing
proportions — statisticians have worked out what the null distribution looks
like *mathematically*. Base R provides dedicated functions for these
**theory-based** tests that return p-values, confidence intervals, and test
statistics in a single call.

<div class="guide-nav">
**In this article:**

- [t-tests — `t.test()`](#t-tests-comparing-means)
- [Correlation — `cor.test()`](#correlation-cor.test)
- [ANOVA — `aov()` and `summary()`](#one-way-anova-aov)
- [Chi-squared test — `chisq.test()`](#chi-squared-test-chisq.test)
- [Linear regression — `lm()` and `summary()`](#linear-regression-lm)

**Also in this series:**

- [Part 1: Simulation and Data Summaries](transitioning-simulation.html)
- [Part 2: Inference and Visualization](transitioning-inference.html)
- [Cheat Sheet](transitioning-cheatsheet.html)
</div>

<div class="warn-box">
**A note on assumptions.** Theory-based tests rely on mathematical
approximations that are only valid when certain assumptions hold (e.g.,
approximate normality, independence). Always check that assumptions are
reasonable before trusting a p-value. The randomization approach from
Parts 1 and 2 is more robust when assumptions are uncertain.
</div>

---

## t-tests: Comparing Means

`t.test()` covers three common scenarios: testing a single mean against a
reference value, comparing two independent groups, and comparing paired
measurements. All three return the same type of output — a t-statistic,
degrees of freedom, p-value, and confidence interval.

### One-sample t-test

Tests whether a population mean equals a specified value $\mu_0$.

$$H_0: \mu = \mu_0 \qquad H_A: \mu \ne \mu_0$$

```{r one_sample_t}
# Are Yale students' average sleep different from the national average of 7 hrs?
set.seed(5502)
sleep_hours <- c(6.1, 7.3, 5.8, 6.5, 7.0, 6.8, 5.5, 7.2, 6.9, 6.3,
                 7.1, 5.9, 6.4, 7.4, 6.0, 6.7, 5.6, 7.5, 6.2, 6.6)

t.test(sleep_hours, mu = 7)
```

Reading the output:

- **t** — the test statistic
- **df** — degrees of freedom (n − 1)
- **p-value** — probability of observing a result this extreme if H₀ were true
- **95% confidence interval** — a range of plausible values for the true mean
- **sample mean** — your observed mean

### Two-sample independent t-test

Tests whether two group means are equal. This is the theory-based equivalent
of the permutation test in Part 2. Using the **calcium supplement study** from
class 17:

$$H_0: \mu_\text{treat} = \mu_\text{control} \qquad H_A: \mu_\text{treat} > \mu_\text{control}$$

```{r two_sample_t}
treat   <- c(7, -4, 18, 17, -3, -5,  1, 10, 11, -2)
control <- c(-1, 12, -1, -3,  3, -5,  5,  2, -11, -1, -3)

t.test(treat, control, alternative = "greater")
```

The `alternative` argument controls the direction of the test:

| Value | Hypothesis |
|---|---|
| `"two.sided"` (default) | $H_A: \mu_1 \ne \mu_2$ |
| `"greater"` | $H_A: \mu_1 > \mu_2$ |
| `"less"` | $H_A: \mu_1 < \mu_2$ |

<div class="tip-box">
By default `t.test()` uses Welch's correction (it does **not** assume equal
variances). This is the safer choice for most real data. If you have strong
reason to assume equal variances, add `var.equal = TRUE`, but this is rarely
necessary.
</div>

### Paired t-test

When observations come in natural pairs (e.g., the same car measured on both
sides), use `paired = TRUE`. This is more powerful than an independent t-test
because it removes the variation between pairs. Based on the **tire wear**
example from class 23:

$$H_0: \mu_\text{diff} = 0 \qquad H_A: \mu_\text{diff} \ne 0$$

```{r paired_t}
# Tread wear (mm) on left and right tires of the same 14 cars
set.seed(7731)
left_tire  <- c(3.1, 4.2, 2.8, 3.9, 4.5, 3.3, 2.6, 4.8, 3.7, 4.1,
                2.9, 3.5, 4.4, 3.0)
right_tire <- left_tire + rnorm(14, mean = 0.2, sd = 0.3)  # right wears slightly more

# Independent t-test (ignores pairing — less powerful)
t.test(left_tire, right_tire)

# Paired t-test (accounts for pairing — more powerful)
t.test(left_tire, right_tire, paired = TRUE)
```

Notice the paired test gives a smaller p-value — it is more sensitive because
it focuses on the *within-car* differences rather than being diluted by
variation between cars.

### Quick reference

| Task | Base R |
|---|---|
| One-sample t-test | `t.test(x, mu = mu_0)` |
| Two-sample independent t-test | `t.test(x, y, alternative = "two.sided")` |
| Paired t-test | `t.test(x, y, paired = TRUE)` |
| Extract the p-value | `t.test(...)$p.value` |
| Extract the CI | `t.test(...)$conf.int` |

---

## Correlation: `cor.test()`

`cor.test()` tests whether the correlation between two quantitative variables
is significantly different from zero. This is the theory-based counterpart to
the permutation test for correlation that you built using `shuffle()` in Part 1.

$$H_0: \rho = 0 \qquad H_A: \rho \ne 0$$

Using simulated data inspired by the **cigarette consumption and lung cancer**
example from class 25:

```{r cor_test}
set.seed(2947)
cigs_per_capita <- runif(44, min = 10, max = 45)
cancer_rate     <- 2 + 0.005 * cigs_per_capita * 1000 + rnorm(44, sd = 5)

cor.test(cigs_per_capita, cancer_rate)
```

The output includes:

- **t** and **df** — the t-statistic and degrees of freedom used for the test
- **p-value** — for H₀: ρ = 0
- **95% confidence interval** — a range of plausible values for the true
  correlation ρ
- **cor** — the sample correlation r

You can also compute the correlation alone with `cor()`:

```{r cor_only}
cor(cigs_per_capita, cancer_rate)
```

### Quick reference

| Task | Base R |
|---|---|
| Test H₀: ρ = 0 | `cor.test(x, y)` |
| Compute correlation only | `cor(x, y)` |
| Spearman (rank-based) correlation | `cor.test(x, y, method = "spearman")` |

---

## One-way ANOVA: `aov()`

A one-way analysis of variance tests whether the means of three or more groups
are all equal. In S&DS 1000, you used `get_F_stat()` to compute the F-statistic
for a randomization test. `get_F_stat()` is itself a thin wrapper around
`aov()`:

```{r get_F_stat_source, eval=FALSE}
# What get_F_stat() does internally:
get_F_stat <- function(data, grouping) {
  fit         <- aov(data ~ grouping)
  fit_summary <- summary(fit)
  fit_summary[[1]]$`F value`[1]   # extract just the F-statistic
}
```

Calling `aov()` directly gives you the full picture — not just the F-statistic,
but the p-value, degrees of freedom, and more, all in one step.

### Running the ANOVA

Based on the **Sudoku completion times by major** from class 24:

$$H_0: \mu_\text{bio} = \mu_\text{cs} = \mu_\text{econ} = \mu_\text{psych}
\qquad H_A: \text{at least one mean differs}$$

```{r anova}
set.seed(1142)
completion_times <- c(rnorm(10, mean = 22, sd = 4),   # Biology
                      rnorm(10, mean = 20, sd = 4),   # CS
                      rnorm(10, mean = 25, sd = 4),   # Econ
                      rnorm(10, mean = 21, sd = 4))   # Psychology
majors <- rep(c("Biology", "CS", "Econ", "Psychology"), each = 10)

fit <- aov(completion_times ~ majors)
summary(fit)
```

Reading the `summary()` output:

| Column | Meaning |
|---|---|
| `Df` | Degrees of freedom (groups − 1; then N − groups) |
| `Sum Sq` | Sum of squares (between-group; within-group) |
| `Mean Sq` | Sum Sq / Df |
| `F value` | The F-statistic — equivalent to `get_F_stat()` |
| `Pr(>F)` | The p-value |

### Post-hoc comparisons with `TukeyHSD()`

A significant ANOVA F-test tells you that *some* group means differ, but not
*which* ones. Use `TukeyHSD()` (Tukey's Honest Significant Difference) to make
all pairwise comparisons while controlling for multiple testing:

```{r tukey}
TukeyHSD(fit)
```

Each row compares two groups. The `p adj` column gives p-values adjusted for
the number of comparisons — use these rather than raw p-values to avoid false
positives.

<div class="tip-box">
`aov()` assumes approximately normal residuals and equal variance across
groups. For a quick check of the equal-variance assumption, look at whether
the group standard deviations are roughly similar. The randomization F-test
from Part 1 (using `get_F_stat()` inside `replicate()`) makes neither
assumption and is a robust alternative.
</div>

### Quick reference

| Task | SDS1000 | Base R |
|---|---|---|
| Compute F-statistic only | `get_F_stat(data, group)` | `summary(aov(data ~ group))[[1]]$'F value'[1]` |
| Full ANOVA with p-value | — | `summary(aov(data ~ group))` |
| Pairwise group comparisons | — | `TukeyHSD(aov(data ~ group))` |

---

## Chi-squared test: `chisq.test()`

A chi-squared goodness-of-fit test asks whether the observed counts across
categories match a set of expected proportions. `get_chisqr_stat()` in SDS1000
is a direct wrapper — it simply calls `chisq.test()` and returns the statistic:

```{r get_chisqr_source, eval=FALSE}
# What get_chisqr_stat() does internally:
get_chisqr_stat <- function(observed_counts, expected_proportions) {
  test_output <- chisq.test(observed_counts, p = expected_proportions)
  test_output$statistic   # extract just the chi-squared value
}
```

Calling `chisq.test()` directly gives you the full test result.

### Goodness-of-fit test

Based on the **Yale student birth months** example from class 23 — are students
equally likely to be born in any month?

$$H_0: \text{birth months are uniformly distributed}
\qquad H_A: \text{some months are more common than others}$$

```{r chisq_gof}
# Observed birth month counts for 198 Yale students (class 23)
observed_counts <- c(14, 11, 21, 17, 15, 13, 19, 16, 18, 22, 14, 18)
names(observed_counts) <- month.abb

# Test against uniform distribution (equal probability for each month)
chisq.test(observed_counts, p = rep(1/12, 12))
```

Reading the output:

- **X-squared** — the chi-squared test statistic (same value `get_chisqr_stat()` returns)
- **df** — degrees of freedom (number of categories − 1)
- **p-value** — probability of observing this much deviation from uniform if H₀ were true

### Chi-squared test of independence

`chisq.test()` also tests whether two categorical variables are *independent*
of each other (i.e., whether they are associated). Pass a contingency table
instead of a count vector:

```{r chisq_independence}
# Are vaccination status and illness related?
vaccine_table <- matrix(c(45, 15, 20, 70),
                        nrow = 2,
                        dimnames = list(c("Vaccinated", "Unvaccinated"),
                                        c("Got Ill", "Stayed Healthy")))
vaccine_table

chisq.test(vaccine_table)
```

### Quick reference

| Task | SDS1000 | Base R |
|---|---|---|
| Chi-squared statistic only | `get_chisqr_stat(obs, expected_p)` | `chisq.test(obs, p = expected_p)$statistic` |
| Full goodness-of-fit test | — | `chisq.test(obs, p = expected_p)` |
| Test of independence | — | `chisq.test(contingency_table)` |

---

## Linear regression: `lm()` and `summary()`

`lm()` fits a linear regression model. `summary()` reports the theory-based
hypothesis test for each coefficient — whether the slope (or intercept) is
significantly different from zero — and `confint()` gives confidence intervals.

Using the **cigarette consumption and lung cancer** data from class 25:

$$H_0: \beta_1 = 0 \qquad H_A: \beta_1 \ne 0$$

```{r lm_full}
set.seed(2947)
cigs_per_capita <- runif(44, min = 10, max = 45)
cancer_rate     <- 2 + 0.005 * cigs_per_capita * 1000 + rnorm(44, sd = 5)

lung_lm <- lm(cancer_rate ~ cigs_per_capita)
summary(lung_lm)
```

Reading the `summary()` output:

| Column | Meaning |
|---|---|
| `Estimate` | The fitted intercept (β₀) and slope (β₁) |
| `Std. Error` | Standard error of each estimate |
| `t value` | Test statistic for H₀: coefficient = 0 |
| `Pr(>|t|)` | p-value (two-sided) |
| `R-squared` | Proportion of variance in y explained by x |

### Confidence intervals for the coefficients

```{r lm_confint}
confint(lung_lm, level = 0.95)
```

The row for `cigs_per_capita` gives a 95% CI for the slope. This is the
theory-based alternative to the bootstrap CI built in Part 1.

### Connecting to the bootstrap

In class 25, you estimated the slope's uncertainty using `resample_pairs()`
inside `do_it()`. The theory-based CI from `confint()` will give a very similar
interval when the regression assumptions (linearity, normality of residuals,
constant variance) are met. When they're not, the bootstrap is more reliable.

| Task | SDS1000 (bootstrap) | Base R (theory-based) |
|---|---|---|
| Slope estimate | `coef(lm(...))["x"]` | `coef(lm(...))["x"]` |
| CI for slope | `resample_pairs()` + `quantile()` | `confint(lm(...))` |
| p-value for slope | `mean(null_dist >= obs)` | `summary(lm(...))$coefficients[2, 4]` |

---

## Putting it all together: which test when?

| Research question | Data types | Test |
|---|---|---|
| Is the mean different from a reference value? | One quantitative variable | `t.test(x, mu = mu_0)` |
| Are two independent group means different? | Quantitative + binary group | `t.test(x, y)` |
| Are two paired measurements different? | Two quantitative variables (paired) | `t.test(x, y, paired = TRUE)` |
| Are three or more group means different? | Quantitative + categorical group | `summary(aov(y ~ group))` |
| Is there a linear association between two variables? | Two quantitative variables | `cor.test(x, y)` |
| Do observed counts match expected proportions? | One categorical variable | `chisq.test(counts, p = expected_p)` |
| Are two categorical variables associated? | Two categorical variables | `chisq.test(table)` |
| How does y change with x (and what is the uncertainty)? | Two quantitative variables | `summary(lm(y ~ x))` + `confint(lm(y ~ x))` |
