---
title: "Homework 3"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---






$\\$






In the following homework, you will gain practice creating linear models to 
predict the values of a response variable from the value of an explanatory 
variable, and you will examine questions related to sampling bias. Please submit 
a compiled pdf with your answers to the exercises to Gradescope 
*by 11pm on Sunday February 8th*. Please also make sure to  indicate the correct 
pages to each question's answers on Gradescope to make sure that we can grade them.


Some useful functions for completing this worksheet are:  `plot()`, `corr()`,
`lm()`, `abline()`, and `coef()`. Also see the "R function" page on Canvas for a
list of functions we have used in class.You might also find the following symbols 
useful:  $\mu$, $\bar{x}$, $\pi$, $\hat{p}$, $\rho$, $\hat{y}$, and $\beta$. 

For all questions that involve R, please use the R chunks (gray sections) to 
create relevant plots and calculations, and then answer the questions in the 
"answer" section below each chunk. Be sure to "show your work" by printing out 
information that answers the question that have been asked (although please do 
not print out pages of irrelevant data). Also, make sure to add appropriate 
labels to all plots.  


As always, if you need help with any of the homework assignments, please attend
the TA office hours which are listed on Canvas and/or ask questions on Ed
Discussions forum. Also, if you have completed the homework, please help others
out by answering questions on Ed Discussions. Finally, reviewing the class
slides and videos will be helpful if you run into difficulties, and attending
the Practice Sessions will be very useful to solidify your knowledge. 





$\\$





## Part 1: Simple linear regression
 



$\\$



On the last homework you practiced calculating statistics "by hand" using data 
on  the number of hot dogs eaten by the winner of Nathan's Famous hot dog eating 
contest. Let's continue to analyze this data set by looking at the question 
of whether it has become harder to win this hot dog eating contest in more recent 
years.

The code below loads data on the winners of Nathan's Famous hot dog eating
contest from the years 2002 to 2019. Please use the R chunk below to analyze
data related to that contest and answer the following questions in the answer
section below. Also, be sure to "show your work" by making sure your answers are
printed out in the R chunk below.

Note: As mentioned on the previous homework, doctors recommend you limit the 
number of hot dogs you eat in a day as also discussed 
[here](https://www.tiktok.com/@paramountplusaustralia/video/7259888863781358875?lang=en).




$\\$





**Exercise 1.1 (4 points)**: In the R chunk below, create a scatter plot of the 
number of hot dogs eaten as a function of year. Is the trend in the data mostly 
positive or negative? 

```{r q1_1}

library(Lock5Data)


# load the data on the hot dog contest winners in a data frame called HotDogs
HotDogs <- HotDogs2019


# Plot the number of hot dogs eaten as a function of the year







```


**Answer** 







$\\$





**Exercise 1.2 (7 points)**: Fit a linear model to this data predicting the 
number of hot dogs eaten as a function of the year, and save the output of your
fitted model to an object called `hotdog_fit`. Then recreate the scatter plot
from 1.1 and add the fitted regression line to your plot (bonus, see if you can
make the line color red). Then from **just looking at the resulting plot**,
answer the questions of whether: 

1. Is the residual larger in 2007 or in 2008?

2. Is the residual positive or negative in 2010?


```{r q1_2}


# Recreate the plot above






# Fit a linear regression model data and add the regression line to the plot






# Add the regression line to the plot







```


**Answer** 

1. 


2. 






$\\$






**Exercise 1.3 (4 points)**: By looking at the scatter plot, guess the 
correlation between year and number of hot dogs eaten. Then use R to find the
correlation value and report what it is. Is the actual correlation value
approximately what you expected it to be?


```{r q1_3}

# Calculate the correlation between the number of hot dogs eaten and the year





```


**Answer** 







$\\$





**Exercise 1.4 (6 points)**: Now extract the coefficients from your linear model 
in the R chunk below. Then in the answers section, write out what the regression
equation is; i.e., write the equation with the values of the coefficients
plugged in.

Hint: using the symbol $\hat{y}$ could be useful here. 


```{r q1_4}

# Print out the regression coefficients here. 
# Then write the equation for the prediction line in the answers section below.





```


**Answer** 






$\\$





**Exercise 1.5 (6 points)**: In the answer section, write an interpretation of 
what the coefficients mean in terms of years and the number of hot dogs
predicted to be eaten.



**Answer** 








$\\$






**Exercise 1.6 (6 points)**: Use your regression equation to predict the number 
of hot dogs expected to be eaten in 2016. Show your work in R. Then find the
actual number of hot dogs eaten by the winner in 2016 on the internet and report
the residual.


```{r q1_6}









```



**Answer** 







$\\$






**Exercise 1.7 (5 points)**: Would it be appropriate to use this regression line 
to predict the winners number of hot dogs in 2050? Please describe why or why not. 



```{r 1_7}







```



**Answer** 








$\\$







## Part 2: Questions examining sampling and bias


The following questions are intended to improve your understanding of sampling, 
bias, sampling distributions and confidence intervals. I also recommend trying 
additional questions from the Lock5 textbook to get more practice.



$\\$





**Exercise 2.1** (6 points): In a sample survey of professors at the University 
of Nebraska, 94% of them described themselves as "above average" teachers. Please 
answer the following questions: 

a. What is the sample? What is the population?

b. Based on the information provided, can we conclude that the study suffers 
from sampling bias? 

c. Is 94% a good estimate for the percentage of above-average teachers at the 
University of Nebraska? If not, why not? 

Note: For this question you can assume that a simple random sample of professors
at the University of Nebraska was taken and that the sample size was fairly large. 



**Answers**:

a. 



b. 



c. 





$\\$





**Exercise 2.2** (7 points): Employment statistics in the US are often based on 
two nationwide monthly surveys: the Current Population Survey (CPS) and the 
Current Employment Statistics (CES) survey. The CPS samples approximately 60,000 
US households and collects the employment status, job type, and demographic 
information of each resident in the household. The CES survey samples 140,000 
nonfarm businesses and government agencies and collects the number of payroll 
jobs, pay rates, and related information for each firm.

a. What is the population in the CPS survey?

b. What is the population in the CES survey?

For each of the following statistical questions, state whether the results from 
the CPS or CES survey would be more relevant.

  i.	 Do larger companies tend to have higher salaries?

  ii.	 What percentage of Americans are self-employed?

  iii. Are married men more or less likely to be employed than single men?




**Answers**:


a. 


b. 


i. 

ii. 

iii. 






$\\$







## Reflection (3 points)


Please reflect on how the homework went by going to Canvas, going to the Quizzes 
link, and clicking on Reflection on homework 3.







