---
title: "Homework 3"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---






$\\$






In the following homework, you will gain practice examining relationships between 
quantitative data. Please submit a compiled pdf with your answers to the exercises 
to Gradescope *by 11pm on Sunday September 21st*. Please also make sure to 
indicate the correct pages to each question's answers on Gradescope to make sure 
that we can grade them.


Some useful functions for completing this worksheet are:  `plot()`, `corr()`,
`lm()`, `abline()`, and `coef()`. Also see the "R function" page on Canvas for a
list of functions we have used in class.


You might also find the following symbols useful:  $\mu$, $\bar{x}$, $\pi$, 
$\hat{p}$, $\rho$, $\hat{y}$. For full credit on problems involving R, be sure 
to label all your figures and to "show your work" by making sure all values that 
are answers to questions are printed/visible in your R Markdown pdf.


For all questions that involve R, please use the R chunks (gray sections) to 
create relevant plots and calculations, and then answer the questions in the 
"answer" section below each chunk. Be sure to "show your work" by printing out 
information that answers the question that have been asked (although please do 
not print out pages of irrelevant data). Also, make sure to add appropriate 
labels to all plots.  


As always, if you need help with any of the homework assignments, please attend
the TA office hours which are listed on Canvas and/or ask questions on Ed
Discussions forum. Also, if you have completed the homework, please help others
out by answering questions on Ed Discussions. Finally, reviewing the class
slides and videos will be helpful if you run into difficulties.





$\\$





## Part 1: Scatter plots and correlation 




$\\$



The following questions are based on questions from the Lock5 textbook and are 
intended to improve your understanding of scatter plots and correlation. I 
recommend you try additional questions from the textbook to get extra practice. 





$\\$




**Exercise 1.1 (4 points)**:  Let's explore two quantitative variables related 
to cars. Describe whether you expect a positive or negative association between
the *distance* a car was driven since the last fill-up of the gas tank, and the
*amount of gas* left in the tank. In 1-3 sentences, please explain the rationale
behind your answer.




**Answer**:  






$\\$





  
**Exercise 1.2 (12 points)**:  Suppose we record the husband's age and the 
wife's age for 100 randomly selected heterosexual married couples. Please answer 
the following questions:

a. What would it mean about ages of couples if these two variables had a
negative relationship?

b. What would it mean about ages of couples if these two variables had a
positive relationship?

c. Which do you think is more likely, a negative or a positive relationship? 

d. Do you expect a strong or weak relationship in the data? Why? 

e. Would a strong correlation imply there is an association between husband age
and wife age?

f. What symbol (of the symbols discussed in class) should we use to denote the
correlation discussed in this problem?



**Answers**:


a. 



b. 



c. 



d. 



e. 



f. 






$\\$







**Exercise 1.3 (6 points)**: The table below shows the relationship between two 
quantitative variables X and Y (you might want to render the document to see the
table better). Please use R to create a scatter plot and calculate the
correlation between these variables.  Hint: the function to create vectors,
`c()`, might be useful here.  *Challenge (0 points):* try to calculate the
correlation without using the `cor()` function.



   |    |    |    |    |    |    |    |    |    
---|----|----|----|----|----|----|----|----|     
X  | 15 | 20 | 25 | 30 | 35 | 40 | 45 | 50 |
Y  |532 |466 |478 |320 |303 |349 |275 |221 |





```{r q1_3}











```







$\\$








**Exercise 1.4 (6 points)**:  When honeybee scouts find a food source or a nice 
site for a new home, they communicate the location to the rest of the swarm by
doing a ["waggle dance"](https://www.youtube.com/watch?v=-7ijI-g4jHg). They
point in the direction of the site and dance longer for sites farther away. The
rest of the bees use the duration of the dance to predict distance to the site.
The code below loads a data frame called `HoneybeeWaggle` that contains the
distance (in meters), and the duration of the dance (in seconds) for seven
honeybee scouts. Using this data:

a. Create a scatter plot of the data, and be sure to label the plot
appropriately. Also describe whether there appears to be a linear trend in the
data. If so, is it positive or negative?

b. Use the `cor()` function to find the correlation between the two variables. 


Note: if you learn more about correlations through (human) dance please check out
[this video](https://www.youtube.com/watch?v=VFjaBh12C6s)



```{r q1_4}


# download the data and load it into R
library(Lock5Data)
data(HoneybeeWaggle)


# Plot the data here







# Calculate the correlation coefficient here






```


**Answers:**

a. 

b. 







$\\$






## Part 2: Simple linear regression
 



$\\$



In class we examined data about the number of hot dogs eaten by the winner of
Nathan's Famous hot dog eating contest. An additional question we can ask using
this data set, is whether it has become harder to win this hot dog eating
contest in more recent years.

The code below loads data on the winners of Nathan's Famous hot dog eating
contest from the years 2002 to 2019. Please use the R chunk below to analyze
data related to that contest and answer the following questions in the answer
section below. Also, be sure to "show your work" by making sure your answers are
printed out in the R chunk below.



$\\$





**Exercise 2.1 (4 points)**: In the R chunk below, create a scatter plot of the 
number of hot dogs eaten as a function of year. Is the trend in the data mostly 
positive or negative? 

```{r q2_1}


# load the data on the hot dog contest winners in a data frame called HotDogs
HotDogs <- HotDogs2019


# Plot the number of hot dogs eaten as a function of the year






```


**Answer** 







$\\$





**Exercise 2.2 (7 points)**: Fit a linear model to this data predicting the 
number of hot dogs eaten as a function of the year, and save the output of your
fitted model to an object called `hotdog_fit`. Then recreate the scatter plot
from 2.1 and add the fitted regression line to your plot (bonus, see if you can
make the line color red). Then from **just looking at the resulting plot**,
answer the questions of whether: 

a. Is the residual larger in 2007 or in 2008?

b. Is the residual positive or negative in 2010?


```{r q2_2}


# Recreate the plot above





# Fit a linear regression model data and add the regression line to the plot





# Add the regression line to the plot







```


**Answer** 

a.  


b. 






$\\$




**Exercise 2.3 (4 points)**: By looking at the scatter plot, guess the 
correlation between year and number of hot dogs eaten. Then use R to find the
correlation value and report what it is. Is the actual correlation value
approximately what you expected it to be?


```{r q2_3}

# Calculate the correlation between the number of hot dogs eaten and the year




```


**Answer** 








$\\$





**Exercise 2.4 (6 points)**: Now extract the coefficients from your linear model 
in the R chunk below. Then in the answers section, write out what the regression
equation is; i.e., write the equation with the values of the coefficients
plugged in.

Hint: using the symbol $\hat{y}$ could be useful here. 


```{r q2_4}

# Print out the regression coefficients here. 
# Then write the equation for the prediction line in the answers section below.








```


**Answer** 









$\\$





**Exercise 2.5 (6 points)**: In the answer section, write an interpretation of 
what the coefficients mean in terms of years and the number of hot dogs
predicted to be eaten.



**Answer** 













$\\$






**Exercise 2.6 (6 points)**: Use your regression equation to predict the number 
of hot dogs expected to be eaten in 2016. Show your work in R. Then find the
actual number of hot dogs eaten by the winner in 2016 on the internet and report
the residual.


```{r q2_6}









```



**Answer** 










$\\$





**Exercise 2.7 (5 points)**: Would it be appropriate to use this regression line 
to predict the winners number of hot dogs in 2050? Please describe why or why not. 



```{r 2_7}









```



**Answer** 











$\\$





## Part 3: Review of descriptive statistics (6 points)

We are now wrapping up the section of the class on descriptive statistics. As was
mentioned at the start of the semester, the material in the class builds on what
was learned in previous classes. Thus, before we go forward and start discussing
statistical inference, it is important that you have a firm understanding of the
material we have covered so far.


In order to make sure you are understanding all the material discussed in the
class, we please review the class material (e.g., slides, textbook, etc.). 
One you are done reviewing this material, please go to Canvas and fill out the
[Descriptive statistics review quiz](https://yale.instructure.com/courses/109046/quizzes/95649)
which will test your knowledge of the concepts we have covered in class so far.


When filling out the quiz, please answer all questions without using any any
external material (i.e., without using any of the class slides, the textbook,
chatGPT, etc). **You will only be graded only on whether you complete the quiz**
and not on how many answers you got correct (i.e., if you complete the quiz, you
will get 6 points for this homework question regardless of how you do on the
quiz). However, if you do poorly on the quiz, please take additional time to
study so that you will be prepared for what we discuss in coming weeks. There
will likely be related questions on the midterm exam, so studying now will help
prepare you for that exam!





$\\$





## Reflection (3 points)


Please reflect on how the homework went by going to Canvas, going to the Quizzes 
link, and clicking on Reflection on homework 3.







