---
title: "Class 10 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---




# Part 1: Normal quantiles

We use quantiles from a standard normal distribution to create confidence
intervals for different confidence levels.

For a given confidence level *C*, we want to find the quentile value q* such
that C percent of the area of a standard normal distribution is between -q* and
q*.



```{r}

# install.packages("mosaic")

quantile_values <- mosaic::xqnorm(c(.025, .975))


quantile_values


```





$\\$





#  Part 2: Simple example of sampling with replacement



```{r}


library(SDS1000)


# simple vector
my_sample <- c(3, 1, 4, 1, 5, 9)


# resampling values in the vector with replacement
boot_sample  <-  sample(my_sample,  6,  replace = TRUE)


```




$\\$




#  Part 3: Simple fake data bootstrap example


```{r}

# simple bootstrap example

my_sample <- c(21, 29, 25, 19, 24, 22, 25, 26, 25, 29)


# calculate the x-bar statistic
obs_mean <- mean(my_sample)


# create the bootstrap distribution
bootstrap_dist <-  do_it(10000) * {    
  
  curr_boot <- sample(my_sample , 10, replace = TRUE)  
  mean(curr_boot)  
  
}


# visualize the bootstrap distribution
hist(bootstrap_dist, 100)


# calculate the standard error
SE_boot <- sd(bootstrap_dist)


# 95% confidence interval
CI_lower <-  obs_mean  - 2 * SE_boot 
CI_upper <-  obs_mean  + 2 * SE_boot 


c(CI_lower, CI_upper)



```




$\\$




#  Part 4: Confidence interval for Gettysburg mean word length


```{r}


# set the random seed
set.seed(123)


# load the gettysburg data
load("gettysburg.Rda")


# store the word lengths in a vector object called word_lengths
word_lengths <- gettysburg$num_letters


# get a random sample from the word lengths

n <- 10

my_sample <- sample(word_lengths, size = n)



# calculate the x-bar statistic
obs_mean <- mean(my_sample)


# create the bootstrap distribution
bootstrap_dist <-  do_it(10000) * {    
  
  curr_boot <- sample(my_sample, n, replace = TRUE)  
  mean(curr_boot)  
  
}


# visualize the bootstrap distribution
hist(bootstrap_dist, 100)


# calculate the standard error
(SE_boot <- sd(bootstrap_dist))


# 95% confidence interval
CI_lower <-  obs_mean  - 2 * SE_boot 
CI_upper <-  obs_mean  + 2 * SE_boot 


c(CI_lower, CI_upper)



```






$\\$





##  Confidence interval example 1: One true love? 


A survey asked 2625 people whether they agreed with the statement: 
"There is only one true love for each person"

1812 of the respondents disagreed.

Compute a 90% confidence interval for the proportion who disagreed.





```{r}

library(SDS1000)


# Create the data set

belives_in_true_love <- c(rep(FALSE, 1812), rep(TRUE, 2625 - 1812))


# Calculate the proportion of people who believe in one true love
# Note: if you use the get_proportion function the word "TRUE" needs to be in quotes

(obs_prop <- get_proportion(belives_in_true_love, "TRUE"))




# Get an example of a bootstrap statistics 

# We can use the sample(data_vec, n) to get a sample of length n:

boot_sample <- sample(belives_in_true_love, 2625, replace = TRUE)
bootstrap_prop <- get_proportion(boot_sample, "TRUE")



# Create a bootstrap distribution in R

boot_dist <- do_it(10000)  *  {

  boot_sample <- sample(belives_in_true_love, 2625, replace = TRUE)
  
  get_proportion(boot_sample, "TRUE")
  
}



# Visualize the bootstrap distribution

hist(boot_dist, 20)




# Get the bootstrap standard error SE*

SE_boot <- sd(boot_dist)





# Create a 90% confidence interval

CI_lower <-  obs_prop  - qnorm(.95) * SE_boot 
CI_upper <-  obs_prop  + qnorm(.95) * SE_boot 

c(CI_lower, CI_upper)



```







$\\$






## Confidence interval example 2: Average body temperature


Let's create a 95% confidence interval for the mean body temperatures of humans
using a sample of 50 people from the data collected by Machowaik et al (1992). 

The data is loaded below from the Lock5 package. Start by visualizing the data
then please create a 95% confidence interval for the mean body temperature using
the bootstrap!



```{r bodytemp_viz}

# Load the data 
library(Lock5Data)

body_temps <- BodyTemp50$BodyTemp




# Start by visualizing the data

hist(body_temps, main = "Body Temperatures", 
     breaks = 20,
     xlab = "Temperature (°F)", col = "lightblue", border = "black")


```




Now create a 95% confidence interval for the mean body temperature using the
bootstrap!

See if you can also create a 80% and 99% confidence interval as well!




```{r bodytemp_CI}


# Calculate the observed mean

(obs_mean <- mean(body_temps))





# Create a bootstrap distribution in R

boot_dist <- do_it(10000)  *  {

  boot_sample <- sample(body_temps, length(body_temps), replace = TRUE)
  
  mean(boot_sample)
  
}



# Visualize the bootstrap distribution

hist(boot_dist, 100)




# Get the bootstrap standard error SE*

SE_boot <- sd(boot_dist)




# Create a 95% confidence interval

CI_lower <-  obs_mean  - 2 * SE_boot 
CI_upper <-  obs_mean  + 2 * SE_boot 

c(CI_lower, CI_upper)



# Create a 80% confidence interval

(CI_80 <-  c(obs_mean - qnorm(.90) * SE_boot, obs_mean + qnorm(.90) * SE_boot))



# Create a 99% confidence interval

(CI_99 <-  c(obs_mean - qnorm(.995) * SE_boot, obs_mean + qnorm(.995) * SE_boot))




```



