---
title: "Class 10 notes and code"
format: pdf
editor: source
editor_options:
  chunk_output_type: console
---




# Part 1: Normal quantiles

We can use quantiles from a standard normal distribution to create confidence
intervals for different confidence levels.

For a given confidence level *C*, we want to find the quantile value z* such
that C percent of the area of a standard normal distribution is between -z* and
z*. To do this we can use the `cnorm()` function from the SDS1000 package.



```{r}

library(SDS1000)









```



$\\$

Suppose we had data where the mean of the data was $\bar{x} = 12$ and we knew
the standard error was $SE = 2$. What are the 80%, 90%, and 95% confidence
intervals for $\mu$?


```{r}

x_bar <- 12
SE <- 2


# 80% confidence interval



# 90% confidence interval



# 95% confidence interval



```


Notice that the intervals get wider as the confidence level increases.




$\\$





#  Part 2: Simple example of sampling with replacement



```{r}


library(SDS1000)


# simple vector
my_sample <- c(3, 1, 4, 1, 5, 9)


# resampling values in the vector with replacement




# print the bootstrap sample (some values can appear more than once)



```




$\\$




#  Part 3: Simple fake data bootstrap example


```{r}

# simple bootstrap example

my_sample <- c(21, 29, 25, 19, 24, 22, 25, 26, 25, 29)


# calculate the x-bar statistic




# create the bootstrap distribution








# visualize the bootstrap distribution




# calculate the standard error




# 95% confidence interval (2 is approximately cnorm(.95) = 1.96)






```




$\\$




#  Part 4: Confidence interval for Gettysburg mean word length


```{r}


# set the random seed
set.seed(123)


# load the gettysburg data
load("gettysburg.Rda")


# store the word lengths in a vector object called word_lengths
word_lengths <- gettysburg$num_letters


# get a random sample from the word lengths

n <- 10





# calculate the x-bar statistic




# create the bootstrap distribution








# visualize the bootstrap distribution




# calculate the standard error




# 95% confidence interval







# since we have the whole population here, we can check if the interval
# captured the population mean word length (mu)





```


