---
title: "Class 10 notes and code"
format: pdf
editor: source
editor_options:
  chunk_output_type: console
---




# Part 1: Normal quantiles

We can use quantiles from a standard normal distribution to create confidence
intervals for different confidence levels.

For a given confidence level *C*, we want to find the quantile value z* such
that C percent of the area of a standard normal distribution is between -z* and
z*. To do this we can use the `cnorm()` function from the SDS1000 package.



```{r}

library(SDS1000)


critical_val_95 <- cnorm(.95)
critical_val_95


critical_val_80 <- cnorm(.80)
critical_val_80


# Also see the mosaic package xcnorm function for visualizing critical values
mosaic::xcnorm(.95)
mosaic::xcnorm(.80)


```



$\\$

Suppose we had data where the mean of the data was $\bar{x} = 12$ and we knew
the standard error was $SE = 2$. What are the 80%, 90%, and 95% confidence
intervals for $\mu$?


```{r}

x_bar <- 12
SE <- 2


# 80% confidence interval
c(x_bar - cnorm(.80) * SE, x_bar + cnorm(.80) * SE)


# 90% confidence interval
c(x_bar - cnorm(.90) * SE, x_bar + cnorm(.90) * SE)


# 95% confidence interval
c(x_bar - cnorm(.95) * SE, x_bar + cnorm(.95) * SE)


```


Notice that the intervals get wider as the confidence level increases.




$\\$





#  Part 2: Simple example of sampling with replacement



```{r}


library(SDS1000)


# simple vector
my_sample <- c(3, 1, 4, 1, 5, 9)


# resampling values in the vector with replacement
boot_sample  <-  sample(my_sample,  6,  replace = TRUE)


# print the bootstrap sample (some values can appear more than once)
boot_sample


```




$\\$




#  Part 3: Simple fake data bootstrap example


```{r}

# simple bootstrap example

my_sample <- c(21, 29, 25, 19, 24, 22, 25, 26, 25, 29)


# calculate the x-bar statistic
obs_mean <- mean(my_sample)
obs_mean


# create the bootstrap distribution
bootstrap_dist <-  do_it(10000) * {

  curr_boot <- sample(my_sample, length(my_sample), replace = TRUE)
  mean(curr_boot)

}


# visualize the bootstrap distribution
hist(bootstrap_dist, 100,
     xlab = "Mean of the bootstrap sample",
     main = "Bootstrap distribution")


# calculate the standard error
SE_boot <- sd(bootstrap_dist)
SE_boot


# 95% confidence interval (2 is approximately cnorm(.95) = 1.96)
CI_lower <-  obs_mean  - 2 * SE_boot
CI_upper <-  obs_mean  + 2 * SE_boot


c(CI_lower, CI_upper)



```




$\\$




#  Part 4: Confidence interval for Gettysburg mean word length


```{r}


# set the random seed
set.seed(123)


# load the gettysburg data
load("gettysburg.Rda")


# store the word lengths in a vector object called word_lengths
word_lengths <- gettysburg$num_letters


# get a random sample from the word lengths

n <- 10

my_sample <- sample(word_lengths, size = n)



# calculate the x-bar statistic
obs_mean <- mean(my_sample)
obs_mean


# create the bootstrap distribution
bootstrap_dist <-  do_it(10000) * {

  curr_boot <- sample(my_sample, n, replace = TRUE)
  mean(curr_boot)

}


# visualize the bootstrap distribution
hist(bootstrap_dist, 100,
     xlab = "Mean word length of the bootstrap sample",
     main = "Bootstrap distribution of the mean word length")


# calculate the standard error
(SE_boot <- sd(bootstrap_dist))


# 95% confidence interval
CI_lower <-  obs_mean  - 2 * SE_boot
CI_upper <-  obs_mean  + 2 * SE_boot


c(CI_lower, CI_upper)


# since we have the whole population here, we can check if the interval
# captured the population mean word length (mu)
mean(word_lengths)


```


The 95% confidence interval contains the population mean word length $\mu$ of
about 4.29.


