---
title: "Class 9 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---




# Part 1: Normal quantiles

We use quantiles from a standard normal distribution to create confidence
intervals for different confidence levels.

For a given confidence level *C*, we want to find the quentile value q* such
that C percent of the area of a standard normal distribution is between -q* and
q*.



```{r}

# install.packages("mosaic")






```





$\\$




#  Part 2: Simple example of sampling with replacement



```{r}


library(SDS1000)


# simple vector
my_sample <- c(3, 1, 4, 1, 5, 9)



# resampling values in the vector with replacement




```




$\\$




#  Part 3: Simple fake data bootstrap example


```{r}

# simple bootstrap example

my_sample <- c(21, 29, 25, 19, 24, 22, 25, 26, 25, 29)



# calculate the x-bar statistic



# create the bootstrap distribution






# visualize the bootstrap distribution




# calculate the standard error





# 95% confidence interval






```




$\\$




#  Part 4: Confidence interval for Gettysburg mean word length


```{r}


# set the random seed
set.seed(123)


# load the gettysburg data
load("gettysburg.Rda")


# store the word lengths in a vector object called word_lengths
word_lengths <- gettysburg$num_letters


# get a random sample from the word lengths






# calculate the x-bar statistic





# create the bootstrap distribution









# visualize the bootstrap distribution





# calculate the standard error





# 95% confidence interval






```








