---
title: "Class 9 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---




# Part 1: Creating sampling distributions


Let's create a sampling distribution by repeatedly sampling the word lengths
from the Gettysburg Address.





$\\$





### Part 1.1: Loading the data


The code below loads the Gettysburg Address into a data frame called `gettysburg`.

Let's start by extracting the word lengths from this data frame into a vector
object called `word_lengths`. Then report how many words are in the Gettysburg Address.

We are viewing the the Gettysburg Address word lengths as our population, and
we will be drawing samples from this population to create a sampling distribution.



```{r load_gettysburg}

# Load SDS1000 library - we will use the do_it() function from this package later 
library(SDS1000)

# Load the Gettysburg Address data frame
load("gettysburg.Rda")


# store the word lengths in a vector object called word_lengths



# print out how many words in are the Gettysburg Address




```





$\\$





### Part 1.2: Generated one random sample and statistic

Now let's generate a random sample of size 10 from the `word_lengths` vector and
store it in an object called `curr_sample`. Then let's calculate the mean of this 
sample.


```{r one_sample}

# set the random seed to 1234
set.seed(1234)


# generate a random sample of size 10 from the word_lengths vector




# calculate the mean of the sample





```




$\\$





### Part 1.3: Using the do_it function


Now let's use the `do_it()` function from the `SDS1000` package to repeat a
piece of code many times.

In particular, let's add the numbers 2 and 3 together 100 times, and store the results
in a vector called `results`. Then print out the `results` vector.



```{r do_it_example}

# use the do_it function to add 2 and 3 together 100 times





# print out the results vector




```




$\\$





### Part 1.4: Creating a sampling distribution

Now let's use the `do_it()` function to create a sampling distribution of the
mean word length in samples of size n = 10 drawn from the `word_lengths` vector.

In particular, use the `do_it()` function to repeat the following steps 10,000 times:

1. Draw a random sample of size 10 from the `word_lengths` vector
2. Calculate the mean of this sample

Store the results in a vector called `sampling_dist`. Then create a histogram showing
the shape of the (approximate) sampling distribution


```{r sampling_distribution}

# use the do_it function to create a sampling distribution of the mean word length








# create a histogram of the sampling distribution







```




$\\$





### Part 1.5: Bias in our sampling distribution? 

Let's examine if there is bias in our sampling distribution by comparing the 
population mean $\mu$ to the average of our sample means; i.e., to $E[\bar{x}]$



```{r sampling_dist_bias}

# population mean




# sampling distribution mean





```




$\\$





### Part 1.6: The standard error

The **standard error (SE)** is the standard deviation of our sampling distribution. 

Let's calculate the standard error of the sampling distribution we created above. 


Note: The standard error (SE) is different from the population standard deviation 
($\sigma$) and from the sample standard deviation ($s$). It is very important
that you understand the difference between these values. If the difference is
not clear to you, please come to office hours (and/or the practice sessions)
to review this!




```{r standard_error}



```





$\\$






### Part 1.7: Exampling how the sample size changes our results

Let's explore how our results changing the sample size changes our results by
rerunning our code with different sample sizes $n$

Let's cut and paste our code above to explore this by looking at:

1. How does the shape of our sampling distribution change with different sample sizes $n$

2. How does the standard error (SE) change with different sample sizes. 

Let's try this with values of `n = 40` and `n = 160`. 


```{r diff_sample_sizes}


# create sampling distributions for sample sizes of n = 40 and n = 160








# plot a histogram of the sampling distribution





```












