---
title: "Class 12 notes and code"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---






# Part 1: More practice creating confidence intervals 



[Pereira-Doel et al (2024)](https://pmc.ncbi.nlm.nih.gov/articles/PMC7028815/)
conducted a study at the University of Surrey in the United Kingdom examining
variables related to water consumption by people taking showers. Let's use the
data from this study create a 90% confidence interval for the mean length of
time that showers last.

The data from the study showing the amount of time people spend in the shower is
loaded into R as a vector called `shower_times` below. To start, please visualize
the data. What do you notice about the distribution of shower times (i.e., 
how would you describe the shape of the distribution, etc.)? 


```{r shower_viz}

load("shower_data.rda")

shower_times <- shower_data$Time



# visualize the data here...











```



**Answer** 






$\\$




Now create a 90% confidence interval for the mean shower time using the
bootstrap!

See if you can also create a 99% confidence interval as well!





```{r shower_CI}


# Calculate the observed mean






# Create a bootstrap distribution in R







# Visualize the bootstrap distribution






# Get the bootstrap standard error SE*






# Create a 90% confidence interval








# Create a 99% confidence interval






```




$\\$




# Part 2: Hypothesis tests

In order to run a hypothesis test, we need to simulate statistics that would
occur if "nothing interesting" is happening; i.e., if the "null hypothesis" is
true.

For hypothesis tests that involve proportions, we can do this by simulating
flipping a coin k times, where k is the number of trials in our original data
set. We can then count the number of heads we get, and see if this is larger
than the number (or proportion) of positive outcomes we observed in our original
data set.

Using the `SDS1000` package, we can simulate coin flips using the `rflip()`
function. For example, if we flip a coin 100 times, where each flip has a 50%
chance of being heads, we can use `rflip(100, 0.5)` to get the number of
heads we observed in our 100 flips.

```{r coin_flip_example}

library(SDS1000)

#set.seed(1234)





```


Can do one simulation of the Doris and Buzz example?  How many times do you 
need to flip the coin to simulate the experiment that was run? 


```{r doris_buzz_simulation}

# How many times do we need to flip the coin?


# Simulate one experiment


```



To see if the result we observed in the Doris and Buzz experiment is unusual, we can
repeat the simulation many times to create a null distribution of statistics. We
can then see how many of the simulated statistics are as extreme or more extreme
than the statistic we observed in the original experiment.



```{r doris_buzz_null_distribution}

# How many times do we need to flip the coin?




# Simulate the null distribution




# Visualize the null distribution





```





