---
title: "Practice Session 11"
format: pdf
editor: source
editor_options: 
  chunk_output_type: console
---

# Part 1: Theory of Hypothesis Testing

## Introduction to Type I and Type II Errors

In hypothesis testing, we need to decide whether we are going to reject the null hypothesis, or fail to reject the null hypothesis. We sometimes, however, make the wrong decision. The types of mistakes we can make are: 

* Type I error: The null hypothesis is **true** but we mistakenly reject it;  i.e., we report a false result. 
* Type II error: The null hypothesis is **false**, but we fail to reject it;  i.e., we fail to report a real result.

What we generally try to do is to control for the Type I error rate so we only make Type I errors a small portion of the time (e.g., we only make type I errors 5% percent of the time) The probability that we make a Type I error is also known as the significance level. Typically, we set our significance level to be 0.05 in hypothesis testing (so we are only report 5% false results). 

## Question 1: Type I, Type II, or Neither?

For each of the following statements, decide whether a Type I or a Type II error was made. If no error was made, indicate that no error was committed.

a) Shiba ran a hypothesis test with the following hypotheses:

* $H_0: \mu = 50$
* $H_A: \mu > 50$

Shiba rejected the null hypothesis. In reality, the true mean is equal to 50.

**Shiba made made a Type I error: she rejected the null hypothesis when the null hypothesis is actually true.**

b) Joe wanted to test if the mean bag weight of Lays Potato chips is less than 5 oz.

* $H_0: \mu = 5$
* $H_A: \mu < 5$

He failed to reject the null. In reality, the true mean is equal to 5 oz.

**Joe made a correct decision: he failed to reject the null hypothesis when the null hypothesis is actually true.**

c) A test for an infectious disease gives a positive result if a patient is believed to have the disease, and a negative result otherwise. However, the test can sometimes be wrong. Label the following statements as a Type I error, Type II error, or neither. Assume the null hypothesis is that the patient doesn't have the disease.

* A positive test result; the patient actually has the disease 

**Neither**

* A positive test result; the patient doesn't actually have the disease

**Type I**

* A negative test result; the patient actually has the disease

**Type II**

* A negative test result; the patient doesn't actually have the disease

**Neither**

In the context of this question, which error (Type I or Type II) would be more problematic? If you could make one of these errors less likely to occur, which one would it be, and why?

**In the context of medical diagnostics, committing a Type II error tends to be more problematic. The reason is that we do not want an infected person to get a negative test result when in reality they have the disease. Hence, we want to make Type II errors less likely to occur. One downside to this is that lowering the Type II error rate often makes Type I errors more likely, so there is a trade-off.**

# Part 2: Review of Inference for (Simple) Linear Regression

As we saw last week, hypothesis testing can be done for more than sample means and proportions. We can also test hypotheses relating to regression parameters, like $\beta_1$, the slope, and $\beta_0$, the y-intercept. The hypotheses typically test whether the parameter is greater than zero, less than zero, or not equal to zero. 

# Question 2: Hypothesis Testing for Simple Linear Regression

Do larger countries have more rural land? Investigate this question using the `TenCountries` data set from the `Lock5Data` library. Fit a regression model, and run a hypothesis test to see if countries with more area tend to have a higher percentage of rural land. In other words, run a test to see if the slope coefficient is greater than zero. **Assume that the percent of rural land is the response variable, and the area of the country is the explanatory variable.**

a) First, create a scatterplot to visualize the relationship between `Area` and `PctRural`. Based on this scatterplot, do you think there is a relationship between the variables? Do you think we will be able to reject the null hypothesis?

```{r}
library(Lock5Data)
Area = TenCountries$Area
PctRural = TenCountries$PctRural

plot(Area, PctRural, xlab = "Area (1000 sq. kilometers)", ylab = "% Rural", 
     main = "Rural Land % by Area for Ten Countries")
```

b) State the null and alternative hypotheses using symbols

* $H_0: \beta_1 = 0$
* $H_A: \beta_1 > 0$

c) Fit the regression model with `PctRural` as the outcome, and `Area` as the predictor. Extract the slope coefficient by using the `coef()` function, and selecting the second value (e.g., `coef(my_model)[2]`).

```{r}
country_model = lm(PctRural ~ Area)
obs_country_coeffs = coef(country_model)
obs_country_slope = obs_country_coeffs[2]
obs_country_slope
```

d) Create a null distribution by using the `do_it` function. The approach you will want to take is to fit a regression model inside the `do_it` call (maybe call it `curr_model`), and you will use `PctRural` as the outcome, and a shuffled `Area` as the predictor. You can shuffle the `Area` variable using the `shuffle()` function. Extract the slope coefficient after fitting each model.

```{r}
library(SDS1000)
null_dist_country = do_it(10000) * {
  curr_mod = lm(PctRural ~ shuffle(Area))
  coef(curr_mod)[2]
}
```

e) Plot a histogram of your null distribution, and add a red vertical line at the observed slope

```{r}
hist(null_dist_country, main = "Null Distribution of Slope")
abline(v = obs_country_slope, col = "red")
```

f) Calculate the p-value by seeing the proportion of null values that are more extreme than the one you observed.

```{r}
pnull(obs_country_slope, null_dist_country, lower.tail = F)
```

g) State your conclusion.

**Since our p-value is less than 0.05, we will reject the null hypothesis. We therefore have evidence that countries with more area tend to have a higher percentage of rural land.**

h) A collaborator raises concerns about this analysis. She notes that there is a disagreement between your permutation results, and the output from the `summary()` function. What could be causing this issue? Could there be an issue with the data set that you used?

```{r}
summary(country_model)
```

**The sample size here is very small ($n = 10$), which makes fitting a regression model very difficult. The results for this analysis are not very trustworthy, which is why we see a disagreement between the two results. We should recommend that we obtain a larger sample size before running this analysis again. We also notice that the normality assumption of the error terms is not likely met. One way to check this normality assumption is to look at a histogram of the residuals. This can serve as a proxy for the true error terms. We hope that this histogram is approximately normal and centered around zero.**

```{r}
hist(country_model$residuals, xlab = "", ylab = "Residual", main = "Histogram of Residuals from Model")
```

**Unfortunately, the plot of residuals above is far from normal, which suggests that the normality assumption for the error terms is not met. We assume that the error terms are normally distributed with a mean of zero, and a standard deviation of one. But that does not appear to be the case here.**

# Question 3: Additional Hypothesis Testing for Simple Linear Regression

Are taller students worse at math? Data was collected from high school students that included their height and scores on the SAT math section. Fit a regression model and run a hypothesis test to see if taller tend to score worse on the math section. **Assume that SAT math score is the response variable, and student height is the explanatory variable.**\ 

The data is available in the `StudentSurvey` dataset from the `Lock5Data` library. We first will remove missing values from the data set using the `na.omit()` function.

a) First, create a scatterplot to visualize the relationship between `SAT math score (MathSAT)` and `Height`.

```{r}
StudentSurvey_clean = na.omit(StudentSurvey)
```

```{r}
sat_math = StudentSurvey_clean$MathSAT
height = StudentSurvey_clean$Height
plot(height, sat_math, xlab = "Height (in)", ylab = "SAT Math Score", 
     main = "SAT Math Score vs Height (in)")
```

b) State the null and alternative hypotheses using symbols

* $H_0: \beta_1 = 0$
* $H_A: \beta_1 < 0$

c) Fit the regression model with `Height` as the outcome, and `Weight` as the predictor. Extract the slope coefficient by using the `coef()` function, and selecting the second value (e.g., `coef(my_model)[2]`).

```{r}
sat_model = lm(sat_math ~ height)
obs_sat_coef = coef(sat_model)
obs_sat_slope = obs_sat_coef[2]
obs_sat_slope
```

d) Create a null distribution by using the `do_it` function. The approach you will want to take is to fit a regression model inside the `do_it` call (maybe call it `curr_model`), and you will use `MathSAT` as the outcome, and a shuffled `Height` as the predictor. You can shuffle the `Height` variable using the `shuffle()` function. Extract the slope coefficient after fitting each model.

```{r}
null_dist_sat = do_it(10000) * {
  curr_mod = lm(sat_math ~ shuffle(height))
  curr_sat_coef = coef(curr_mod)
  curr_sat_coef[2]
}
```

e) Plot a histogram of your null distribution, and add a red vertical line at the observed slope

```{r}
hist(null_dist_sat, main = "Null Distribution of Slope")
abline(v = obs_sat_slope, col = "red")
```

f) Calculate the p-value by seeing the proportion of null values that are more extreme than the one you observed.

```{r}
pnull(obs_sat_slope, null_dist_sat, lower.tail = T)
```

g) State your conclusion.

**Since our p-value is greater than 0.05, we will fail to reject the null hypothesis. We therefore do not have evidence that taller students score worse on the SAT math section.**

# Part 3: Inference for (Multiple) Linear Regression

So far, we have fit regression models where we had a single predictor or variable relating to our outcome variable. However, we often are interested in understanding the relationship among the outcome and two or more other variables. Regression with more than one predictor is known as multiple linear regression. 

## Question 4: Hypothesis Testing for Multiple Linear Regression

In this next exercise, we will predict arm span from both height and foot length. A study collected a sample of high school students and measured their arm span, height, and foot length. The data set is called `PASeniors`, and it is located in the `Lock5Data` library.

```{r}
library(SDS1000)
library(Lock5Data)
data(PASeniors)

cPASeniors<- na.omit(PASeniors)
```

1) Fit a simple linear regression model to predict arm span from height. See the outputs using the `summary()` function.

```{r}
# fit a simple linear regression to predict arm span from height
Armspan <- cPASeniors$Armspan
Height <- cPASeniors$Height
  
arm_height_mod = lm(Armspan ~ Height)
summary(arm_height_mod)
```

2) Fit a simple linear regression model to predict arm span from foot length.

```{r}
# fit a simple linear regression to predict arm span from foot
Foot <- cPASeniors$Foot

arm_foot_mod <- lm(Armspan ~ Foot)
summary(arm_foot_mod)
```

3) Let's now compare the two models that we fit in parts 2) and 3).

a) **Foot ($b1$ = 3.4835) has a larger slope than Height ($b1= 0.91491$).**

b) **Height ($se = 9.414$) has a smaller standard deviation of error than foot ($se = 9.937$).**

c) **Height with $R^2=51.5 \%$ explains more variability in arms span than foot with $R^2=46.0 \%$.**

d) **Both the smaller standard deviation of error in (b) and the larger $R^2$. In (c) indicate that Height is somewhat more effective than Foot for predicting arm span. The larger slope for foot is not so relevant since it also has a much larger standard error for the slope.**

&nbsp;

4) Now fit a multiple linear regression to predict arm span using both height and foot length.

```{r}
# fit a simple linear regression to predict `Armspan` from ` Foot` and ` Height`
arm_foot_height_mod <- lm(Armspan ~ Height + Foot)
summary(arm_foot_height_mod)
```

a) **The fitted model is arm span $=8.52+0.7356 \cdot$ Height $+1.4477 \cdot$ Foot. For a student with Height $=180$ and Foot $=26$ the predicted arm span is given by:**

$$
\widehat{\operatorname{Armsp}} a n=8.52+0.7356 \cdot 180+1.4477 \cdot 26=178.55
$$

b) **The p-values for the individual t-tests for both predictors are essentially zero, so we have strong evidence that both Height and Foot are effective in this model to predict arm span.**

c) **The output shows $R^2=61.75 \%$, so the model based on Height and Foot explains $61.75 \%$ of the variability of the arm span measurements for these students.**

